### 2022-12-22 manuscript.Rmd
initial commit

---
title: Material Culture Networks in Tonto Basin
output:
  bookdown::word_document2:
    toc: true
    number_sections: false
    reference_docx: TF.docx
  rticles::tf_article: null
author:
  - name: Robert J. Bischoff 1
    affil: a
    email: rbischoff@asu.edu
affiliation:
  - num: a
    address: School of Human Evolution and Social Change, Arizona State University, Tempe, AZ
keywords: |
 American Southwest; network analysis; multilayer networks; Hohokam; projectile points; ceramics; computational archaeology; gender
abstract: |
  Tonto Basin is located in a border zone between the Hohokam, Mogollon, and Ancestral Pueblo regions and was primarily occupied during the Roosevelt phase (AD 1275-1325) and the Gila phase (AD 1325-1450). The Roosevelt phase is notable for the appearance of Salado pottery, which has been interpreted as evidence of Kayenta immigration into Tonto Basin and the beginning of the Salado phenomenon. Salado pottery production was widespread, with production often centered at the location of former immigrant enclaves. This study uses data from the Roosevelt Community Platform Mounds Study to analyze social relations through networks based on architecture, ceramics, projectile points, and site locations. The results show a strong correlation between projectile points and space, and differences in the ceramic and point networks suggest that different social processes, possibly relating to gender, influenced the structure of the networks. Sites in this study with roomblocks--associated with immigration--are highly central in the ceramic networks, but have low centrality in the point networks, indication differences in the networks and suggesting differences in the levels of social integration between the genders.
bibliography: references2.bib
link-citations: yes
editor_options:
  chunk_output_type: inline
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
  \def\UrlBreaks{\do\/\do-\do?}
csl: american-antiquity.csl
---

```{r setup, echo = F}
out.format <- knitr::opts_knit$get("out.format")
options(knitr.table.format = "latex")
knitr::opts_chunk$set(
  echo = F,
  message = FALSE,
  warning = FALSE,
  dpi = 450,
  out.width = "100%"
)
```

```{r load, include=FALSE}
library(tidyverse,quietly = T)
library(magrittr)
library(Momocs,quietly = T)
library(kableExtra)
library(ggthemes, quietly = T)
library(here)
library(rio)
library(igraph)
library(ggnetwork)
library(vegan)
library(multinet)

set.seed(1010)

# library(ggimage)
i_am("manuscript.Rmd")
# keep track of figures and tables
figs = 1
tbls = 1
set.seed(1010)

# load data and remove sites with only one projectile point

sites = import(here("data/TontoBasinNetworkSites.xlsx"),setclass = 'tibble') %>% 
  ungroup() %>% 
  filter(!ArchaMapName %in% c("AZ V:5:130 (ASM)","AZ U:4:009 (ASM)","AZ U:8:318 (ASM)","AZ V:5:121 (ASM)"))

# data frame for network nodes
nodes = sites %>%
  select(ArchaMapName,Area,Architecture = architecture) %>% 
  mutate(Architecture = case_when(str_detect(Architecture,"compound, platform mound")~"platform mound",str_detect(Architecture,"compound, roomblock")~"roomblock",TRUE~Architecture)) %>% 
  rowid_to_column("id") %>% 
  ungroup()

# coords = outTonto$fac %>% 
#   select(ArchaMapName,UTMEAST,UTMNORTH) %>% 
#   distinct_all()
# coords %>% export(here('data/siteCoords.xlsx'))
coords = import(here('data/siteCoords.xlsx')) %>% 
  inner_join(nodes) %>% 
  arrange(id)
CeramicData = rio::import(here("data/CeramicData.xlsx"),
                          setclass='tibble')

# number of types per site
count = CeramicData %>% as.data.frame() %>%
  group_by(ArchaMapName,SWSN_type) %>%
  summarize(Count = sum(Count)) %>%
  ungroup()
table(count$ArchaMapName)
ceramicDF = CeramicData %>% as.data.frame() %>%
  group_by(ArchaMapName,SWSN_type) %>%
  summarize(Count = sum(Count)) %>%
  ungroup() %>%
  inner_join(nodes %>% select(ArchaMapName,id)) %>% 
  select(-ArchaMapName) %>% 
  pivot_wider(names_from = SWSN_type, values_from = Count) %>%
  column_to_rownames("id") %>% 
  select(-`Type, Undifferentiated`)

outTonto = readRDS(here("data/TontoNetworkOutlines.Rds"))

outTonto$fac %<>% select(-id,-Area) %>% left_join(nodes)
# outTonto %<>%
# mutate(ArchaMapName = droplevels(ArchaMapName))
# outTonto %>% saveRDS(here("data/TontoNetworkOutlines.Rds"))
```

```{r sessionInfo,eval=FALSE, include=FALSE}
# display software information
sessioninfo::session_info()
```

# Introduction {-}

Tonto Basin holds great potential for archaeological research due to the large cultural resource management projects conducted in the region [@Ahlstrom1991-eb; @Ciolek-Torrello1994-ds; @Doelle1992-lu; @Rice1998-ku]. This analysis uses the available information from the Roosevelt Community Platform Mounds Study (RPMS), the largest of the studies [@Rice1998-ku], and builds on prior research [@Clark2001-ck; @Hill2015-at; @Huntley2016-em; @Lyons2003-yy; @Lyons2006-to; @Lyons2012-pr;  @Neuzil2008-zd; @Stark1998-mu; @Watts2013-ub] to combine data on architecture, ceramics, projectile points, and site locations to analyze the networks of social relations that existed in the past through networks based on these data. The primary question is to determine whether these data are correlated. Because spatial data is included, this would mean that near sites are more similar than distant sites in architecture, ceramics, and projectile points. A lack of correlation would indicate that factors other than distance are influencing the social networks. The results of the network analysis demonstrate a strong correlation between  projectile points and space. The differences in the ceramic and point networks indicate that different social processes, perhaps relating to gender, are affecting the structure of the network.

Tonto Basin is located in east-central Arizona and features Tonto Creek flowing from the Mogollon Rim into the Salt River (figure `r figs`). The junction of the Salt River and Tonto Creek now forms Roosevelt Lake since the damming of the Salt River. It was the expansion of the dam that precipitated the RPMS and its related projects in the late 1980s-1990s. All of the artifacts/data from the RPMS project are hosted at Arizona State University, and the data in this study come exclusively from this project.

```{r TontoBasinSites, fig.cap= 'Map of archaeology sites from the Roosevelt Platform Mounds Study in Tonto Basin. Groups use the original assignments.'}

figs = figs + 1
knitr::include_graphics(here("figures/TontoBasinSitesv3.png"))
```

Most of the occupation of the Tonto Basin, particularly the sites excavated during the project, date between AD 1275 and 1325, which corresponds to the Roosevelt Phase. The Roosevelt phase is notable for the appearance of Salado pottery. Moderate occupation continued into the Gila phase (AD 1325-1450), after which recognizable occupation ceased in the area until the historic period.

One aspect of Tonto Basin that has caught the attention of many researchers is its location in a border zone between the Hohokam, Mogollon, and Ancestral Pueblo regions. Of particular note is the presence of masonry roomblock architecture and pottery uncharacteristic of the local Hohokam traditions, which has been interpreted as Kayenta immigration into Tonto Basin and the beginning of the Salado phenomenon [@Clark2001-ck; @Lyons2003-yy; @Lyons2006-to; @Stark1998-mu]. Salado pottery production was widespread, but often the largest sources of production was centered at the location of a former Kayenta enclave [@Hill2015-at; @Huntley2016-em; @Lyons2012-pr; @Neuzil2008-zd].

Ideally, studying social relations would be approached at the individual level, but rarely can archaeologists address archaeological data with that level of specificity. Watts [-@Watts2013-ub] used projectile points from Tonto Basin in just such a study, however. His study of flake scar patterning identified individual knappers, or at least clusters of knappers with similar knapping styles, and the distribution of their projectile points around Tonto Basin. He found strong connections between many parts of the eastern Tonto Basin.

One way to examine individuals who make similar types of projectile points--and similar types of pottery for that matter--is to talk about identity. Identity can be a troubling topic for anthropologists. There are numerous meanings given to it and numerous scales at which it applies [@Brubaker2000-na], but Peeples [-@Peeples2018-ib] has introduced archaeologists to a more practical approach that is useful here. This approach views identity as existing along two axes: one categorical and the other relational. Relational identification is the process of identifying with someone due to frequent interaction. Categorical identification is the process of identifying with someone because you belong to a recognized social group. For example, members of the same moiety would share a categorical identity. They would likely also share a relational identity if they frequently interacted. These identities can be reflected in material culture. Pottery makes a good example. A potter can produce large, black and red chevrons on a pot that everyone who sees it can associate with a particular social group. Members of that group may wish to use that pottery as well as those who wish to curry favor with that group. Others may avoid that pottery. This pottery would be a marker of categorical identity. Relational identity can be seen in the way the potter prepares the clay recipe and smooths the coils. The particulars of these actions would be learned from close interaction with a teacher or other potters. In this way, relational identity can be compared to communities of practice [@Lave1991-ik; @Wenger1998-rx].

In this analysis, I argue that ceramics and architecture are more likely to represent categorical identity. The architecture discussed is highly visible and representative of historical group membership. The ceramics are grouped into types based primarily on decoration, which is strongly indicative of group membership [see @Clark2001-ck for a discussion on material culture and its relationship to identity]. Projectile points can in some cases indicate group membership, but in this case the triangular and side-notched points were likely used for different purposes [see @Loendorf2015-ww]. Styles were difficult to determine for the original researchers, and I had the same difficulty. In my opinion, the subtle differences between projectile point outlines are more representative of interactions between knappers than markers of group identity for this particular case. There are no environmental reasons to assume differences in projectile points. Each hunter would have been using the points for the same game.

Another way that people identify themselves is gender. Gender roles, like identity, can also vary, but for the purposes of this analysis I will use a simplified model. Women made pottery and men made projectile points. This was not always true of course, but it generally fits the ethnographic record in the Southwest. This does not mean that a man could not move pots from one site to another or that a woman could not do the same with a projectile point, but I will make the assumption that ceramic networks and point networks are gendered. Therefore differences in these networks represent different interactions between men's and women's networks.

From this discussion, I expect the ceramic and point networks to differ in two ways. I expect the ceramic network to represent categorical identity among women, and I expect the point network to represent relational identity among men. This makes the networks somewhat more difficult to compare. Ideally, we would have networks of the same identity type, but it does provide the expectation that we should see significant differences in the networks.

The presence of migrant communities in Tonto Basin invites questions regarding the relationship between sites--particularly between migrant and local communities. A network approach is an ideal way to examine the relationships between sites. Evidence that projectile points move between sites [@Watts2013-ub] indicates that this is also a productive line of analysis, and one that is less-commonly pursued compared to architecture and ceramics.

# Network Analysis {-}

Network approaches are best suited where the relationships between entities are of primary interest. In this case, I compare the similarity between sites using architecture, ceramics, and projectile points. Thus, a network can be defined using a site from Tonto Basin as a node and a type of similarity as a link (or edge) between the nodes. Similarity networks are a commonly used type of network in archaeology [e.g., @Birch2018-xx; @Borck2015-jy; @Cochrane2010-oa; @Golitko2015-we; @Lulewicz2019-lu; @Mills2013-wq; @Peeples2018-ib; @Terrell2010-hd]. These studies use a variety of artifacts and methods to construct their networks, but each has demonstrated the utility of network methods within archaeological contexts.

Table `r tbls` 1 lists the `r nrow(sites)` sites with sufficient data to include in this study with the types of architecture present, periods occupied, and total number of ceramics and complete projectile points found at each site. The lack of complete projectile points at some sites is the largest caveat in this study, but complete points were necessary for the type of projectile point analysis used. The periods, architecture types, and ceramic types were assigned during the RPMS project and can be found in the project reports [@Rice1998-ku].

```{r sites, echo=FALSE}
tbls = tbls + 1
cap = "Project Sites and Data"
sites %>% 
  select(Site = ArchaMapName,Area,Architecture = architecture,Phase,Ceramics,Points) %>% 
knitr::kable(caption= cap, 
             booktabs = T,
             format.args = list(big.mark = ",",
  scientific = FALSE)) %>% 
  kable_styling(latex_options = "scale_down")
```

The types of similarity used in this study are varied depending on the type of network. One parameter that is consistent is how similar links should be in order to keep them. There is rarely a clear dividing line between similar and not similar. This can be a challenge for network analysis, because we can end up with networks where every node is connected to every other node. The decision to binarize a network--remove the weakest links and then cosider each link of equal value--has is drawbacks [@Peeples2013-tj], but is necessary in this case. A solution is to assign weights to each link that defines the strength of the tie. These networks are often difficult to visualize, and some network algorithms do not allow for weights. A common approach is to keep only the strongest ties by either ranking the ties or using a cutoff value. Part of this analysis is a visual approach and some of the methods require non-weighted links, thus I will be only using the strongest links.

I used ranked links to keep consistent values between networks. For example, a network could be made of the `r nrow(sites)` sites as nodes with each node being connected to the three sites that it has the strongest similarity to for ceramics, architecture, projectile points, and the three sites that it is physically closest to for the spatial network. In practice, not every node will have three connections. More than three ties can exist when the ties are tied in rank. Some nodes may not have three connections. And finally, because the network is not directed (meaning the ties indicate similarity in both directions) one node can have several ties pointing to it because that node was in the top three of multiple other nodes. In the latter case, this is a good indication of a central node in the network. Because I am using several types of networks, there is no common cutoff value I can use. Instead of arbitrarily picking one value, I have calculated each metric using networks composed of the top three to ten ties. Examining the range of these metrics provides a more robust analysis.

Two network metrics are discussed in this analysis: eigenvector centrality and multilayer network correlation--the second will be discussed in the next section. Node centrality is a common way to quantify networks [@Borgatti2005-iz]. Centrality is a measure of the influence of the node on the network. Nodes with higher centrality generally derive or generate greater benefits from the network. Eigenvector centrality is one way to measure node centrality and is commonly used in archaeology. This metric describes how well a node is connected to the network as a whole and is helpful in comparing different networks containing the same nodes. Eigenvector centrality is also more robust to missing nodes than other measures [@Peeples2017-kl], a major problem in most archaeological studies.

## Multilayer Networks {-}

A single network is typically used in network analyses, but multilayer networks can be more informative. Multilayer networks are layered networks where nodes have different types of connections [@Kivela2014-mh]. In this analysis, each network has the same nodes--each archaeological site--but different types of relationships between them. The combination of these individual networks is a multilayer network (also called a multiplex network).

Multilayer networks allow for methods to be applied on multiple networks at once [see @Brodka2018-vz]. The method applicable to this analysis is layer correlation. Either Pearson or Spearman rank correlation can be computed to determine the strength and direction of correlation between each layer. Bródka and colleagues [-@Brodka2018-vz] have provided an R package to compute these statistics, which I have used in this study. They recommend the Pearson correlation in most circumstances. I will display the results of the Pearson correlation, but the Spearman correlation provided similar results.

```{r}
# network functions 
makeGraph = function(d, nLinks = NULL){
  # igraph
  g = graph_from_adjacency_matrix(d,mode = 'undirected',diag = F,weighted = T)
  V(g)$name = nodes$id %>% as.character()
  V(g)$ArchaMapName = nodes$ArchaMapName
  V(g)$Area = nodes$Area
  V(g)$Architecture = nodes$Architecture
  V(g)$centrality = eigen_centrality(g, weights = E(g)$weight)$vector

  if(!is.null(nLinks)){
    # calculate top n (with ties) edges and convert to igraph object
    # convert to dataframe
    n = igraph::as_data_frame(g)
    n = bind_rows(n,n %>% rename(from = to, to = from))
    n %<>% # sort data frame
      arrange(from,desc(weight)) %>%
      # rank by site
      group_by(from) %>%
      mutate(rank = rank(-weight)) %>% 
      ungroup()
    nSub = n %>% 
      group_by(from) %>% 
      slice_max(n = nLinks, order_by = desc(rank)) %>% 
      ungroup()

    gRanks = graph_from_data_frame(nSub,directed = F, vertices = nodes)
    gRanks %<>% simplify()
    V(gRanks)$centrality = eigen_centrality(gRanks, weights = E(gRanks)$weight)$vector
    return(gRanks)
  } else {
    return(g)
  }
}

values2graphicsMod = function (values, output = "color", palette = "Paired") 
{
  if (!output %in% c("color", "shape")) {
    stop("wrong parameter: output")
  }
  if (is.data.frame(values)) {
    values <- values[[1]]
  }
  types <- as.factor(values)
  num_types <- length(levels(types))
  color_map = brewer.pal(num_types, palette)
  color_map[types]
  if ((output == "shape") & num_types > 5) {
    warning("only 5 distinct shapes available: some shapes have been repeated")
  }
  l = vector("list")
  l[["legend.text"]] = levels(types)
  l[["legend.pch"]] = 21
  if (output == "color") {
    l[["legend.col"]] = color_map[1:num_types]
    l[["color"]] = color_map[types]
  }
  if (output == "shape") {
    l[["legend.pch"]] = ((1:num_types) - 1)%%5 + 21
    l[["shape"]] = (as.numeric(types) - 1)%%5 + 21
  }
  l
}

dists = list()
```

```{r}
# Spatial network
d = dist(coords[,c("UTMEAST","UTMNORTH")]) %>% as.matrix
colnames(d) = rownames(d) = coords$id
# convert to scaled similarity (1 highest)
d = 1 - (d / max(d))
dists$spatial = d
```

## Spatial {-}

The simplest network is the spatial network. This network was created by calculating the Euclidean distance from every site to every other site. This method is equivalent to proximal point analysis, which was used in one of the earliest network studies in archaeology [@Terrell1977-co] and continues to be used to study potential pathways of interaction [see @Broodbank2000-pg; @Collar2013-fb].

## Architecture {-}

```{r}
# architecture network
# convert to adjacency matrix
tmp = sites %>% 
  separate_rows(architecture, sep = ", ") %>% 
  left_join(nodes %>% select(ArchaMapName,id))
m = table(tmp$id,tmp$architecture) %>% as.matrix
# calculate jaccard distance
d = vegdist(m,method = "jaccard", na.rm = T) %>% as.matrix
# convert to jaccard similarity
d = 1-d
dists$architecture = d
```

The architectural components consisted of compounds, roomblocks, and platform mounds. Some sites had pit houses from earlier occupations, but most excavation was focused on the Roosevelt and Gila phases so these few pit houses were removed from the analysis [see @Rice1998-ku]. Each site was classified as a platform mound if one was present regardless of other architecture or as a roomblock if a platform mound was not present regardless of other architecture. Thus, if a site is labeled a platform mound it may have other architecture. The network for architecture, and for the remaining networks, was created using the Jaccard similarity coefficient [@Jaccard1912-na] to determine the links between each site. This is a simple measure of similarity that does not take into account abundance. It was chosen due to the large differences in sample sizes between sites for the ceramics and projectile points and because count data were irrelevant or not readily available for the architectural data.

## Ceramics {-}

```{r}
# ceramic stats
# cdata = ceramicDF %>% rownames_to_column("sites") %>%  janitor::adorn_totals()
# summary(cdata[16,-1] %>% unlist %>% as.integer)

# ceramic network
# convert to adjacency matrix
m = ceramicDF %>% as.matrix
# calculate jaccard distance
d = vegdist(m,method = "jaccard", na.rm = T) %>% as.matrix
# convert to jaccard similarity
d = 1-d
dists$ceramics = d
```

The ceramic network consists of 55 types--plainware and decorated--ranging from 181,666 Plain Brown Ware sherds to a single Black Mesa Black-on-white sherd. A single sherd is not much, but is still indicative of trade. However, the median sherd count was 157 for all types. The ceramic data was accessed from the cyberSW database [@Mills2020-tb], which keeps the full citations for the ceramic data.

## Projectile Points {-}

```{r}
# projectile point network
# convert to adjacency matrix
m = table(outTonto$fac$id,outTonto$fac$pointCluster) %>% as.matrix
# calculate jaccard distance
d = vegdist(m,method = "jaccard") %>% as.matrix
# convert to jaccard similarity
d = 1-d
dists$points = d
```

The ceramic and architectural data is available in suitable formats for analysis, but the original projectile point analysis was too general for the purposes of this study [see @Rice1994-rk for details]. Geometric morphometrics is a set of methods designed to quantitatively analyze shapes and is ideal for projectile point analysis. See Bischoff [-@Bischoff2022-rg] for details on geometric morphometrics and the methods used to analyze these projectile points. A geometric morphometric analysis was conducted on the 2D outlines of each complete triangular or side-notched projectile point. The majority of projectile points were triangular or side-notched. The few other points were likely curated points from earlier periods. Figures `r figs` and `r figs + 1` show the results of these analyses. These results were classified into closely related clusters and used in the same way as ceramic types.

```{r TontoSideDistanceNetwork, fig.cap= 'Network graph displaying side-notched points from Tonto Basin as nodes with ties showing the morphometric distance between points. Darker colors represent stronger ties. Note that only the strongest 10\\% of ties are shown. From Bischoff (2022: Figure 14).'}
figs = figs + 1
knitr::include_graphics(here("figures/TontoSideDistanceNetwork.png"))
```

```{r TontoTriangularDistanceNetwork, fig.cap= 'Network graph displaying triangular points from Tonto Basin as nodes with ties showing the morphometric distance between points. Darker colors represent stronger ties. Note that only the strongest 10\\% of ties are shown. From Bischoff (2022: Figure 15).'}
figs = figs + 1
knitr::include_graphics(here("figures/TontoTriangularDistanceNetwork.png"))
```

# Results {-}

```{r include=FALSE}
corr = list()
eigen = list()
for (nLinks in 3:10){
  print(nLinks)
  net = ml_empty("Tonto")
  g2List = list()
  nms = names(dists)
  for(n in nms){
    print(n)
    g = makeGraph(dists[[n]], nLinks = nLinks)
    eigen[[paste0(nLinks,"-",n)]] = V(g)$centrality
    g2List[[n]] = g
    add_igraph_layer_ml(n = net, g = g,name = n)
  }

  # layer_comparison_ml(net, method = "jaccard.edges", mode = "all") %>% 
  # print
  corr[[nLinks]] = layer_comparison_ml(net, method = "pearson.degree", mode = "all") %>% as.data.frame() %>% 
    mutate(nLinks = nLinks)
}

corrDF = bind_rows(corr) %>% 
  rownames_to_column("from") %>% 
  pivot_longer(-c(from,nLinks), names_to = "to") %>% 
  mutate(from = str_remove_all(from,"[^A-z]")) %>% 
  mutate(nLinks = factor(as.character(nLinks))) %>% 
  rowwise()  %>% 
  filter(from != to) %>% 
  mutate(id = paste(sort(c(from,to)),collapse = "->")) %>% 
  select(-from,-to) %>% 
  distinct_all()#%>% 
# pivot_wider(names_from = nLinks)
corrDF %>% 
  ggplot(aes(x = value, y = id)) + 
  geom_boxplot() +
  theme_gdocs() +
  xlab("Pearson correlation") +
  ylab("Network comparison")
# ggsave(here("figures/TontoNetworksCorrelation.png"),dpi = 450, width = 6.5, height = 4.5)
# browseURL(here("figures/TontoNetworksCorrelation.png"))

eigenDF = bind_rows(eigen) %>% 
  rowid_to_column("id") %>% 
  pivot_longer(-id) %>% 
  left_join(nodes) %>% 
  mutate(name = str_remove_all(name,"[^A-z]"),
         id = factor(id))

eigenDF %>% 
  ggplot(aes(x = value, y = id, color = Architecture)) + 
  geom_boxplot() +
  theme_gdocs() + 
  xlab("Eigenvector centrality") +
  ylab("Site ID") +
  guides(color = guide_legend(title = element_blank(), nrow = 1)) +
  theme(legend.position = "bottom") +
  facet_wrap(~name, scales = "free_x")
# ggsave(here("figures/TontoNetworksEigenvector.png"),dpi = 450, width = 6.5, height = 8.5)
# browseURL(here("figures/TontoNetworksEigenvector.png"))

g2List = list()
nms = names(dists)
for(n in nms){
  print(n)
  g = makeGraph(dists[[n]], nLinks = 5)
  eigen[[paste0(nLinks,"-",n)]] = V(g)$centrality
  g2List[[n]] = g
}

netDF = map2_df(g2List,names(g2List),function(g,n){
  ggnetwork(g) %>% mutate(layer = n)
})

ggplot(netDF,aes(x = x,y = y, xend = xend, yend = yend, shape = Architecture, color = Area, label = name, size = centrality)) +
  geom_edges(color = "gray", size = .75) +
  geom_nodes() +
  geom_nodetext_repel(color = "black", size = 3) +
  scale_shape_manual(values = 15:18) +
  scale_color_brewer(palette = "Set1") +
  scale_size_continuous(range = c(3,4.5)) +
  facet_wrap(~layer) +
  theme_blank() +
  theme(panel.border = element_rect(fill = NA),
        strip.background = element_rect(fill = "white"),
        legend.position = "bottom") +
  guides(size = "none", shape = guide_legend(ncol = 2,title = NULL), color = guide_legend(ncol = 2, title = NULL))
# ggsave(here("figures/TontoNetworksMultiNet.png"),dpi = 450, width = 6.5, height = 8)
# browseURL(here("figures/TontoNetworksMultiNet.png"))
paste0(nodes$id,": ",nodes$ArchaMapName) %>% paste(collapse = "; ")
meanEigenVals = eigenDF %>% 
  filter(name %in% c("ceramics","points")) %>% 
  group_by(Architecture,name) %>% 
  summarize(value = round(mean(value),2)) %>% 
  arrange(name) %>% 
  pivot_wider(names_from = name, values_from = value)
```

Figure `r figs` shows the networks with the top five strongest ties between each node. This number of ties represents the fewest ties that connects all nodes in the spatial network. Fewer ties leaves the Cline Mesa sites disconnected from the rest of the network. Thus, Bass Point Mound forms a crucial bridge between the Cline Mesa sites and the rest of the network. The Rock Island area consists of a single site, Bass Point Mound, which was a platform mound, although it was not heavily excavated. If spatial distance was a primary driver in social interaction, then we would expect Bass Point Mound to consistently be a highly central node.

```{r TontoNetworksMultiNet, fig.cap= 'Graphs showing the networks used in this analsyis. Each network is displayed with the five strongest ties between each node. Legend: 1: AZ U:3:128 (ASM); 2: AZ U:4:032 (ASM); 3: AZ V:5:119 (ASM); 4: Bass Point Mound; 5: Cline Terrace Mound, Monster Ruin; 6: Indian Point Complex; 7: Pinto Point Mound; 8: Saguaro Muerto; 9: Sand Dune Site; 10: Schoolhouse Point Mesa Complex; 11: Schoolhouse Point Mound'}
figs = figs + 1
knitr::include_graphics(here("figures/TontoNetworksMultiNet.png"))
```

I chose eigenvector as the centrality measure (see figure `r figs`), which you can see in the spatial eigenvector results poorly reflects the centrality of Bass Point Mound. Betweenness centrality would better capture how objects would have to flow through Bass Point Mound to get from the Cline Mesa sites to the other parts of the network, but this is not a road network. People could have simply walked around Bass Point Mound. What eigenvector centrality does indicate is whether it is connected to other strongly connected sites. This may not be important for a spatial network, but it is important for the other networks. Bass Point Mound's centrality is very low for architecture. It is the most central site for points based on eigenvector analysis and visual inspection and it is high for ceramics as well. This indicates a strong correlation with space and points, at least for the Bass Point Mound, and that Bass Point Mound was also important for Ceramic circulation (whether that is actual pots or the ideas of how to make pots).

```{r TontoNetworksEigenvector, fig.cap= 'Boxplots showing the eigenvector centrality for each network. Values were calculated using networks created using between 3 to 10 of the strongest ranking ties. Legend: 1: AZ U:3:128 (ASM); 2: AZ U:4:032 (ASM); 3: AZ V:5:119 (ASM); 4: Bass Point Mound; 5: Cline Terrace Mound, Monster Ruin; 6: Indian Point Complex; 7: Pinto Point Mound; 8: Saguaro Muerto; 9: Sand Dune Site; 10: Schoolhouse Point Mesa Complex; 11: Schoolhouse Point Mound.'}
figs = figs + 1
knitr::include_graphics(here("figures/TontoNetworksEigenvector.png"))
```

Bass Point Mound is not the only bridge node. Pinto Point Mound connects the platform mounds with the compounds in the architecture network. The reason for this is that the site also has a compound present. Similar to Bass Point Mound, eigenvector centrality does not favor bridge nodes, but Pinto Point Mound has the second highest centrality value for ceramics and the highest for spatial distance (meaning it is located near many other sites), and it has an above average centrality for points. This suggests architecture is an important component in the ceramic network and possibly in the point network.

Besides the two bridge nodes, visual inspection of the graphs provides several other observations. One is that the four geographic areas represented do not form strong components in the non-spatial networks like they do in the spatial network. The architecture is thoroughly mixed, and the ceramics and points form single component networks. This is typical of artifact similarity networks. The Livingston sites exhibit the most clustering in the ceramic network, although some of the connections go through Cline Mesa sites, which are as geographically distant as possible for these sites. The Schoolhouse Mesa sites always group together in the ceramic and point networks, as do the Livingston compound sites and separately the Livingston roomblock and platform mound. The Cline Mesa sites exhibit more variability.

Another observation from the visual inspection of these networks is that there is some clustering due to architecture in the ceramic and point networks. Four of the five compound-only sites cluster at one end of the ceramic network, and a different mix of four out of five cluster together at one end of the projectile point network. The platform mounds are split in the ceramic network, but all cluster together in the projectile point network. Table `r tbls` shows the mean eigenvector centrality vectors for the ceramic and point networks by type of architecture. Surprisingly the platform mounds are the least central, on average, for the ceramic network, but they are the most central for the projectile point network. The roomblocks, on the hand, are the most central for the ceramic network and by far the least central in the projectile point network. They take a larger drop in centrality in the projectile point network than the platform mound sites do in the ceramic network. It is expected that roomblocks have high centrality in the ceramic network, but not that platform mounds should have the lowest centrality.

```{r meanEigenVals}
cap = "Mean Eigenvalues for Ceramic and Point Networks by Type of Architecture"
meanEigenVals %>% 
  knitr::kable(caption= cap, 
             booktabs = T)
```

There is a chain of association between immigrants and roomblocks, roomblocks or locations with roomblocks as centers of Salado production, and Salado pottery as a widespread phenomenon that provides an expectation for roomblock sites to be central to pottery networks. This expectation is born out in this network analysis. Two of the platform mounds and three of the compound sites also have high centrality, demonstrating that ceramic circulation occurred between groups at sites with all types of architecture. What is clear from this analysis is that centrality in the ceramic network does not correlate with centrality in the point network.

The visual analysis, combined with the eigenvector centrality, indicated some spatial and architectural correlation. The multilayer Pearson correlation provides a more direct way to compare these layers, as shown in figure `r figs`. This analysis provides a clear contrast between layers. Only one comparison had a strong correlation--the point and spatial networks were strongly correlated. The visual inspection and centrality analysis provided some indication of this, but the multilayer network comparison provides a clearer indication. All other layer comparisons had an approximately equivalent, positive correlation, but with only a week strength. The one exception was architecture and points where the correlation was approximately zero. The visual inspection and eigenvector analysis demonstrated some interesting interactions with architecture in the network, but the overall correlation demonstrates that architecture cannot be used to predict the presence of absence of network ties in the projectile point network.

```{r TontoNetworksCorrelation, fig.cap= 'Boxplots showing the Pearson correlation between each network. Values were calculated using networks created from 3 to 10 of the strongest ranking ties.'}
figs = figs + 1
knitr::include_graphics(here("figures/TontoNetworksCorrelation.png"))
```

# Discusssion {-}

This analysis has three main findings regarding networks in Tonto Basin: (1) that point networks correlate with space; (2) that roomblock sites are highly central in the ceramic network and have low centrality in the point network; and (3) that ceramic and point networks are significantly different from each other. The purpose of this analysis is to understand something about the social relations in Tonto Basin. Presumably, these networks indicate something about how people interacted, and the ceramics and projectile points reveal something about the identity of those who made, used, and discarded them.

As discussed previously, my simplistic model of interaction in Tonto Basin assumes that the ceramic network represents categorical identification among women. Recall also that I expected architecture to represent categorical identity. Thus, the correlation between architecture and ceramics, at least as represented by roomblocks, is an expected find and good corroborating evidence that these types of material culture represent markers of identity demonstrating belonging to a particular social group.

The point networks were expected to represent relational identification among men. This is not always the case with projectile points. In some cases they can represent group identity [see @Wiessner1983-ei], but in this case there is little variation in style. Because relational identification is related to frequent interaction, spatial distance is a crucial component. It is much harder to interact with someone when they are far away. Thus, the correlation between the point and spatial networks also makes theoretical sense.

The remaining question is whether the differences between the ceramic and projectile point networks are due to the differences in relational and categorical identification or are due to differences in behavior related to gender. I argue that in this case gender likely played an important role. It is most probable that potters were making the same types of pottery, rather than switching between styles. Not to say that potters were not making plain ware and decorated at the same time, but that they were not making radically different styles of decorated wares at the same time. Thus, the relational ceramic network likely follows the categorical network. This is however, an assumption that needs to be tested by examining sourcing data. The differences between the point and ceramic networks are large enough, that I am confident enough to hazard an opinion that these networks do represent gendered networks.

# Conclusion {-}

This analysis used architectural, ceramic, projectile point, and spatial data from `r nrow(sites)` sites in Tonto Basin, a region in central Arizona. The data come primarily from occupations between AD 1275 and 1325. A network was created for each type of data and combined into a multilayer network. Visual network analysis, eigenvector centrality, and multilayer network Pearson correlation was used to study and compare the networks.

The analysis found correlations between certain types of architecture and ceramic networks and between the point and spatial networks. These correlations are consistent with architecture and ceramics as markers of categorical identity--meaning these types of material culture were used as markers of belonging to a recognized social group. The correlation between projectile points and spatial location is also consistent with projectile point form as a marker of relational identity--meaning that the similarity between projectile points indicates frequent interaction.

Generally, projectile points were used in the Southwest by men in hunting and warfare. Women are most often associated with making and transporting pottery. I suggest that the large differences found between the point and ceramic networks indicate that men and women had different social networks in Tonto Basin. The large differences in centrality between the sites with roomblocks in the ceramic networks versus the point networks is also a possible indication of another social reality. If these networks are indeed gendered, then might immigrant women have been more socially connected than the men? Perhaps the simplest explanation is that male immigrants made projectile points they were more familiar with and felt no need to adapt to local customs, whereas the women felt greater pressure to integrate. Although it seems it was the immigrants' traditions that became the norm for pottery. One point that may support the idea of a lower level of integration among the men is the idea of gift giving or exchange of projectile points. Watt's [-@Watts2013-ub] analysis found that projectile points made by individuals were found throughout the eastern Tonto Basin, which supports the exchange of these items. There are several examples of the exchange of bows and arrows ethnographically throughout the world [e.g., @Mauss1966-nm; @Nishiaki2013-lc; @Wiessner1983-ei], in North America generally [e.g., @Hoffman1896-xl; @Radin1923-gf], and in the Southwest specifically [e.g., Beaglehole1936-ul; @Dittert1959-rz; @Fewkes1898-oj; @griffen1969a; @Parsons1939-xg; @Simpson1953-ob]. The frequent exchange of points would have increased the centrality of roomblock sites in the point networks. The fact that this did not happen may be an indication that male immigrant networks were not as well integrated as female networks centered around potters. Whichever is the case, this analysis demonstrates that multilayer network comparisons using different types of material culture is a profitable avenue of research and one that deserves further attention.

# Acknowledgements {-}

Josh Watts, Matt Peeples, Melissa Powell, Chris Caseldine, and several volunteers assisted with this research. The many individuals who contributed to the Roosevelt Platform Mound Study also deserve recognition.

# Disclosure Statement {-}

The author declares that no conflicts of interest exist in relation to the content of the article.

# References {-}

### 2022-12-28 manuscript.Rmd
finished updating paper based on Chris Caseldine's comments

@@
 output:
+  rticles::tf_article: null
   bookdown::word_document2:
@@
     reference_docx: TF.docx
-  rticles::tf_article: null
 author:
-  - name: Robert J. Bischoff 1
+  - name: Robert J. Bischoff
     affil: a
@@
 abstract: |
-  Tonto Basin is located in a border zone between the Hohokam, Mogollon, and Ancestral Pueblo regions and was primarily occupied during the Roosevelt phase (AD 1275-1325) and the Gila phase (AD 1325-1450).
-The Roosevelt phase is notable for the appearance of Salado pottery, which has been interpreted as evidence of Kayenta immigration into Tonto Basin and the beginning of the Salado phenomenon.
-Salado pottery production was widespread, with production often centered at the location of former immigrant enclaves.
-This study uses data from the Roosevelt Community Platform Mounds Study to analyze social relations through networks based on architecture, ceramics, projectile points, and site locations.
-The results show a strong correlation between projectile points and space, and differences in the ceramic and point networks suggest that different social processes, possibly relating to gender, influenced the structure of the networks.
-Sites in this study with roomblocks--associated with immigration--are highly central in the ceramic networks, but have low centrality in the point networks, indication differences in the networks and suggesting differences in the levels of social integration between the genders.
+  Network analysis has a strong foundation in Southwest archaeology, yet the combination of multiple types of artifacts in one analysis--multilayer network analysis--has not been formally applied except within a single artifact type.
+Many studies consider material culture holistically, yet network analysis has the advantage of focusing specifically on the relationships between entities (often communities) and how the structure of a network can benefit particular entities.
+This study takes architecture, ceramic, projectile point, and site location data from the Roosevelt Platform Mound Study and combines these data into a multilayer network analysis.
+This analysis provides a way to test the co-variance of these types of material culture with each other and with spatial variation.
+The results demonstrate a strong association between projectile points and spatial distance.
+Sites containing roomblock architecture are highly central in the ceramic network but have low centrality in the projectile point network.
+Overall, the ceramic and projectile point networks exhibit significant differences.
+This indicates that the social networks that created these patterns had different social mechanisms.
+One potential cause of these differences is gendered spheres of interaction with men exchanging projectile points and women exchanging ceramics.
 bibliography: references2.bib
@@
 csl: american-antiquity.csl
+always_allow_html: true
 ---
@@
 out.format <- knitr::opts_knit$get("out.format")
-options(knitr.table.format = "latex")
+# options(knitr.table.format = "latex")
 knitr::opts_chunk$set(
@@
 library(kableExtra)
+library(huxtable)
 library(ggthemes, quietly = T)
@@
 
-Tonto Basin holds great potential for archaeological research due to the large cultural resource management projects conducted in the region [@Ahlstrom1991-eb; @Ciolek-Torrello1994-ds; @Doelle1992-lu; @Rice1998-ku].
-This analysis uses the available information from the Roosevelt Community Platform Mounds Study (RPMS), the largest of the studies [@Rice1998-ku], and builds on prior research [@Clark2001-ck; @Hill2015-at; @Huntley2016-em; @Lyons2003-yy; @Lyons2006-to; @Lyons2012-pr;  @Neuzil2008-zd; @Stark1998-mu; @Watts2013-ub] to combine data on architecture, ceramics, projectile points, and site locations to analyze the networks of social relations that existed in the past through networks based on these data.
-The primary question is to determine whether these data are correlated.
-Because spatial data is included, this would mean that near sites are more similar than distant sites in architecture, ceramics, and projectile points.
-A lack of correlation would indicate that factors other than distance are influencing the social networks.
-The results of the network analysis demonstrate a strong correlation between  projectile points and space.
-The differences in the ceramic and point networks indicate that different social processes, perhaps relating to gender, are affecting the structure of the network.
+Network science has many applications for archaeologists.
+It can be particularly useful for breaking out of traditional spatial categories and examining data through new lenses [@Feinman2020-nf; @Holland-Lulewicz2021-eb].
+Multilayer network analysis combines multiple networks into one analytical framework [@Kivela2014-mh].
+Social systems are complex webs of interrelations.
+There is too much involved to model the material culture of even a technologically simple society.
+But certainly analyzing more than one type of material culture provides a more complete understanding of the social dynamics involved in the networks of interaction we wish to understand.
+Yet, there are few examples of multilayer networks in archaeology [@Giomi2022-dn; @Upton2019-yg] and none I am aware of that extend beyond ceramics.
+
+Network studies are particularly prominent in the Southwest and there are examples of network studies that consider multiple types of material culture.
+For example, Mills and colleagues [-@Mills2013-gc] used ceramics and obsidian to understand the transformation of large scale social networks.
+Peeples [-@Peeples2018-ib] used ceramics and architecture to understand identity and social change in the Cibola region.
+This paper is an application of multilayer network analysis that seeks to further network research by examining multiple types of material culture in one framework.
+Given the frequent focus on single types of artifacts in network research, a primary question I seek to address is whether networks based on different types of material culture are positively correlated.
+This is an important question because archaeologists often make inferences based solely on ceramics or other types of material culture.
+Evidence regarding the ways material culture co-vary will help archaeologists direct future research.
+Furthermore, I also seek causal explanations for the co-variance, or lack thereof, for material culture in the case study I present here.
+
+Tonto Basin holds great potential for archaeological research due to the large cultural resource management projects conducted in the region primarily in the 1980s-1990s [@Ahlstrom1991-eb; @Ciolek-Torrello1994-ds; @Doelle1992-lu; @Rice1998-ku], as well as the large number of syntheses and other studies focusing on the area [e.g., @Clark2004-uw; @Dean2000-uc; @Elson1995-tb; @Elson2000-zq;  @Hill2015-at; @Huntley2016-em; @Lange1992-gp; @Lyons2003-yy; @Lyons2006-to; @Lyons2012-pr;  @Neuzil2008-zd; @Oliver2001-tu; @Stark1998-mu; @Watts2013-ub].
+This analysis uses data from the Roosevelt Platform Mounds Study (RPMS), the largest of these projects [@Rice1998-ku], to examine basic architectural data, typed ceramics, and a recent projectile point analysis [@Bischoff2022-rg].
+There are two null hypotheses tested in this analysis: (1) that the data is spatially correlated--meaning that sites nearest to each other will be most alike--and (2) that the architecture, ceramics, and projectile points will all be positively correlated--meaning that similar types of architecture, ceramics, and projectile points will be found at the same sites.
+In reality, I expected significant variation between these types of material culture.
+As will be demonstrated, the results indicate a positive correlation between projectile point similarity and spatial distance.
+Furthermore, ceramic and projectile point networks (hereafter point networks) exhibit significant differences.
+I posit that the differences in these networks are connected to the identities of the individuals creating the original social networks: particularly that of gender.
+
+## Tonto Basin
 
 Tonto Basin is located in east-central Arizona and features Tonto Creek flowing from the Mogollon Rim into the Salt River (figure `r figs`).
-The junction of the Salt River and Tonto Creek now forms Roosevelt Lake since the damming of the Salt River.
+Parts of the region have been intensively studied, while certain time periods and much of the uplands have been less intensively sampled [see @Clark2021-lu for a recent overview].
+The basin was also conveniently located along regional travel routes, allowing major settlements--particularly those with platform mounds--to participate and benefit from this exchange [@Caseldine2022-uu; @Wood2000-ze, p.
+129-133].
+The junction of the Salt River and Tonto Creek now forms Roosevelt Lake since the damming of the confluence of the Salt River and Tonto Creek.
 It was the expansion of the dam that precipitated the RPMS and its related projects in the late 1980s-1990s.
-All of the artifacts/data from the RPMS project are hosted at Arizona State University, and the data in this study come exclusively from this project.
+Some important regional differences may have differentiated sites along the Salt River and Tonto Creek arms of Roosevelt Lake [@Lyons2013-ya; @Simon2001-am] based on non-local ceramics and obsidian.
+All of the artifacts/data from the RPMS project are hosted at the Center for Archaeology and Society at Arizona State University, and the data in this study come exclusively from this project.
 
@@
 
-Most of the occupation of the Tonto Basin, particularly the sites excavated during the project, date between AD 1275 and 1325, which corresponds to the Roosevelt Phase.
-The Roosevelt phase is notable for the appearance of Salado pottery.
-Moderate occupation continued into the Gila phase (AD 1325-1450), after which recognizable occupation ceased in the area until the historic period.
+Tonto Basin was occupied, although likely not continuously, from the Archaic through the late Classic period.
+The height of occupation, particularly the sites excavated during the project, dates between AD 1275 and 1325, which corresponds to the Roosevelt Phase.
+The Roosevelt phase is notable for the appearance of Salado pottery and platform mounds--probably introduced from the Phoenix Basin [@Elson1996-ym].
+Moderate occupation continued into the Gila phase (AD 1325-1450), which was characterized by population loss and aggregation into larger sites.
+This phase marks the end of recognizable occupation in the area until the historic period.
 
-One aspect of Tonto Basin that has caught the attention of many researchers is its location in a border zone between the Hohokam, Mogollon, and Ancestral Pueblo regions.
-Of particular note is the presence of masonry roomblock architecture and pottery uncharacteristic of the local Hohokam traditions, which has been interpreted as Kayenta immigration into Tonto Basin and the beginning of the Salado phenomenon [@Clark2001-ck; @Lyons2003-yy; @Lyons2006-to; @Stark1998-mu].
-Salado pottery production was widespread, but often the largest sources of production was centered at the location of a former Kayenta enclave [@Hill2015-at; @Huntley2016-em; @Lyons2012-pr; @Neuzil2008-zd].
+One aspect of Tonto Basin that has caught the attention of researchers is its location in a border zone between the Hohokam, Mogollon, and Ancestral Pueblo regions [e.g., @Caseldine2022-uu; @Clark2001-ck; @Elson1994-hk; @Hill2015-at; @Huntley2016-em; @Lyons2012-pr; @Lyons2013-ya; @Neuzil2008-zd; @Wood2000-ze].
+Of particular note is the presence of masonry roomblock architecture and pottery uncharacteristic of the local Hohokam traditions, which has been interpreted as Kayenta immigration into Tonto Basin [@Clark2001-ck; @Lyons2003-yy; @Lyons2006-to; @Stark1998-mu].
+This migration into the Tonto Basinwas part of a larger migration from north to south and is associated with the origin of the Salado phenomenon.
+Salado pottery production was widespread across southern Arizona, but often the largest sources of production was centered at the location of a former Kayenta enclave [@Hill2015-at; @Huntley2016-em; @Lyons2012-pr; @Neuzil2008-zd].
+The connection between immigration and an important type of decorated pottery in the region suggests that sites with roomblocks may have higher centrality in ceramic networks.
+Having high centrality means they have more connections to other nodes (i.e., sites).
+The presence of migrant communities in Tonto Basin invites questions regarding the relationship between sites--particularly between migrant and local communities.
+A network approach is an ideal way to examine the relationships between sites.
+Evidence that projectile points move between sites [e.g., @Watts2013-ub] indicates that this is also a productive line of analysis, and one that is less-commonly pursued compared to architecture and ceramics.
 
+# Network Analysis {-}
+
+Network analysis is a flexible tool for analyzing many types of data [see @Brughmans2023-uj].
+The key element is that some way must be determined to tie each entity together.
+This analysis uses sites as nodes in the network and some type of similarity as a tie to connect each node.
+Each type of similarity depends on the type of data.
+Table `r tbls` lists the `r nrow(sites)` sites with sufficient data to include in this study with the types of architecture present, periods occupied, and total number of ceramics and complete projectile points found at each site.
+The lack of complete projectile points at some sites is the largest caveat in this study, but complete points were necessary for the type of projectile point analysis used.
+The periods, architecture types, and ceramic types were assigned during the RPMS project and can be found in the project reports [@Rice1998-ku].
+This is not a random sample of Tonto Basin sites.
+It is merely a sample of convenience.
+Sampling can have a major impact on conclusions.
+In this case, I am not drawing conclusions based on a complete network or the existence of a random sample (though biases can still affect the results).
+Rather, I am interested in whether the types of material culture present in this study co-vary with each other and with space.
+These questions do not require a complete network.
+
+## Defining Networks and Relations
+
+Network approaches are best suited where the relationships between entities are of primary interest.
+In this case, the entities are defined as archaeology sites.
+The archaeology sites in turn represent a group of individuals who lived or visited the area of the site and left material remains behind.
 Ideally, studying social relations would be approached at the individual level, but rarely can archaeologists address archaeological data with that level of specificity.
-Watts [-@Watts2013-ub] used projectile points from Tonto Basin in just such a study, however.
-His study of flake scar patterning identified individual knappers, or at least clusters of knappers with similar knapping styles, and the distribution of their projectile points around Tonto Basin.
+Watts [-@Watts2013-ub] used projectile points from Tonto Basin in just such a study.
+His analysis of flake scar patterning identified individual knappers, or at least clusters of knappers with similar knapping styles, and the distribution of their projectile points around Tonto Basin.
 He found strong connections between many parts of the eastern Tonto Basin.
+Unfortunately, this study will only attempt a site-level analysis.
+Watt's study does, however, illustrate how relations can be defined between nodes in a network analysis.
+He used similarity networks where the points that had similar flaking styles were connected together.
+Similarity networks are a commonly used type of network in archaeology [e.g., @Birch2018-xx; @Borck2015-jy; @Cochrane2010-oa; @Golitko2015-we; @Lulewicz2019-lu; @Mills2013-wq; @Peeples2018-ib; @Terrell2010-hd].
+These studies use a variety of artifacts and methods to construct their networks, but each has demonstrated the utility of network methods within archaeological contexts.
+This study also uses similarity networks to group sites by similarity in ceramic assemblages and similarity in projectile point forms.
 
-One way to examine individuals who make similar types of projectile points--and similar types of pottery for that matter--is to talk about identity.
+Yet, it can be difficult to understand how a similarity network is related to a past social network.
+What kinds of interactions are represented by similarities in projectile points or in ceramics?
+Answering this question is also a crucial step in interpreting networks.
+One way to examine relations between individuals who make or use similar types of projectile points or pots is to talk about identity.
 Identity can be a troubling topic for anthropologists.
@@
 Pottery makes a good example.
-A potter can produce large, black and red chevrons on a pot that everyone who sees it can associate with a particular social group.
-Members of that group may wish to use that pottery as well as those who wish to curry favor with that group.
-Others may avoid that pottery.
-This pottery would be a marker of categorical identity.
+Peeples [-@Peeples2018-ib, p.
+151] notes that bowls in the Cibola region were sometimes painted with a bright, red slip with designs painted on the interior and they were among the first in the Southwest to have designs on the exterior.
+This is a strong indication of categorical identity.
+The potters were attempting to make a clear signal for whoever saw the pot.
 Relational identity can be seen in the way the potter prepares the clay recipe and smooths the coils.
@@
 The architecture discussed is highly visible and representative of historical group membership.
-The ceramics are grouped into types based primarily on decoration, which is strongly indicative of group membership [see @Clark2001-ck for a discussion on material culture and its relationship to identity].
-Projectile points can in some cases indicate group membership, but in this case the triangular and side-notched points were likely used for different purposes [see @Loendorf2015-ww].
-Styles were difficult to determine for the original researchers, and I had the same difficulty.
+The ceramics are grouped into types based primarily on decoration, which is strongly indicative of group membership.
+What this means is that links between nodes in the cer The details of architecture and ceramics used in this study are discussed in a future section, but it is important to note that these designations as categorical or relational are contextually dependent on this study.
+Clark [-@Clark2001-ck] has an excellent discussion on material culture and its relationship to identity.
+He identifies certain patterns, but his synthesis and other research [@Carr1995-mq; @Carr1995-sh; @Dietler1998-zo; @Gosselain1998-qd; @Gosselain2000-wf; @Gosselain2016-as; @Hodder1982-no; @Huntley2008-qu; @Lemonnier1986-le; @Lyons2003-yy; @Neuzil2008-zd; @Sassman2001-sw; @Stark1998-mu; @Wiessner1983-ei; @Wiessner1997-su] strongly indicates that the relationship between material culture and identity is culturally relative.
+Projectile points can in some cases indicate group membership [@Wiessner1983-ei; @Wiessner1997-su], but in this case the triangular and side-notched points were likely used for different purposes [see @Loendorf2015-ww; @Sliva2002-oz].
+Styles were difficult to determine for the original researchers [@Rice1994-rk], and I had the same difficulty.
 In my opinion, the subtle differences between projectile point outlines are more representative of interactions between knappers than markers of group identity for this particular case.
-There are no environmental reasons to assume differences in projectile points.
-Each hunter would have been using the points for the same game.
+This is indicative of relational identity.
+There are no environmental reasons to assume differences in projectile points, as each site was in a similar ecozone.
+Each hunter would have been using the points for the same game and differences in point styles would have been primarily cultural [@Sliva2002-oz].
 
-Another way that people identify themselves is gender.
-Gender roles, like identity, can also vary, but for the purposes of this analysis I will use a simplified model.
+There is one other aspect to these relations that I believe played an important role.
+A central way people identify themselves is by gender.
+Gender roles, like identity, can also vary in complex ways, but for the purposes of this analysis I will use a simplified model.
 Women made pottery and men made projectile points.
-This was not always true of course, but it generally fits the ethnographic record in the Southwest.
-This does not mean that a man could not move pots from one site to another or that a woman could not do the same with a projectile point, but I will make the assumption that ceramic networks and point networks are gendered.
-Therefore differences in these networks represent different interactions between men's and women's networks.
+This was not always true of course, but this fits the available data for the Hohokam [@Crown1996-xb; @VanPool2010-im, p.
+253].
+An examination of the burial record for the RPMS study shows no obvious indication that women made pottery, only that both men and women were buried with pottery [@Loendorf1998-ln, Table 10.7].
+On the other hand, projectile points were almost exclusively buried with males.
+This does not mean that a man could not move pots from one site to another or that a woman could not do the same with a projectile point, but, in general, differences in point and ceramic networks are most likely to indicate differences in gender networks.
 
@@
 
-The presence of migrant communities in Tonto Basin invites questions regarding the relationship between sites--particularly between migrant and local communities.
-A network approach is an ideal way to examine the relationships between sites.
-Evidence that projectile points move between sites [@Watts2013-ub] indicates that this is also a productive line of analysis, and one that is less-commonly pursued compared to architecture and ceramics.
-
-# Network Analysis {-}
-
-Network approaches are best suited where the relationships between entities are of primary interest.
-In this case, I compare the similarity between sites using architecture, ceramics, and projectile points.
-Thus, a network can be defined using a site from Tonto Basin as a node and a type of similarity as a link (or edge) between the nodes.
-Similarity networks are a commonly used type of network in archaeology [e.g., @Birch2018-xx; @Borck2015-jy; @Cochrane2010-oa; @Golitko2015-we; @Lulewicz2019-lu; @Mills2013-wq; @Peeples2018-ib; @Terrell2010-hd].
-These studies use a variety of artifacts and methods to construct their networks, but each has demonstrated the utility of network methods within archaeological contexts.
-
-Table `r tbls` 1 lists the `r nrow(sites)` sites with sufficient data to include in this study with the types of architecture present, periods occupied, and total number of ceramics and complete projectile points found at each site.
-The lack of complete projectile points at some sites is the largest caveat in this study, but complete points were necessary for the type of projectile point analysis used.
-The periods, architecture types, and ceramic types were assigned during the RPMS project and can be found in the project reports [@Rice1998-ku].
+The network for architecture, ceramics, and projectile points, were created using the Jaccard similarity coefficient [@Jaccard1912-na] to determine the links between each site.
+This is a simple measure of similarity that does not take into account abundance.
+It was chosen due to the large differences in sample sizes between sites for the ceramics and projectile points and because count data were irrelevant or not readily available for the architectural data.
 
@@
   select(Site = ArchaMapName,Area,Architecture = architecture,Phase,Ceramics,Points) %>%
+  # as_hux() %>% set_width(1) %>% set_number_format(everywhere,c(5,6),fmt_pretty()) %>% 
+  # theme_article() %>% 
+  # set_caption(cap)
   knitr::kable(caption= cap,
@@
 The types of similarity used in this study are varied depending on the type of network.
-One parameter that is consistent is how similar links should be in order to keep them.
+Part of this analysis is a visual approach and some of the methods require non-weighted links, thus I will be only using the strongest links.
 There is rarely a clear dividing line between similar and not similar.
 This can be a challenge for network analysis, because we can end up with networks where every node is connected to every other node.
-The decision to binarize a network--remove the weakest links and then cosider each link of equal value--has is drawbacks [@Peeples2013-tj], but is necessary in this case.
+The decision to binarize a network--remove the weakest links and then consider each link of equal value--has its drawbacks [@Peeples2013-tj], but is necessary in this case.
 A solution is to assign weights to each link that defines the strength of the tie.
@@
 A common approach is to keep only the strongest ties by either ranking the ties or using a cutoff value.
-Part of this analysis is a visual approach and some of the methods require non-weighted links, thus I will be only using the strongest links.
 
 I used ranked links to keep consistent values between networks.
-For example, a network could be made of the `r nrow(sites)` sites as nodes with each node being connected to the three sites that it has the strongest similarity to for ceramics, architecture, projectile points, and the three sites that it is physically closest to for the spatial network.
-In practice, not every node will have three connections.
-More than three ties can exist when the ties are tied in rank.
-Some nodes may not have three connections.
-And finally, because the network is not directed (meaning the ties indicate similarity in both directions) one node can have several ties pointing to it because that node was in the top three of multiple other nodes.
+This involved calculating the strength of similarity between each site in the network and then ranking the strength of similarity.
+The top **n** connections between each site were kept and the rest were discarded.
+In practice, not every node will have the same number of connections.
+Some may have fewer ties if there are not enough nodes that are similar.
+More ties can exist when there are ties in the ranks.
+Because the network is not directed (meaning the ties indicate similarity in both directions) one node can have several ties pointing to it because that node was in the top **n** of multiple other nodes.
 In the latter case, this is a good indication of a central node in the network.
-Because I am using several types of networks, there is no common cutoff value I can use.
-Instead of arbitrarily picking one value, I have calculated each metric using networks composed of the top three to ten ties.
+Because I am using several types of networks, there is no common cutoff value I can use for the number of ties to keep.
+Instead of arbitrarily picking one value, I have calculated each metric using networks composed of the top 3-10 ties.
+Meaning that each metric is calculated for a network composed of ties with the top 3 connections, and the process is repeated for new networks composed of ties with the top 4 connections, and so forth.
 Examining the range of these metrics provides a more robust analysis.
@@
 Nodes with higher centrality generally derive or generate greater benefits from the network.
+For example, if I have many friends, then I can call in more favors.
 Eigenvector centrality is one way to measure node centrality and is commonly used in archaeology.
 This metric describes how well a node is connected to the network as a whole and is helpful in comparing different networks containing the same nodes.
+For example, if my friends have many friends, then I will have a greater advantage then someone with an equal number of friends, but whose friends have few friends.
+Essentially, this is a way of measuring second-order and beyond connections.
 Eigenvector centrality is also more robust to missing nodes than other measures [@Peeples2017-kl], a major problem in most archaeological studies.
-
-## Multilayer Networks {-}
 
@@
 The combination of these individual networks is a multilayer network (also called a multiplex network).
-
 Multilayer networks allow for methods to be applied on multiple networks at once [see @Brodka2018-vz].
@@
 I will display the results of the Pearson correlation, but the Spearman correlation provided similar results.
+Eigenvector centrality was mentioned in the previous section, but it is also calculated as a multilayer eigenvector centrality.
+In this case the centrality measure is simply the mean of the eigenvector centraliy for each separate layer, as the layers are not interdependent [@Frost2022-lr].
+Multilayer eigenvector centrality results are only presented for the combination of ceramic and point layers.
+These networks can be thought of as dependent variables for the spatial and architectural networks--meaning spatial distance and architecture are considered to be important variables determining the structure of the ceramic and point networks.
+
+The robustness of results were tested by randomly sampling the underlying network matrices 10,000 times and comparing the results to the random samples to obtain a p-value.
+This provides a baseline that determines how likely it is to obtain a given result by chance.
+
+Certain caveats must be acknowledged.
+All archaeologists deal with missing data, which can greatly affect the results of analysis.
+Network analysts have grappled with this question and often use resampling or similar methods to deal with this problem [@Bischoff2021-sf; @Bolland1988-zm; @Borgatti2006-jl; @Brughmans2023-uj; @Costenbader2003-st; @Galaskiewicz1991-qb; @Gjesfjeld2015-hw; @Lee2009-wr; @Rivera-Hutinel2012-ik].
+Regardless of statistical methods, unaccounted for biases will still produce invalid results [@Bischoff2021-sf].
+The most important bias that may affect these findings is chronology.
+For example, Bass Point Mound and Cline Terrace Mound were probably not occupied at the same time [@Jacobs1997-ob; @Lindauer1995-wq].
+Does that invalidate the network analysis?
+I do not think so.
+While the networks involve nodes as sites, what the networks are really meant to represent are groups of people.
+Furthermore, they are not individuals, but the aggregate decisions of people over more than one generation.
+Current models of Hohokam social groups suggest a certain residential stability [@Craig2017-rg, p.
+336].
+Perhaps the residents of these sites were not physically located at the site at the same time as the other inhabitants were located at their site, but they may have been at another nearby location.
+The spacing of settlements may indicate a corporate form of social organization [@Clark2004-uw; @Rice1998-rz, p.96].
+If this is accurate, then the communities inhabiting the sites may likely have existed in some form prior to the establishment of any particular site and have continued on.
+It is these communities that the network analysis is attempting to capture.
+Still, evidence indicates that most sites were contemporaneous.
+For example, Cline Terrace Mound and Schoolhouse Point were most probably occupied at the same time [@Lyons2013-ya].
+It is best, though, to think of these interpretations as estimations based on incomplete data.
 
@@
 This network was created by calculating the Euclidean distance from every site to every other site.
-This method is equivalent to proximal point analysis, which was used in one of the earliest network studies in archaeology [@Terrell1977-co] and continues to be used to study potential pathways of interaction [see @Broodbank2000-pg; @Collar2013-fb].
+Euclidean distance is a shortcut that does not represent actual travel routes.
+Least cost paths provide more accurate data [see @Caseldine2022-uu for an application in Tonto Basin], yet an analysis in similar terrain indicates that for distances longer than those in this study the benefits of least cost path distances in place of Euclidean distance are negligible [@Bischoff2017-gx].
+Least cost paths are also hampered by the presence of Roosevelt Lake and other modern impacts on the landscape.
+The spatial network created here is equivalent to proximal point analysis, which was used in one of the earliest network studies in archaeology [@Terrell1977-co] and continues to be used to study potential pathways of interaction [see @Broodbank2000-pg; @Collar2013-fb].
+The inclusion of a spatial network serves as a null hypothesis to test the other networks against.
 
@@
 The architectural components consisted of compounds, roomblocks, and platform mounds.
+Compounds were the typical residential structures in the area and featured multiple houses built from masonry and/or adobe around a courtyard.
+Platform mounds were formed from large masonry and adobe retaining walls with a cell-like interior filled with trash and rubble.
+They were common in the Hohokam region throughout the Classic period.
 Some sites had pit houses from earlier occupations, but most excavation was focused on the Roosevelt and Gila phases so these few pit houses were removed from the analysis [see @Rice1998-ku].
@@
 Thus, if a site is labeled a platform mound it may have other architecture.
-The network for architecture, and for the remaining networks, was created using the Jaccard similarity coefficient [@Jaccard1912-na] to determine the links between each site.
-This is a simple measure of similarity that does not take into account abundance.
-It was chosen due to the large differences in sample sizes between sites for the ceramics and projectile points and because count data were irrelevant or not readily available for the architectural data.
+Sites were thus labeled to emphasize the most distinctive form of architecture, but all types of architecture were present in the analysis, and all architectural features included in the analysis can be seen in Table 1.
 
@@
 The ceramic network consists of 55 types--plainware and decorated--ranging from 181,666 Plain Brown Ware sherds to a single Black Mesa Black-on-white sherd.
-A single sherd is not much, but is still indicative of trade.
-However, the median sherd count was 157 for all types.
+A single sherd is not much, but it is still indicative of trade or movement.
+The median sherd count was 157 for all types.
 The ceramic data was accessed from the cyberSW database [@Mills2020-tb], which keeps the full citations for the ceramic data.
+The standard caveats for ceramic analysis apply to this analysis as well: problems with sherd misidentification and data errors.
 
@@
 
-The ceramic and architectural data is available in suitable formats for analysis, but the original projectile point analysis was too general for the purposes of this study [see @Rice1994-rk for details].
-Geometric morphometrics is a set of methods designed to quantitatively analyze shapes and is ideal for projectile point analysis.
-See Bischoff [-@Bischoff2022-rg] for details on geometric morphometrics and the methods used to analyze these projectile points.
-A geometric morphometric analysis was conducted on the 2D outlines of each complete triangular or side-notched projectile point.
+The ceramic and architectural data are available in suitable formats for analysis, but the original projectile point analysis was too general for the purposes of this study [see @Rice1994-rk for details].
+Points were divided into longer or shorter categories and subdivided by several attributes (e.g., blade, base, or notch shape).
+Geometric morphometrics (GM) is a set of methods designed to quantitatively analyze shapes and is ideal for projectile point analysis.
+See Bischoff [-@Bischoff2022-rg] for details on GM and the methods used to analyze these projectile points.
+A GM analysis was conducted on the 2D outlines of each complete triangular or side-notched projectile point.
 The majority of projectile points were triangular or side-notched.
@@
 Figures `r figs` and `r figs + 1` show the results of these analyses.
+These figures demonstrate how GM methods can calculate the similarity between each projectile point shape.
 These results were classified into closely related clusters and used in the same way as ceramic types.
@@
 paste0(nodes$id,": ",nodes$ArchaMapName) %>% paste(collapse = "; ")
+
 meanEigenVals = eigenDF %>% 
-  filter(name %in% c("ceramics","points")) %>% 
-  group_by(Architecture,name) %>% 
+  group_by(id,Area,Architecture,name) %>% 
   summarize(value = round(mean(value),2)) %>% 
   arrange(name) %>% 
-  pivot_wider(names_from = name, values_from = value)
+  pivot_wider(names_from = name, values_from = value) %>% 
+  mutate(id = as.integer(id)) %>% 
+  left_join(nodes %>% select(id,Site = ArchaMapName)) %>% 
+  ungroup()
+
+meanEigenValsSites = meanEigenVals %>% 
+  select(Site,Area,Architecture,ceramics,points) %>% 
+  rowwise() %>% 
+  mutate(Centrality = round(mean(c(ceramics,points)),2)) %>% 
+  select(-ceramics,-points)
+
+# # resampling analysis
+# # purpose reshuffle the points and ceramics to determine if the correlations can appear by chance
+# library(furrr)
+# future::plan('multisession')
+# sensitivityResults = furrr::future_map(1:10000, function(s){
+#   corr = list()
+#   eigen = list()
+#   for (nLinks in 3:10){
+#     net = ml_empty("Tonto")
+#     g2List = list()
+#     nms = names(dists)
+#     for(n in nms){
+#       m = dists[[n]]
+#       m = matrix(sample(m,121, replace = T),ncol = 11)
+#       g = makeGraph(m, nLinks = nLinks)
+#       eigen[[paste0(nLinks,"-",n)]] = V(g)$centrality
+#       g2List[[n]] = g
+#       add_igraph_layer_ml(n = net, g = g,name = n)
+#     }
+#     corr[[nLinks]] = layer_comparison_ml(net, method = "pearson.degree", mode = "all") %>% as.data.frame() %>%
+#       mutate(nLinks = nLinks)
+#   }
+# 
+#   corrDF = bind_rows(corr) %>%
+#     rownames_to_column("from") %>%
+#     pivot_longer(-c(from,nLinks), names_to = "to") %>%
+#     mutate(from = str_remove_all(from,"[^A-z]")) %>%
+#     mutate(nLinks = factor(as.character(nLinks))) %>%
+#     rowwise()  %>%
+#     filter(from != to) %>%
+#     mutate(id = paste(sort(c(from,to)),collapse = "->")) %>%
+#     select(-from,-to) %>%
+#     distinct_all() %>%
+#     mutate(run = s)
+# 
+#   eigenDF = bind_rows(eigen) %>%
+#     rowid_to_column("id") %>%
+#     pivot_longer(-id) %>%
+#     left_join(nodes) %>%
+#     mutate(name = str_remove_all(name,"[^A-z]"),
+#            id = factor(id),
+#            run = s)
+#   return(list(corrDF = corrDF, eigenDF = eigenDF))
+# })
+# sensitivityCorr = map_df(sensitivityResults,function(i)bind_rows(i$corrDF))
+# saveRDS(sensitivityCorr,"data/SensitivityCorr.Rds")
+sensitivityCorr = readRDS("data/SensitivityCorr.Rds")
+# sensitivityEigen = map_df(sensitivityResults,function(i)bind_rows(i$eigenDF))
+# saveRDS(sensitivityEigen,"data/sensitivityEigen.Rds")
+sensitivityEigen = readRDS("data/sensitivityEigen.Rds")
+pF = function(val,var, df = "sensitivityEigen"){
+  if(df == "sensitivityEigen"){
+  df = get(df)
+  sub = df %>% filter(name %in% var)
+  ecdfF = ecdf(sub$value)
+  round(ecdfF(val),2)
+  }else{
+  df = get(df)
+  sub = df %>% filter(id %in% var)
+  ecdfF = ecdf(sub$value)
+  round(ecdfF(val),2)
+  }
+}
+# pF(.95,"ceramics",df = "sensitivityEigen")
+# pF(.35,"architecture->ceramics",df = "sensitivityCorr")
 ```
+
+A major benefit of network analysis is the visual exploration of the network.
+The results of the network analysis are provided partially to address the major questions of this paper--the correlations between types of material culture and their potential causes--and partially as a visual exploration with corresponding evidence from network metrics.
 
@@
 Fewer ties leaves the Cline Mesa sites disconnected from the rest of the network.
-Thus, Bass Point Mound forms a crucial bridge between the Cline Mesa sites and the rest of the network.
-The Rock Island area consists of a single site, Bass Point Mound, which was a platform mound, although it was not heavily excavated.
-If spatial distance was a primary driver in social interaction, then we would expect Bass Point Mound to consistently be a highly central node.
+Bass Point Mound forms a crucial bridge between the Cline Mesa sites and the rest of the network.
+Because the network is not complete, one cannot argue that there were not other sites between Cline Mesa and the Schoolhouse Mesa sites, but it still represents an intermediary location.
+Its geographic position overlooking the confluence of the Salt River and Tonto Creek would make an ideal meeting place for parties coming down from the Tonto Creek arm of what is now Roosevelt Lake or coming up from the Salt River arm.
+The Rock Island area consists of a single site, Bass Point Mound (a platform mound), although it was not heavily excavated.
+If spatial distance was an important factor in social interaction, then we would expect Bass Point Mound to consistently be a highly central node due to its central location.
+Table `r tbls` shows the multilayer eigenvector centrality for points and ceramics only.
+Bass Point Mound has the highest multilayer network centrality with 0.95 (p = `r pF(.95,c("ceramics","points"),df = "sensitivityEigen")`).
+Figure `r figs + 1` provides the eigenvector data by type of network.
+Bass Point Mound has low centrality in the architectural network and the networks that do not connect the spatial components (three and four closest connections), which gives it a lower spatial centrality as well.
+The ceramic and point networks provide strong evidence that Bass Point Mound's spatial location was advantageous for forming connections between the sites in this study.
 
@@
 
-I chose eigenvector as the centrality measure (see figure `r figs`), which you can see in the spatial eigenvector results poorly reflects the centrality of Bass Point Mound.
-Betweenness centrality would better capture how objects would have to flow through Bass Point Mound to get from the Cline Mesa sites to the other parts of the network, but this is not a road network.
-People could have simply walked around Bass Point Mound.
-What eigenvector centrality does indicate is whether it is connected to other strongly connected sites.
-This may not be important for a spatial network, but it is important for the other networks.
-Bass Point Mound's centrality is very low for architecture.
-It is the most central site for points based on eigenvector analysis and visual inspection and it is high for ceramics as well.
-This indicates a strong correlation with space and points, at least for the Bass Point Mound, and that Bass Point Mound was also important for Ceramic circulation (whether that is actual pots or the ideas of how to make pots).
+```{r meanEigen, echo=FALSE}
+tbls = tbls + 1
+cap = "Multilayer Eigenvector Centrality for Ceramics and Points"
+meanEigenValsSites %>% 
+  select(Site,Centrality) %>% 
+  arrange(desc(Centrality)) %>% 
+  knitr::kable(caption= cap,
+               booktabs = T,
+               format.args = list(big.mark = ",",
+                                  scientific = FALSE))
+  # as_hux() %>% 
+  # theme_article() %>% 
+  # set_caption(cap)
+
+```
+Visual inspection of the network graphs provides several other observations.
+One is that the four geographic areas represented do not form strong components in the non-spatial networks like they do in the spatial network.
+The architecture is thoroughly mixed, and the ceramics and points form single component networks.
+The formation of single components--rather than distinct sections of the graph--is typical of artifact similarity networks.
+The Livingston sites exhibit the most clustering in the ceramic network, although some of the connections go through Cline Mesa sites, which are as geographically distant as possible for these sites.
+The Schoolhouse Mesa sites always group together in the ceramic and point networks, as do the Livingston compound sites and separately the Livingston roomblock and platform mound.
+The Cline Mesa sites exhibit more variability.
 
@@
 ```
-
-Bass Point Mound is not the only bridge node.
-Pinto Point Mound connects the platform mounds with the compounds in the architecture network.
-The reason for this is that the site also has a compound present.
-Similar to Bass Point Mound, eigenvector centrality does not favor bridge nodes, but Pinto Point Mound has the second highest centrality value for ceramics and the highest for spatial distance (meaning it is located near many other sites), and it has an above average centrality for points.
-This suggests architecture is an important component in the ceramic network and possibly in the point network.
-
-Besides the two bridge nodes, visual inspection of the graphs provides several other observations.
-One is that the four geographic areas represented do not form strong components in the non-spatial networks like they do in the spatial network.
-The architecture is thoroughly mixed, and the ceramics and points form single component networks.
-This is typical of artifact similarity networks.
-The Livingston sites exhibit the most clustering in the ceramic network, although some of the connections go through Cline Mesa sites, which are as geographically distant as possible for these sites.
-The Schoolhouse Mesa sites always group together in the ceramic and point networks, as do the Livingston compound sites and separately the Livingston roomblock and platform mound.
-The Cline Mesa sites exhibit more variability.
 
@@
 It is expected that roomblocks have high centrality in the ceramic network, but not that platform mounds should have the lowest centrality.
+Clearly, centrality in the projectile point network does not equal centrality in the ceramic network.
 
@@
 cap = "Mean Eigenvalues for Ceramic and Point Networks by Type of Architecture"
-meanEigenVals %>% 
+eigenDF %>% 
+  filter(name %in% c("ceramics","points")) %>% 
+  group_by(name,Architecture) %>% 
+  summarize(value = round(mean(value),2)) %>% 
+  pivot_wider(names_from = name, values_from = value) %>% 
+  ungroup() %>% 
+  rowwise() %>% 
+  mutate(mean = round(mean(c(ceramics,points)),2)) %>% 
+  arrange(desc(mean)) %>%
+  # as_hux() %>% 
+  # theme_article() %>% 
+  # set_caption(cap)
   knitr::kable(caption= cap,
@@
 
-There is a chain of association between immigrants and roomblocks, roomblocks or locations with roomblocks as centers of Salado production, and Salado pottery as a widespread phenomenon that provides an expectation for roomblock sites to be central to pottery networks.
-This expectation is born out in this network analysis.
-Two of the platform mounds and three of the compound sites also have high centrality, demonstrating that ceramic circulation occurred between groups at sites with all types of architecture.
-What is clear from this analysis is that centrality in the ceramic network does not correlate with centrality in the point network.
-
 The visual analysis, combined with the eigenvector centrality, indicated some spatial and architectural correlation.
@@
 This analysis provides a clear contrast between layers.
-Only one comparison had a strong correlation--the point and spatial networks were strongly correlated.
-The visual inspection and centrality analysis provided some indication of this, but the multilayer network comparison provides a clearer indication.
-All other layer comparisons had an approximately equivalent, positive correlation, but with only a week strength.
+None of the results, on average, showed a negative correlation--meaning that ties existed in one network where ties did not exist in the other network.
+Only one comparison had a strong correlation--the point and spatial networks were strongly correlated (**x** = `r psCorr = corrDF %>% filter(id == "points->spatial") %>% pull(value) %>% mean %>% round(.,2); psCorr`; p = `r pF(psCorr,"points->spatial",df = "sensitivityCorr")`).
+The visual inspection and centrality analysis provided some indication of this, but the multilayer network comparison provides strong verification.
+All other layer comparisons had an approximately equivalent, positive correlation, but with only a weak strength.
 The one exception was architecture and points where the correlation was approximately zero.
-The visual inspection and eigenvector analysis demonstrated some interesting interactions with architecture in the network, but the overall correlation demonstrates that architecture cannot be used to predict the presence of absence of network ties in the projectile point network.
+The visual inspection and eigenvector analysis demonstrated some interesting interactions with architecture in the network, but the overall correlation demonstrates that architecture cannot be used to predict the presence or absence of network ties in the projectile point network.
 
@@
 This analysis has three main findings regarding networks in Tonto Basin: (1) that point networks correlate with space; (2) that roomblock sites are highly central in the ceramic network and have low centrality in the point network; and (3) that ceramic and point networks are significantly different from each other.
-The purpose of this analysis is to understand something about the social relations in Tonto Basin.
-Presumably, these networks indicate something about how people interacted, and the ceramics and projectile points reveal something about the identity of those who made, used, and discarded them.
+In terms of the null hypotheses stated in the introduction, only one of the networks exhibited strong spatial correlation and while all but one pairing (architecture and points) exhibited some positive correlation, all of the correlations except the points and space were weak.
+Given these network results, how might they be interpreted in terms of social behavior?
+
+Exchange would be a primary driver in these network dynamics, and clear evidence exists for exchange of several types of material culture.
+Trade within the basin does not appear to have been dominated by any one group and may have been competitive in nature [@Rice1998-cu, p.
+127-128].
+Ceramic exchange certainly occurred at a regional level, but even local ceramics were circulated within the basin [@Heidke2000-nr; @Miksa1995-qw; @Stark1992-vu], possibly in return for food [@Clark2004-uw, p.290-291].
+Projectile points (either as part of the arrow or separately) may have also been commonly exchanged.
+There are several examples of the exchange of bows and arrows ethnographically throughout the world [e.g., @Mauss1966-nm; @Nishiaki2013-lc; @Wiessner1983-ei], in North America generally [e.g., @Hoffman1896-xl; @Radin1923-gf], and in the Southwest specifically [e.g., @Beaglehole1936-ul; @Dittert1959-rz; @Fewkes1898-oj; @griffen1969a; @Parsons1939-xg; @Simpson1953-ob].
+Besides Watt's study of knappers, Sliva's [-@Sliva2002-oz, p.
+539] analysis of sites from the Roosevelt Community Development Study suggests craft specialization in projectile point production.
+Mortuary offerings were possibly the intended function for many points [@Sliva2002-oz, p.
+543], which may have spurred the increased craft specialization.
+Shell and stone jewelry, among other artifacts were also produced and exchanged within the basin.
+A potential loci for exchange are the platform mound sites.
+The purpose of these sites is still debated, but many argue that they served an integrative function [e.g., @Abbott2006-fg; @Adler1990-ws; @Clark2004-uw; @Craig1995-qo; @Craig1994-bg, p.
+112-165; @Doelle1995-qd, p.
+439].
+If platform mounds were places where exchange between communities regularly took place, then we would expect platform mounds to have higher centrality.
+Centrality varied among the platform mounds, but on average, platform mounds did have the highest centrality.
+
+As discussed previously, cultural differences were expected between the Tonto Creek and Salt River arms of Roosevelt Lake.
+This was due in part to differences in long-distance ceramic exchange and obsidian sourcing [@Lyons2013-ya; @Simon2001-am].
+This analysis did not use obsidian or focus on non-local ceramics and therefore did not capture differences based on these factors.
+There was some expectation that more separation in the material networks would be apparent between the Cline Mesa sites in the Tonto Creek arm and the other sites (minus Bass Point Mound that lies at the confluence).
+This was not born out in the visual inspection of the analysis with significant mixing of the areas in the network.
+There were, however, indications that spatial distance was an important factor.
+Spatial distance would align with the proposed differences in the Tonto Creek and Salt River arms of the lake.
+The only strong correlation between networks was the point and spatial network, which indicates that being near another site is a good indication of the shape of a projectile point.
+There was however, little correlation overall between ceramics and spatial distance.
+
+There is a potential chain of association between immigrants and roomblocks, roomblocks or locations with roomblocks as centers of Salado production, and Salado pottery as a widespread phenomenon that provides an expectation for roomblock sites to be central to pottery networks.
+While this association is circumstantial and not expected to be uniform across the Hohokam region, roomblocks did have the highest centrality in the ceramic networks.
+Another reason for the high centrality of roomblock sites is that they lacked access to adequate farmland and had to trade pottery for food [@Clark2004-uw, p.
+291].
+This second explanation may help explain why roomblock sites had low centrality in the point networks, presuming that points were not exchanged for food.
+It is beyond this simple analysis to determine the precise reasons, but it is clear that roomblock sites were important for ceramic circulation in Tonto Basin.
+On the other hand, sites with roomblocks had the lowest centrality in the point networks.
+This suggests that the influence the occupants of roomblock sites had in pottery networks may not have extended to other spheres of interaction.
 
@@
 
-The point networks were expected to represent relational identification among men.
-This is not always the case with projectile points.
-In some cases they can represent group identity [see @Wiessner1983-ei], but in this case there is little variation in style.
+The point networks were expected to represent relational identification among men, at least in this case study.
 Because relational identification is related to frequent interaction, spatial distance is a crucial component.
@@
 
-The remaining question is whether the differences between the ceramic and projectile point networks are due to the differences in relational and categorical identification or are due to differences in behavior related to gender.
-I argue that in this case gender likely played an important role.
-It is most probable that potters were making the same types of pottery, rather than switching between styles.
-Not to say that potters were not making plain ware and decorated at the same time, but that they were not making radically different styles of decorated wares at the same time.
-Thus, the relational ceramic network likely follows the categorical network.
-This is however, an assumption that needs to be tested by examining sourcing data.
-The differences between the point and ceramic networks are large enough, that I am confident enough to hazard an opinion that these networks do represent gendered networks.
+Perhaps the reason roomblocks were more central to the ceramic network is because immigrants to the basin had less access to farmland and had to make pottery to get food, as mentioned.
+This would explain higher centrality for roomblocks and lower centrality for point networks.
+It does not explain the remaining differences between point and ceramic networks.
+Projectile points at least were highly gendered, and ceramics probably were as well.
+The evidence presented here demonstrates that ceramic and point networks vary significantly, and I believe gender likely played an important role.
 
@@
 
-This analysis used architectural, ceramic, projectile point, and spatial data from `r nrow(sites)` sites in Tonto Basin, a region in central Arizona.
+This is one of the first archaeological applications of multilayer network analysis to consider multiple types of material culture.
+This is advantageous for studying how types of material culture do or do not co-vary, which aids interpretations of the social interactions that created these patterns.
+This analysis used architectural, ceramic, projectile point, and spatial data from `r nrow(sites)` sites in Tonto Basin.
 The data come primarily from occupations between AD 1275 and 1325.
 A network was created for each type of data and combined into a multilayer network.
-Visual network analysis, eigenvector centrality, and multilayer network Pearson correlation was used to study and compare the networks.
+Visual network analysis, eigenvector centrality, and multilayer network Pearson correlation were used to study and compare the networks.
+The findings indicate that the projectile point network was strongly correlated with the spatial network--indicating that sites near each other were more likely to have similar projectile points.
+Furthermore, sites with roomblocks indicative of immigrants from the north and east of Tonto Basin were, on average, the most highly central sites in the ceramic network; however, they were the least central in the point network.
+Immigrants may have relied on pottery production to integrate with the local networks, but this relationship did not hold for projectile points.
+Finally, the results demonstrate major differences between the ceramic and point networks.
+This findings has major implications for network studies, because many rely on only a single artifact type.
+If different types of material culture do not co-vary regularly, then that indicates archaeologists must do more to include other types of material culture to better understand the complex social networks that existed in the past.
+Furthermore, projectile points and ceramics have strong associations with gender.
+Differences in the ceramic and point networks suggest differences existed between the social networks of each gender.
 
-The analysis found correlations between certain types of architecture and ceramic networks and between the point and spatial networks.
-These correlations are consistent with architecture and ceramics as markers of categorical identity--meaning these types of material culture were used as markers of belonging to a recognized social group.
-The correlation between projectile points and spatial location is also consistent with projectile point form as a marker of relational identity--meaning that the similarity between projectile points indicates frequent interaction.
-
-Generally, projectile points were used in the Southwest by men in hunting and warfare.
-Women are most often associated with making and transporting pottery.
-I suggest that the large differences found between the point and ceramic networks indicate that men and women had different social networks in Tonto Basin.
-The large differences in centrality between the sites with roomblocks in the ceramic networks versus the point networks is also a possible indication of another social reality.
-If these networks are indeed gendered, then might immigrant women have been more socially connected than the men?
-Perhaps the simplest explanation is that male immigrants made projectile points they were more familiar with and felt no need to adapt to local customs, whereas the women felt greater pressure to integrate.
-Although it seems it was the immigrants' traditions that became the norm for pottery.
-One point that may support the idea of a lower level of integration among the men is the idea of gift giving or exchange of projectile points.
-Watt's [-@Watts2013-ub] analysis found that projectile points made by individuals were found throughout the eastern Tonto Basin, which supports the exchange of these items.
-There are several examples of the exchange of bows and arrows ethnographically throughout the world [e.g., @Mauss1966-nm; @Nishiaki2013-lc; @Wiessner1983-ei], in North America generally [e.g., @Hoffman1896-xl; @Radin1923-gf], and in the Southwest specifically [e.g., Beaglehole1936-ul; @Dittert1959-rz; @Fewkes1898-oj; @griffen1969a; @Parsons1939-xg; @Simpson1953-ob].
-The frequent exchange of points would have increased the centrality of roomblock sites in the point networks.
-The fact that this did not happen may be an indication that male immigrant networks were not as well integrated as female networks centered around potters.
-Whichever is the case, this analysis demonstrates that multilayer network comparisons using different types of material culture is a profitable avenue of research and one that deserves further attention.
+This analysis uses a handful of sites from Tonto Basin.
+Much more data exists and the results discussed here would be greatly strengthened by including more sites and data.
+What would be perhaps even more useful would be to include sourced obsidian and ceramics.
+Networks created from artifacts where the origin is known can be used to more strongly infer various types of social interaction.
+This type of multilayer analysis would also be useful to conduct in other regions.
+It would be particularly useful for comparative studies to know in what circumstances different types of material culture do or do not co-vary.
 
@@
 Josh Watts, Matt Peeples, Melissa Powell, Chris Caseldine, and several volunteers assisted with this research.
+Chris Caseldine's comments greatly improved this paper.
+All mistakes are my own.
 The many individuals who contributed to the Roosevelt Platform Mound Study also deserve recognition.
@@
 
-The author declares that no conflicts of interest exist in relation to the content of the article.
+The author declares no conflicts of interest.
 

### 2022-12-28 manuscript.tex
finished updating paper based on Chris Caseldine's comments

--- 2022-12-28 manuscript.Rmd
@@
----
-title: Material Culture Networks in Tonto Basin
-output:
-  rticles::tf_article: null
-  bookdown::word_document2:
-    toc: true
-    number_sections: false
-    reference_docx: TF.docx
-author:
-  - name: Robert J. Bischoff
-    affil: a
-    email: rbischoff@asu.edu
-affiliation:
-  - num: a
-    address: School of Human Evolution and Social Change, Arizona State University, Tempe, AZ
-keywords: |
- American Southwest; network analysis; multilayer networks; Hohokam; projectile points; ceramics; computational archaeology; gender
-abstract: |
-  Network analysis has a strong foundation in Southwest archaeology, yet the combination of multiple types of artifacts in one analysis--multilayer network analysis--has not been formally applied except within a single artifact type.
-Many studies consider material culture holistically, yet network analysis has the advantage of focusing specifically on the relationships between entities (often communities) and how the structure of a network can benefit particular entities.
-This study takes architecture, ceramic, projectile point, and site location data from the Roosevelt Platform Mound Study and combines these data into a multilayer network analysis.
-This analysis provides a way to test the co-variance of these types of material culture with each other and with spatial variation.
-The results demonstrate a strong association between projectile points and spatial distance.
-Sites containing roomblock architecture are highly central in the ceramic network but have low centrality in the projectile point network.
-Overall, the ceramic and projectile point networks exhibit significant differences.
-This indicates that the social networks that created these patterns had different social mechanisms.
-One potential cause of these differences is gendered spheres of interaction with men exchanging projectile points and women exchanging ceramics.
-bibliography: references2.bib
-link-citations: yes
-editor_options:
-  chunk_output_type: inline
-header-includes: |
+% interactcadsample.tex
+% v1.03 - April 2017
+
+\documentclass[]{interact}
+
+\usepackage{epstopdf}% To incorporate .eps illustrations using PDFLaTeX, etc.
+\usepackage{subfigure}% Support for small, `sub' figures and tables
+%\usepackage[nolists,tablesfirst]{endfloat}% To `separate' figures and tables from text if required
+
+\usepackage{natbib}% Citation support using natbib.sty
+\bibpunct[, ]{(}{)}{;}{a}{}{,}% Citation support using natbib.sty
+\renewcommand\bibfont{\fontsize{10}{12}\selectfont}% Bibliography support using natbib.sty
+
+\theoremstyle{plain}% Theorem-like structures provided by amsthm.sty
+\newtheorem{theorem}{Theorem}[section]
+\newtheorem{lemma}[theorem]{Lemma}
+\newtheorem{corollary}[theorem]{Corollary}
+\newtheorem{proposition}[theorem]{Proposition}
+
+\theoremstyle{definition}
+\newtheorem{definition}[theorem]{Definition}
+\newtheorem{example}[theorem]{Example}
+
+\theoremstyle{remark}
+\newtheorem{remark}{Remark}
+\newtheorem{notation}{Notation}
+
+% tightlist command for lists without linebreak
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+
 \usepackage{hyperref}
@@
 \def\UrlBreaks{\do\/\do-\do?}
-csl: american-antiquity.csl
-always_allow_html: true
----
 
-```{r setup, echo = F}
-out.format <- knitr::opts_knit$get("out.format")
-# options(knitr.table.format = "latex")
-knitr::opts_chunk$set(
-  echo = F,
-  message = FALSE,
-  warning = FALSE,
-  dpi = 450,
-  out.width = "100%"
-)
-```
+\usepackage{booktabs}
+\usepackage{longtable}
+\usepackage{array}
+\usepackage{multirow}
+\usepackage{wrapfig}
+\usepackage{float}
+\usepackage{colortbl}
+\usepackage{pdflscape}
+\usepackage{tabu}
+\usepackage{threeparttable}
+\usepackage{threeparttablex}
+\usepackage[normalem]{ulem}
+\usepackage{makecell}
+\usepackage{xcolor}
 
-```{r load, include=FALSE}
-library(tidyverse,quietly = T)
-library(magrittr)
-library(Momocs,quietly = T)
-library(kableExtra)
-library(huxtable)
-library(ggthemes, quietly = T)
-library(here)
-library(rio)
-library(igraph)
-library(ggnetwork)
-library(vegan)
-library(multinet)
+\begin{document}
 
-set.seed(1010)
+\articletype{}
 
-# library(ggimage)
-i_am("manuscript.Rmd")
-# keep track of figures and tables
-figs = 1
-tbls = 1
-set.seed(1010)
+\title{Material Culture Networks in Tonto Basin}
 
-# load data and remove sites with only one projectile point
+\author{\name{Robert J. Bischoff$^{a}$}
+\affil{$^{a}$School of Human Evolution and Social Change, Arizona State
+University, Tempe, AZ}
+}
 
-sites = import(here("data/TontoBasinNetworkSites.xlsx"),setclass = 'tibble') %>% 
-  ungroup() %>% 
-  filter(!ArchaMapName %in% c("AZ V:5:130 (ASM)","AZ U:4:009 (ASM)","AZ U:8:318 (ASM)","AZ V:5:121 (ASM)"))
+\thanks{CONTACT Robert J.
+Bischoff. Email: \href{mailto:rbischoff@asu.edu}{\nolinkurl{rbischoff@asu.edu}}}
 
-# data frame for network nodes
-nodes = sites %>%
-  select(ArchaMapName,Area,Architecture = architecture) %>% 
-  mutate(Architecture = case_when(str_detect(Architecture,"compound, platform mound")~"platform mound",str_detect(Architecture,"compound, roomblock")~"roomblock",TRUE~Architecture)) %>% 
-  rowid_to_column("id") %>% 
-  ungroup()
+\maketitle
 
-# coords = outTonto$fac %>% 
-#   select(ArchaMapName,UTMEAST,UTMNORTH) %>% 
-#   distinct_all()
-# coords %>% export(here('data/siteCoords.xlsx'))
-coords = import(here('data/siteCoords.xlsx')) %>% 
-  inner_join(nodes) %>% 
-  arrange(id)
-CeramicData = rio::import(here("data/CeramicData.xlsx"),
-                          setclass='tibble')
+\begin{abstract}
+Network analysis has a strong foundation in Southwest archaeology, yet
+the combination of multiple types of artifacts in one
+analysis--multilayer network analysis--has not been formally applied
+except within a single artifact type. Many studies consider material
+culture holistically, yet network analysis has the advantage of focusing
+specifically on the relationships between entities (often communities)
+and how the structure of a network can benefit particular entities. This
+study takes architecture, ceramic, projectile point, and site location
+data from the Roosevelt Platform Mound Study and combines these data
+into a multilayer network analysis. This analysis provides a way to test
+the co-variance of these types of material culture with each other and
+with spatial variation. The results demonstrate a strong association
+between projectile points and spatial distance. Sites containing
+roomblock architecture are highly central in the ceramic network but
+have low centrality in the projectile point network. Overall, the
+ceramic and projectile point networks exhibit significant differences.
+This indicates that the social networks that created these patterns had
+different social mechanisms. One potential cause of these differences is
+gendered spheres of interaction with men exchanging projectile points
+and women exchanging ceramics.
+\end{abstract}
 
-# number of types per site
-count = CeramicData %>% as.data.frame() %>%
-  group_by(ArchaMapName,SWSN_type) %>%
-  summarize(Count = sum(Count)) %>%
-  ungroup()
-table(count$ArchaMapName)
-ceramicDF = CeramicData %>% as.data.frame() %>%
-  group_by(ArchaMapName,SWSN_type) %>%
-  summarize(Count = sum(Count)) %>%
-  ungroup() %>%
-  inner_join(nodes %>% select(ArchaMapName,id)) %>% 
-  select(-ArchaMapName) %>% 
-  pivot_wider(names_from = SWSN_type, values_from = Count) %>%
-  column_to_rownames("id") %>% 
-  select(-`Type, Undifferentiated`)
+\begin{keywords}
+American Southwest; network analysis; multilayer networks; Hohokam;
+projectile points; ceramics; computational archaeology; gender
+\end{keywords}
 
-outTonto = readRDS(here("data/TontoNetworkOutlines.Rds"))
+\hypertarget{introduction}{%
+\section*{Introduction}\label{introduction}}
+\addcontentsline{toc}{section}{Introduction}
 
-outTonto$fac %<>% select(-id,-Area) %>% left_join(nodes)
-# outTonto %<>%
-# mutate(ArchaMapName = droplevels(ArchaMapName))
-# outTonto %>% saveRDS(here("data/TontoNetworkOutlines.Rds"))
-```
+Network science has many applications for archaeologists. It can be
+particularly useful for breaking out of traditional spatial categories
+and examining data through new lenses
+\citep{Feinman2020-nf, Holland-Lulewicz2021-eb}. Multilayer network
+analysis combines multiple networks into one analytical framework
+\citep{Kivela2014-mh}. Social systems are complex webs of
+interrelations. There is too much involved to model the material culture
+of even a technologically simple society. But certainly analyzing more
+than one type of material culture provides a more complete understanding
+of the social dynamics involved in the networks of interaction we wish
+to understand. Yet, there are few examples of multilayer networks in
+archaeology \citep{Giomi2022-dn, Upton2019-yg} and none I am aware of
+that extend beyond ceramics.
 
-```{r sessionInfo,eval=FALSE, include=FALSE}
-# display software information
-sessioninfo::session_info()
-```
+Network studies are particularly prominent in the Southwest and there
+are examples of network studies that consider multiple types of material
+culture. For example, Mills and colleagues \citeyearpar{Mills2013-gc}
+used ceramics and obsidian to understand the transformation of large
+scale social networks. Peeples \citeyearpar{Peeples2018-ib} used
+ceramics and architecture to understand identity and social change in
+the Cibola region. This paper is an application of multilayer network
+analysis that seeks to further network research by examining multiple
+types of material culture in one framework. Given the frequent focus on
+single types of artifacts in network research, a primary question I seek
+to address is whether networks based on different types of material
+culture are positively correlated. This is an important question because
+archaeologists often make inferences based solely on ceramics or other
+types of material culture. Evidence regarding the ways material culture
+co-vary will help archaeologists direct future research. Furthermore, I
+also seek causal explanations for the co-variance, or lack thereof, for
+material culture in the case study I present here.
 
-# Introduction {-}
+Tonto Basin holds great potential for archaeological research due to the
+large cultural resource management projects conducted in the region
+primarily in the 1980s-1990s
+\citep{Ahlstrom1991-eb, Ciolek-Torrello1994-ds, Doelle1992-lu, Rice1998-ku},
+as well as the large number of syntheses and other studies focusing on
+the area
+\citep[e.g.,][]{Clark2004-uw, Dean2000-uc, Elson1995-tb, Elson2000-zq, Hill2015-at, Huntley2016-em, Lange1992-gp, Lyons2003-yy, Lyons2006-to, Lyons2012-pr, Neuzil2008-zd, Oliver2001-tu, Stark1998-mu, Watts2013-ub}.
+This analysis uses data from the Roosevelt Platform Mounds Study (RPMS),
+the largest of these projects \citep{Rice1998-ku}, to examine basic
+architectural data, typed ceramics, and a recent projectile point
+analysis \citep{Bischoff2022-rg}. There are two null hypotheses tested
+in this analysis: (1) that the data is spatially correlated--meaning
+that sites nearest to each other will be most alike--and (2) that the
+architecture, ceramics, and projectile points will all be positively
+correlated--meaning that similar types of architecture, ceramics, and
+projectile points will be found at the same sites. In reality, I
+expected significant variation between these types of material culture.
+As will be demonstrated, the results indicate a positive correlation
+between projectile point similarity and spatial distance. Furthermore,
+ceramic and projectile point networks (hereafter point networks) exhibit
+significant differences. I posit that the differences in these networks
+are connected to the identities of the individuals creating the original
+social networks: particularly that of gender.
 
-Network science has many applications for archaeologists.
-It can be particularly useful for breaking out of traditional spatial categories and examining data through new lenses [@Feinman2020-nf; @Holland-Lulewicz2021-eb].
-Multilayer network analysis combines multiple networks into one analytical framework [@Kivela2014-mh].
-Social systems are complex webs of interrelations.
-There is too much involved to model the material culture of even a technologically simple society.
-But certainly analyzing more than one type of material culture provides a more complete understanding of the social dynamics involved in the networks of interaction we wish to understand.
-Yet, there are few examples of multilayer networks in archaeology [@Giomi2022-dn; @Upton2019-yg] and none I am aware of that extend beyond ceramics.
+\hypertarget{tonto-basin}{%
+\subsection{Tonto Basin}\label{tonto-basin}}
 
-Network studies are particularly prominent in the Southwest and there are examples of network studies that consider multiple types of material culture.
-For example, Mills and colleagues [-@Mills2013-gc] used ceramics and obsidian to understand the transformation of large scale social networks.
-Peeples [-@Peeples2018-ib] used ceramics and architecture to understand identity and social change in the Cibola region.
-This paper is an application of multilayer network analysis that seeks to further network research by examining multiple types of material culture in one framework.
-Given the frequent focus on single types of artifacts in network research, a primary question I seek to address is whether networks based on different types of material culture are positively correlated.
-This is an important question because archaeologists often make inferences based solely on ceramics or other types of material culture.
-Evidence regarding the ways material culture co-vary will help archaeologists direct future research.
-Furthermore, I also seek causal explanations for the co-variance, or lack thereof, for material culture in the case study I present here.
+Tonto Basin is located in east-central Arizona and features Tonto Creek
+flowing from the Mogollon Rim into the Salt River (figure 1). Parts of
+the region have been intensively studied, while certain time periods and
+much of the uplands have been less intensively sampled \citep[see][ for
+a recent overview]{Clark2021-lu}. The basin was also conveniently
+located along regional travel routes, allowing major
+settlements--particularly those with platform mounds--to participate and
+benefit from this exchange
+\citep[p.~129-133]{Caseldine2022-uu, Wood2000-ze}. The junction of the
+Salt River and Tonto Creek now forms Roosevelt Lake since the damming of
+the confluence of the Salt River and Tonto Creek. It was the expansion
+of the dam that precipitated the RPMS and its related projects in the
+late 1980s-1990s. Some important regional differences may have
+differentiated sites along the Salt River and Tonto Creek arms of
+Roosevelt Lake \citep{Lyons2013-ya, Simon2001-am} based on non-local
+ceramics and obsidian. All of the artifacts/data from the RPMS project
+are hosted at the Center for Archaeology and Society at Arizona State
+University, and the data in this study come exclusively from this
+project.
 
-Tonto Basin holds great potential for archaeological research due to the large cultural resource management projects conducted in the region primarily in the 1980s-1990s [@Ahlstrom1991-eb; @Ciolek-Torrello1994-ds; @Doelle1992-lu; @Rice1998-ku], as well as the large number of syntheses and other studies focusing on the area [e.g., @Clark2004-uw; @Dean2000-uc; @Elson1995-tb; @Elson2000-zq;  @Hill2015-at; @Huntley2016-em; @Lange1992-gp; @Lyons2003-yy; @Lyons2006-to; @Lyons2012-pr;  @Neuzil2008-zd; @Oliver2001-tu; @Stark1998-mu; @Watts2013-ub].
-This analysis uses data from the Roosevelt Platform Mounds Study (RPMS), the largest of these projects [@Rice1998-ku], to examine basic architectural data, typed ceramics, and a recent projectile point analysis [@Bischoff2022-rg].
-There are two null hypotheses tested in this analysis: (1) that the data is spatially correlated--meaning that sites nearest to each other will be most alike--and (2) that the architecture, ceramics, and projectile points will all be positively correlated--meaning that similar types of architecture, ceramics, and projectile points will be found at the same sites.
-In reality, I expected significant variation between these types of material culture.
-As will be demonstrated, the results indicate a positive correlation between projectile point similarity and spatial distance.
-Furthermore, ceramic and projectile point networks (hereafter point networks) exhibit significant differences.
-I posit that the differences in these networks are connected to the identities of the individuals creating the original social networks: particularly that of gender.
+\begin{figure}
+\includegraphics[width=1\linewidth]{figures/TontoBasinSitesv3} \caption{Map of archaeology sites from the Roosevelt Platform Mounds Study in Tonto Basin.
+Groups use the original assignments.}\label{fig:TontoBasinSites}
+\end{figure}
 
-## Tonto Basin
+Tonto Basin was occupied, although likely not continuously, from the
+Archaic through the late Classic period. The height of occupation,
+particularly the sites excavated during the project, dates between AD
+1275 and 1325, which corresponds to the Roosevelt Phase. The Roosevelt
+phase is notable for the appearance of Salado pottery and platform
+mounds--probably introduced from the Phoenix Basin \citep{Elson1996-ym}.
+Moderate occupation continued into the Gila phase (AD 1325-1450), which
+was characterized by population loss and aggregation into larger sites.
+This phase marks the end of recognizable occupation in the area until
+the historic period.
 
-Tonto Basin is located in east-central Arizona and features Tonto Creek flowing from the Mogollon Rim into the Salt River (figure `r figs`).
-Parts of the region have been intensively studied, while certain time periods and much of the uplands have been less intensively sampled [see @Clark2021-lu for a recent overview].
-The basin was also conveniently located along regional travel routes, allowing major settlements--particularly those with platform mounds--to participate and benefit from this exchange [@Caseldine2022-uu; @Wood2000-ze, p.
-129-133].
-The junction of the Salt River and Tonto Creek now forms Roosevelt Lake since the damming of the confluence of the Salt River and Tonto Creek.
-It was the expansion of the dam that precipitated the RPMS and its related projects in the late 1980s-1990s.
-Some important regional differences may have differentiated sites along the Salt River and Tonto Creek arms of Roosevelt Lake [@Lyons2013-ya; @Simon2001-am] based on non-local ceramics and obsidian.
-All of the artifacts/data from the RPMS project are hosted at the Center for Archaeology and Society at Arizona State University, and the data in this study come exclusively from this project.
+One aspect of Tonto Basin that has caught the attention of researchers
+is its location in a border zone between the Hohokam, Mogollon, and
+Ancestral Pueblo regions
+\citep[e.g.,][]{Caseldine2022-uu, Clark2001-ck, Elson1994-hk, Hill2015-at, Huntley2016-em, Lyons2012-pr, Lyons2013-ya, Neuzil2008-zd, Wood2000-ze}.
+Of particular note is the presence of masonry roomblock architecture and
+pottery uncharacteristic of the local Hohokam traditions, which has been
+interpreted as Kayenta immigration into Tonto Basin
+\citep{Clark2001-ck, Lyons2003-yy, Lyons2006-to, Stark1998-mu}. This
+migration into the Tonto Basinwas part of a larger migration from north
+to south and is associated with the origin of the Salado phenomenon.
+Salado pottery production was widespread across southern Arizona, but
+often the largest sources of production was centered at the location of
+a former Kayenta enclave
+\citep{Hill2015-at, Huntley2016-em, Lyons2012-pr, Neuzil2008-zd}. The
+connection between immigration and an important type of decorated
+pottery in the region suggests that sites with roomblocks may have
+higher centrality in ceramic networks. Having high centrality means they
+have more connections to other nodes (i.e., sites). The presence of
+migrant communities in Tonto Basin invites questions regarding the
+relationship between sites--particularly between migrant and local
+communities. A network approach is an ideal way to examine the
+relationships between sites. Evidence that projectile points move
+between sites \citep[e.g.,][]{Watts2013-ub} indicates that this is also
+a productive line of analysis, and one that is less-commonly pursued
+compared to architecture and ceramics.
 
-```{r TontoBasinSites, fig.cap= 'Map of archaeology sites from the Roosevelt Platform Mounds Study in Tonto Basin. Groups use the original assignments.'}
+\hypertarget{network-analysis}{%
+\section*{Network Analysis}\label{network-analysis}}
+\addcontentsline{toc}{section}{Network Analysis}
 
-figs = figs + 1
-knitr::include_graphics(here("figures/TontoBasinSitesv3.png"))
-```
+Network analysis is a flexible tool for analyzing many types of data
+\citep[see][]{Brughmans2023-uj}. The key element is that some way must
+be determined to tie each entity together. This analysis uses sites as
+nodes in the network and some type of similarity as a tie to connect
+each node. Each type of similarity depends on the type of data. Table 1
+lists the 11 sites with sufficient data to include in this study with
+the types of architecture present, periods occupied, and total number of
+ceramics and complete projectile points found at each site. The lack of
+complete projectile points at some sites is the largest caveat in this
+study, but complete points were necessary for the type of projectile
+point analysis used. The periods, architecture types, and ceramic types
+were assigned during the RPMS project and can be found in the project
+reports \citep{Rice1998-ku}. This is not a random sample of Tonto Basin
+sites. It is merely a sample of convenience. Sampling can have a major
+impact on conclusions. In this case, I am not drawing conclusions based
+on a complete network or the existence of a random sample (though biases
+can still affect the results). Rather, I am interested in whether the
+types of material culture present in this study co-vary with each other
+and with space. These questions do not require a complete network.
 
-Tonto Basin was occupied, although likely not continuously, from the Archaic through the late Classic period.
-The height of occupation, particularly the sites excavated during the project, dates between AD 1275 and 1325, which corresponds to the Roosevelt Phase.
-The Roosevelt phase is notable for the appearance of Salado pottery and platform mounds--probably introduced from the Phoenix Basin [@Elson1996-ym].
-Moderate occupation continued into the Gila phase (AD 1325-1450), which was characterized by population loss and aggregation into larger sites.
-This phase marks the end of recognizable occupation in the area until the historic period.
+\hypertarget{defining-networks-and-relations}{%
+\subsection{Defining Networks and
+Relations}\label{defining-networks-and-relations}}
 
-One aspect of Tonto Basin that has caught the attention of researchers is its location in a border zone between the Hohokam, Mogollon, and Ancestral Pueblo regions [e.g., @Caseldine2022-uu; @Clark2001-ck; @Elson1994-hk; @Hill2015-at; @Huntley2016-em; @Lyons2012-pr; @Lyons2013-ya; @Neuzil2008-zd; @Wood2000-ze].
-Of particular note is the presence of masonry roomblock architecture and pottery uncharacteristic of the local Hohokam traditions, which has been interpreted as Kayenta immigration into Tonto Basin [@Clark2001-ck; @Lyons2003-yy; @Lyons2006-to; @Stark1998-mu].
-This migration into the Tonto Basinwas part of a larger migration from north to south and is associated with the origin of the Salado phenomenon.
-Salado pottery production was widespread across southern Arizona, but often the largest sources of production was centered at the location of a former Kayenta enclave [@Hill2015-at; @Huntley2016-em; @Lyons2012-pr; @Neuzil2008-zd].
-The connection between immigration and an important type of decorated pottery in the region suggests that sites with roomblocks may have higher centrality in ceramic networks.
-Having high centrality means they have more connections to other nodes (i.e., sites).
-The presence of migrant communities in Tonto Basin invites questions regarding the relationship between sites--particularly between migrant and local communities.
-A network approach is an ideal way to examine the relationships between sites.
-Evidence that projectile points move between sites [e.g., @Watts2013-ub] indicates that this is also a productive line of analysis, and one that is less-commonly pursued compared to architecture and ceramics.
+Network approaches are best suited where the relationships between
+entities are of primary interest. In this case, the entities are defined
+as archaeology sites. The archaeology sites in turn represent a group of
+individuals who lived or visited the area of the site and left material
+remains behind. Ideally, studying social relations would be approached
+at the individual level, but rarely can archaeologists address
+archaeological data with that level of specificity. Watts
+\citeyearpar{Watts2013-ub} used projectile points from Tonto Basin in
+just such a study. His analysis of flake scar patterning identified
+individual knappers, or at least clusters of knappers with similar
+knapping styles, and the distribution of their projectile points around
+Tonto Basin. He found strong connections between many parts of the
+eastern Tonto Basin. Unfortunately, this study will only attempt a
+site-level analysis. Watt's study does, however, illustrate how
+relations can be defined between nodes in a network analysis. He used
+similarity networks where the points that had similar flaking styles
+were connected together. Similarity networks are a commonly used type of
+network in archaeology
+\citep[e.g.,][]{Birch2018-xx, Borck2015-jy, Cochrane2010-oa, Golitko2015-we, Lulewicz2019-lu, Mills2013-wq, Peeples2018-ib, Terrell2010-hd}.
+These studies use a variety of artifacts and methods to construct their
+networks, but each has demonstrated the utility of network methods
+within archaeological contexts. This study also uses similarity networks
+to group sites by similarity in ceramic assemblages and similarity in
+projectile point forms.
 
-# Network Analysis {-}
+Yet, it can be difficult to understand how a similarity network is
+related to a past social network. What kinds of interactions are
+represented by similarities in projectile points or in ceramics?
+Answering this question is also a crucial step in interpreting networks.
+One way to examine relations between individuals who make or use similar
+types of projectile points or pots is to talk about identity. Identity
+can be a troubling topic for anthropologists. There are numerous
+meanings given to it and numerous scales at which it applies
+\citep{Brubaker2000-na}, but Peeples \citeyearpar{Peeples2018-ib} has
+introduced archaeologists to a more practical approach that is useful
+here. This approach views identity as existing along two axes: one
+categorical and the other relational. Relational identification is the
+process of identifying with someone due to frequent interaction.
+Categorical identification is the process of identifying with someone
+because you belong to a recognized social group. For example, members of
+the same moiety would share a categorical identity. They would likely
+also share a relational identity if they frequently interacted. These
+identities can be reflected in material culture. Pottery makes a good
+example. Peeples \citeyearpar[p.~151]{Peeples2018-ib} notes that bowls
+in the Cibola region were sometimes painted with a bright, red slip with
+designs painted on the interior and they were among the first in the
+Southwest to have designs on the exterior. This is a strong indication
+of categorical identity. The potters were attempting to make a clear
+signal for whoever saw the pot. Relational identity can be seen in the
+way the potter prepares the clay recipe and smooths the coils. The
+particulars of these actions would be learned from close interaction
+with a teacher or other potters. In this way, relational identity can be
+compared to communities of practice \citep{Lave1991-ik, Wenger1998-rx}.
 
-Network analysis is a flexible tool for analyzing many types of data [see @Brughmans2023-uj].
-The key element is that some way must be determined to tie each entity together.
-This analysis uses sites as nodes in the network and some type of similarity as a tie to connect each node.
-Each type of similarity depends on the type of data.
-Table `r tbls` lists the `r nrow(sites)` sites with sufficient data to include in this study with the types of architecture present, periods occupied, and total number of ceramics and complete projectile points found at each site.
-The lack of complete projectile points at some sites is the largest caveat in this study, but complete points were necessary for the type of projectile point analysis used.
-The periods, architecture types, and ceramic types were assigned during the RPMS project and can be found in the project reports [@Rice1998-ku].
-This is not a random sample of Tonto Basin sites.
-It is merely a sample of convenience.
-Sampling can have a major impact on conclusions.
-In this case, I am not drawing conclusions based on a complete network or the existence of a random sample (though biases can still affect the results).
-Rather, I am interested in whether the types of material culture present in this study co-vary with each other and with space.
-These questions do not require a complete network.
+In this analysis, I argue that ceramics and architecture are more likely
+to represent categorical identity. The architecture discussed is highly
+visible and representative of historical group membership. The ceramics
+are grouped into types based primarily on decoration, which is strongly
+indicative of group membership. What this means is that links between
+nodes in the cer The details of architecture and ceramics used in this
+study are discussed in a future section, but it is important to note
+that these designations as categorical or relational are contextually
+dependent on this study. Clark \citeyearpar{Clark2001-ck} has an
+excellent discussion on material culture and its relationship to
+identity. He identifies certain patterns, but his synthesis and other
+research
+\citep{Carr1995-mq, Carr1995-sh, Dietler1998-zo, Gosselain1998-qd, Gosselain2000-wf, Gosselain2016-as, Hodder1982-no, Huntley2008-qu, Lemonnier1986-le, Lyons2003-yy, Neuzil2008-zd, Sassman2001-sw, Stark1998-mu, Wiessner1983-ei, Wiessner1997-su}
+strongly indicates that the relationship between material culture and
+identity is culturally relative. Projectile points can in some cases
+indicate group membership \citep{Wiessner1983-ei, Wiessner1997-su}, but
+in this case the triangular and side-notched points were likely used for
+different purposes \citep[see][]{Loendorf2015-ww, Sliva2002-oz}. Styles
+were difficult to determine for the original researchers
+\citep{Rice1994-rk}, and I had the same difficulty. In my opinion, the
+subtle differences between projectile point outlines are more
+representative of interactions between knappers than markers of group
+identity for this particular case. This is indicative of relational
+identity. There are no environmental reasons to assume differences in
+projectile points, as each site was in a similar ecozone. Each hunter
+would have been using the points for the same game and differences in
+point styles would have been primarily cultural \citep{Sliva2002-oz}.
 
-## Defining Networks and Relations
+There is one other aspect to these relations that I believe played an
+important role. A central way people identify themselves is by gender.
+Gender roles, like identity, can also vary in complex ways, but for the
+purposes of this analysis I will use a simplified model. Women made
+pottery and men made projectile points. This was not always true of
+course, but this fits the available data for the Hohokam
+\citep[p.~253]{Crown1996-xb, VanPool2010-im}. An examination of the
+burial record for the RPMS study shows no obvious indication that women
+made pottery, only that both men and women were buried with pottery
+\citep[Table 10.7]{Loendorf1998-ln}. On the other hand, projectile
+points were almost exclusively buried with males. This does not mean
+that a man could not move pots from one site to another or that a woman
+could not do the same with a projectile point, but, in general,
+differences in point and ceramic networks are most likely to indicate
+differences in gender networks.
 
-Network approaches are best suited where the relationships between entities are of primary interest.
-In this case, the entities are defined as archaeology sites.
-The archaeology sites in turn represent a group of individuals who lived or visited the area of the site and left material remains behind.
-Ideally, studying social relations would be approached at the individual level, but rarely can archaeologists address archaeological data with that level of specificity.
-Watts [-@Watts2013-ub] used projectile points from Tonto Basin in just such a study.
-His analysis of flake scar patterning identified individual knappers, or at least clusters of knappers with similar knapping styles, and the distribution of their projectile points around Tonto Basin.
-He found strong connections between many parts of the eastern Tonto Basin.
-Unfortunately, this study will only attempt a site-level analysis.
-Watt's study does, however, illustrate how relations can be defined between nodes in a network analysis.
-He used similarity networks where the points that had similar flaking styles were connected together.
-Similarity networks are a commonly used type of network in archaeology [e.g., @Birch2018-xx; @Borck2015-jy; @Cochrane2010-oa; @Golitko2015-we; @Lulewicz2019-lu; @Mills2013-wq; @Peeples2018-ib; @Terrell2010-hd].
-These studies use a variety of artifacts and methods to construct their networks, but each has demonstrated the utility of network methods within archaeological contexts.
-This study also uses similarity networks to group sites by similarity in ceramic assemblages and similarity in projectile point forms.
+From this discussion, I expect the ceramic and point networks to differ
+in two ways. I expect the ceramic network to represent categorical
+identity among women, and I expect the point network to represent
+relational identity among men. This makes the networks somewhat more
+difficult to compare. Ideally, we would have networks of the same
+identity type, but it does provide the expectation that we should see
+significant differences in the networks.
 
-Yet, it can be difficult to understand how a similarity network is related to a past social network.
-What kinds of interactions are represented by similarities in projectile points or in ceramics?
-Answering this question is also a crucial step in interpreting networks.
-One way to examine relations between individuals who make or use similar types of projectile points or pots is to talk about identity.
-Identity can be a troubling topic for anthropologists.
-There are numerous meanings given to it and numerous scales at which it applies [@Brubaker2000-na], but Peeples [-@Peeples2018-ib] has introduced archaeologists to a more practical approach that is useful here.
-This approach views identity as existing along two axes: one categorical and the other relational.
-Relational identification is the process of identifying with someone due to frequent interaction.
-Categorical identification is the process of identifying with someone because you belong to a recognized social group.
-For example, members of the same moiety would share a categorical identity.
-They would likely also share a relational identity if they frequently interacted.
-These identities can be reflected in material culture.
-Pottery makes a good example.
-Peeples [-@Peeples2018-ib, p.
-151] notes that bowls in the Cibola region were sometimes painted with a bright, red slip with designs painted on the interior and they were among the first in the Southwest to have designs on the exterior.
-This is a strong indication of categorical identity.
-The potters were attempting to make a clear signal for whoever saw the pot.
-Relational identity can be seen in the way the potter prepares the clay recipe and smooths the coils.
-The particulars of these actions would be learned from close interaction with a teacher or other potters.
-In this way, relational identity can be compared to communities of practice [@Lave1991-ik; @Wenger1998-rx].
+The network for architecture, ceramics, and projectile points, were
+created using the Jaccard similarity coefficient \citep{Jaccard1912-na}
+to determine the links between each site. This is a simple measure of
+similarity that does not take into account abundance. It was chosen due
+to the large differences in sample sizes between sites for the ceramics
+and projectile points and because count data were irrelevant or not
+readily available for the architectural data.
 
-In this analysis, I argue that ceramics and architecture are more likely to represent categorical identity.
-The architecture discussed is highly visible and representative of historical group membership.
-The ceramics are grouped into types based primarily on decoration, which is strongly indicative of group membership.
-What this means is that links between nodes in the cer The details of architecture and ceramics used in this study are discussed in a future section, but it is important to note that these designations as categorical or relational are contextually dependent on this study.
-Clark [-@Clark2001-ck] has an excellent discussion on material culture and its relationship to identity.
-He identifies certain patterns, but his synthesis and other research [@Carr1995-mq; @Carr1995-sh; @Dietler1998-zo; @Gosselain1998-qd; @Gosselain2000-wf; @Gosselain2016-as; @Hodder1982-no; @Huntley2008-qu; @Lemonnier1986-le; @Lyons2003-yy; @Neuzil2008-zd; @Sassman2001-sw; @Stark1998-mu; @Wiessner1983-ei; @Wiessner1997-su] strongly indicates that the relationship between material culture and identity is culturally relative.
-Projectile points can in some cases indicate group membership [@Wiessner1983-ei; @Wiessner1997-su], but in this case the triangular and side-notched points were likely used for different purposes [see @Loendorf2015-ww; @Sliva2002-oz].
-Styles were difficult to determine for the original researchers [@Rice1994-rk], and I had the same difficulty.
-In my opinion, the subtle differences between projectile point outlines are more representative of interactions between knappers than markers of group identity for this particular case.
-This is indicative of relational identity.
-There are no environmental reasons to assume differences in projectile points, as each site was in a similar ecozone.
-Each hunter would have been using the points for the same game and differences in point styles would have been primarily cultural [@Sliva2002-oz].
+\begin{table}
 
-There is one other aspect to these relations that I believe played an important role.
-A central way people identify themselves is by gender.
-Gender roles, like identity, can also vary in complex ways, but for the purposes of this analysis I will use a simplified model.
-Women made pottery and men made projectile points.
-This was not always true of course, but this fits the available data for the Hohokam [@Crown1996-xb; @VanPool2010-im, p.
-253].
-An examination of the burial record for the RPMS study shows no obvious indication that women made pottery, only that both men and women were buried with pottery [@Loendorf1998-ln, Table 10.7].
-On the other hand, projectile points were almost exclusively buried with males.
-This does not mean that a man could not move pots from one site to another or that a woman could not do the same with a projectile point, but, in general, differences in point and ceramic networks are most likely to indicate differences in gender networks.
+\caption{\label{tab:sites}Project Sites and Data}
+\centering
+\resizebox{\linewidth}{!}{
+\begin{tabular}[t]{llllrr}
+\toprule
+Site & Area & Architecture & Phase & Ceramics & Points\\
+\midrule
+AZ U:3:128 (ASM) & Cline Mesa & compound & Roosevelt, Gila & 7,522 & 5\\
+AZ U:4:032 (ASM) & Cline Mesa & compound & Sacaton, Roosevelt & 5,352 & 3\\
+AZ V:5:119 (ASM) & Livingston & compound & Roosevelt & 1,679 & 3\\
+Bass Point Mound & Rock Island & platform mound & Roosevelt & 27,602 & 13\\
+Cline Terrace Mound, Monster Ruin & Cline Mesa & platform mound & Roosevelt, Gila & 169,567 & 45\\
+\addlinespace
+Indian Point Complex & Cline Mesa & compound, roomblock & Roosevelt, Gila & 39,493 & 14\\
+Pinto Point Mound & Livingston & compound, platform mound & Roosevelt & 19,174 & 21\\
+Saguaro Muerto & Livingston & roomblock & Roosevelt & 13,458 & 5\\
+Sand Dune Site & Livingston & compound & Roosevelt & 10,545 & 12\\
+Schoolhouse Point Mesa Complex & Schoolhouse Mesa & compound & Roosevelt & 74,592 & 12\\
+\addlinespace
+Schoolhouse Point Mound & Schoolhouse Mesa & platform mound & Roosevelt, Gila & 240,635 & 36\\
+\bottomrule
+\end{tabular}}
+\end{table}
 
-From this discussion, I expect the ceramic and point networks to differ in two ways.
-I expect the ceramic network to represent categorical identity among women, and I expect the point network to represent relational identity among men.
-This makes the networks somewhat more difficult to compare.
-Ideally, we would have networks of the same identity type, but it does provide the expectation that we should see significant differences in the networks.
+The types of similarity used in this study are varied depending on the
+type of network. Part of this analysis is a visual approach and some of
+the methods require non-weighted links, thus I will be only using the
+strongest links. There is rarely a clear dividing line between similar
+and not similar. This can be a challenge for network analysis, because
+we can end up with networks where every node is connected to every other
+node. The decision to binarize a network--remove the weakest links and
+then consider each link of equal value--has its drawbacks
+\citep{Peeples2013-tj}, but is necessary in this case. A solution is to
+assign weights to each link that defines the strength of the tie. These
+networks are often difficult to visualize, and some network algorithms
+do not allow for weights. A common approach is to keep only the
+strongest ties by either ranking the ties or using a cutoff value.
 
-The network for architecture, ceramics, and projectile points, were created using the Jaccard similarity coefficient [@Jaccard1912-na] to determine the links between each site.
-This is a simple measure of similarity that does not take into account abundance.
-It was chosen due to the large differences in sample sizes between sites for the ceramics and projectile points and because count data were irrelevant or not readily available for the architectural data.
-
-```{r sites, echo=FALSE}
-tbls = tbls + 1
-cap = "Project Sites and Data"
-sites %>% 
-  select(Site = ArchaMapName,Area,Architecture = architecture,Phase,Ceramics,Points) %>%
-  # as_hux() %>% set_width(1) %>% set_number_format(everywhere,c(5,6),fmt_pretty()) %>% 
-  # theme_article() %>% 
-  # set_caption(cap)
-  knitr::kable(caption= cap,
-               booktabs = T,
-               format.args = list(big.mark = ",",
-                                  scientific = FALSE)) %>%
-  kable_styling(latex_options = "scale_down")
-```
-
-The types of similarity used in this study are varied depending on the type of network.
-Part of this analysis is a visual approach and some of the methods require non-weighted links, thus I will be only using the strongest links.
-There is rarely a clear dividing line between similar and not similar.
-This can be a challenge for network analysis, because we can end up with networks where every node is connected to every other node.
-The decision to binarize a network--remove the weakest links and then consider each link of equal value--has its drawbacks [@Peeples2013-tj], but is necessary in this case.
-A solution is to assign weights to each link that defines the strength of the tie.
-These networks are often difficult to visualize, and some network algorithms do not allow for weights.
-A common approach is to keep only the strongest ties by either ranking the ties or using a cutoff value.
-
-I used ranked links to keep consistent values between networks.
-This involved calculating the strength of similarity between each site in the network and then ranking the strength of similarity.
-The top **n** connections between each site were kept and the rest were discarded.
-In practice, not every node will have the same number of connections.
-Some may have fewer ties if there are not enough nodes that are similar.
-More ties can exist when there are ties in the ranks.
-Because the network is not directed (meaning the ties indicate similarity in both directions) one node can have several ties pointing to it because that node was in the top **n** of multiple other nodes.
-In the latter case, this is a good indication of a central node in the network.
-Because I am using several types of networks, there is no common cutoff value I can use for the number of ties to keep.
-Instead of arbitrarily picking one value, I have calculated each metric using networks composed of the top 3-10 ties.
-Meaning that each metric is calculated for a network composed of ties with the top 3 connections, and the process is repeated for new networks composed of ties with the top 4 connections, and so forth.
+I used ranked links to keep consistent values between networks. This
+involved calculating the strength of similarity between each site in the
+network and then ranking the strength of similarity. The top \textbf{n}
+connections between each site were kept and the rest were discarded. In
+practice, not every node will have the same number of connections. Some
+may have fewer ties if there are not enough nodes that are similar. More
+ties can exist when there are ties in the ranks. Because the network is
+not directed (meaning the ties indicate similarity in both directions)
+one node can have several ties pointing to it because that node was in
+the top \textbf{n} of multiple other nodes. In the latter case, this is
+a good indication of a central node in the network. Because I am using
+several types of networks, there is no common cutoff value I can use for
+the number of ties to keep. Instead of arbitrarily picking one value, I
+have calculated each metric using networks composed of the top 3-10
+ties. Meaning that each metric is calculated for a network composed of
+ties with the top 3 connections, and the process is repeated for new
+networks composed of ties with the top 4 connections, and so forth.
 Examining the range of these metrics provides a more robust analysis.
 
-Two network metrics are discussed in this analysis: eigenvector centrality and multilayer network correlation--the second will be discussed in the next section.
-Node centrality is a common way to quantify networks [@Borgatti2005-iz].
-Centrality is a measure of the influence of the node on the network.
-Nodes with higher centrality generally derive or generate greater benefits from the network.
-For example, if I have many friends, then I can call in more favors.
-Eigenvector centrality is one way to measure node centrality and is commonly used in archaeology.
-This metric describes how well a node is connected to the network as a whole and is helpful in comparing different networks containing the same nodes.
-For example, if my friends have many friends, then I will have a greater advantage then someone with an equal number of friends, but whose friends have few friends.
-Essentially, this is a way of measuring second-order and beyond connections.
-Eigenvector centrality is also more robust to missing nodes than other measures [@Peeples2017-kl], a major problem in most archaeological studies.
+Two network metrics are discussed in this analysis: eigenvector
+centrality and multilayer network correlation--the second will be
+discussed in the next section. Node centrality is a common way to
+quantify networks \citep{Borgatti2005-iz}. Centrality is a measure of
+the influence of the node on the network. Nodes with higher centrality
+generally derive or generate greater benefits from the network. For
+example, if I have many friends, then I can call in more favors.
+Eigenvector centrality is one way to measure node centrality and is
+commonly used in archaeology. This metric describes how well a node is
+connected to the network as a whole and is helpful in comparing
+different networks containing the same nodes. For example, if my friends
+have many friends, then I will have a greater advantage then someone
+with an equal number of friends, but whose friends have few friends.
+Essentially, this is a way of measuring second-order and beyond
+connections. Eigenvector centrality is also more robust to missing nodes
+than other measures \citep{Peeples2017-kl}, a major problem in most
+archaeological studies.
 
-A single network is typically used in network analyses, but multilayer networks can be more informative.
-Multilayer networks are layered networks where nodes have different types of connections [@Kivela2014-mh].
-In this analysis, each network has the same nodes--each archaeological site--but different types of relationships between them.
-The combination of these individual networks is a multilayer network (also called a multiplex network).
-Multilayer networks allow for methods to be applied on multiple networks at once [see @Brodka2018-vz].
-The method applicable to this analysis is layer correlation.
-Either Pearson or Spearman rank correlation can be computed to determine the strength and direction of correlation between each layer.
-Bródka and colleagues [-@Brodka2018-vz] have provided an R package to compute these statistics, which I have used in this study.
-They recommend the Pearson correlation in most circumstances.
-I will display the results of the Pearson correlation, but the Spearman correlation provided similar results.
-Eigenvector centrality was mentioned in the previous section, but it is also calculated as a multilayer eigenvector centrality.
-In this case the centrality measure is simply the mean of the eigenvector centraliy for each separate layer, as the layers are not interdependent [@Frost2022-lr].
-Multilayer eigenvector centrality results are only presented for the combination of ceramic and point layers.
-These networks can be thought of as dependent variables for the spatial and architectural networks--meaning spatial distance and architecture are considered to be important variables determining the structure of the ceramic and point networks.
+A single network is typically used in network analyses, but multilayer
+networks can be more informative. Multilayer networks are layered
+networks where nodes have different types of connections
+\citep{Kivela2014-mh}. In this analysis, each network has the same
+nodes--each archaeological site--but different types of relationships
+between them. The combination of these individual networks is a
+multilayer network (also called a multiplex network). Multilayer
+networks allow for methods to be applied on multiple networks at once
+\citep[see][]{Brodka2018-vz}. The method applicable to this analysis is
+layer correlation. Either Pearson or Spearman rank correlation can be
+computed to determine the strength and direction of correlation between
+each layer. Bródka and colleagues \citeyearpar{Brodka2018-vz} have
+provided an R package to compute these statistics, which I have used in
+this study. They recommend the Pearson correlation in most
+circumstances. I will display the results of the Pearson correlation,
+but the Spearman correlation provided similar results. Eigenvector
+centrality was mentioned in the previous section, but it is also
+calculated as a multilayer eigenvector centrality. In this case the
+centrality measure is simply the mean of the eigenvector centraliy for
+each separate layer, as the layers are not interdependent
+\citep{Frost2022-lr}. Multilayer eigenvector centrality results are only
+presented for the combination of ceramic and point layers. These
+networks can be thought of as dependent variables for the spatial and
+architectural networks--meaning spatial distance and architecture are
+considered to be important variables determining the structure of the
+ceramic and point networks.
 
-The robustness of results were tested by randomly sampling the underlying network matrices 10,000 times and comparing the results to the random samples to obtain a p-value.
-This provides a baseline that determines how likely it is to obtain a given result by chance.
+The robustness of results were tested by randomly sampling the
+underlying network matrices 10,000 times and comparing the results to
+the random samples to obtain a p-value. This provides a baseline that
+determines how likely it is to obtain a given result by chance.
 
-Certain caveats must be acknowledged.
-All archaeologists deal with missing data, which can greatly affect the results of analysis.
-Network analysts have grappled with this question and often use resampling or similar methods to deal with this problem [@Bischoff2021-sf; @Bolland1988-zm; @Borgatti2006-jl; @Brughmans2023-uj; @Costenbader2003-st; @Galaskiewicz1991-qb; @Gjesfjeld2015-hw; @Lee2009-wr; @Rivera-Hutinel2012-ik].
-Regardless of statistical methods, unaccounted for biases will still produce invalid results [@Bischoff2021-sf].
-The most important bias that may affect these findings is chronology.
-For example, Bass Point Mound and Cline Terrace Mound were probably not occupied at the same time [@Jacobs1997-ob; @Lindauer1995-wq].
-Does that invalidate the network analysis?
-I do not think so.
-While the networks involve nodes as sites, what the networks are really meant to represent are groups of people.
-Furthermore, they are not individuals, but the aggregate decisions of people over more than one generation.
-Current models of Hohokam social groups suggest a certain residential stability [@Craig2017-rg, p.
-336].
-Perhaps the residents of these sites were not physically located at the site at the same time as the other inhabitants were located at their site, but they may have been at another nearby location.
-The spacing of settlements may indicate a corporate form of social organization [@Clark2004-uw; @Rice1998-rz, p.96].
-If this is accurate, then the communities inhabiting the sites may likely have existed in some form prior to the establishment of any particular site and have continued on.
-It is these communities that the network analysis is attempting to capture.
-Still, evidence indicates that most sites were contemporaneous.
-For example, Cline Terrace Mound and Schoolhouse Point were most probably occupied at the same time [@Lyons2013-ya].
-It is best, though, to think of these interpretations as estimations based on incomplete data.
+Certain caveats must be acknowledged. All archaeologists deal with
+missing data, which can greatly affect the results of analysis. Network
+analysts have grappled with this question and often use resampling or
+similar methods to deal with this problem
+\citep{Bischoff2021-sf, Bolland1988-zm, Borgatti2006-jl, Brughmans2023-uj, Costenbader2003-st, Galaskiewicz1991-qb, Gjesfjeld2015-hw, Lee2009-wr, Rivera-Hutinel2012-ik}.
+Regardless of statistical methods, unaccounted for biases will still
+produce invalid results \citep{Bischoff2021-sf}. The most important bias
+that may affect these findings is chronology. For example, Bass Point
+Mound and Cline Terrace Mound were probably not occupied at the same
+time \citep{Jacobs1997-ob, Lindauer1995-wq}. Does that invalidate the
+network analysis? I do not think so. While the networks involve nodes as
+sites, what the networks are really meant to represent are groups of
+people. Furthermore, they are not individuals, but the aggregate
+decisions of people over more than one generation. Current models of
+Hohokam social groups suggest a certain residential stability
+\citep[p.~336]{Craig2017-rg}. Perhaps the residents of these sites were
+not physically located at the site at the same time as the other
+inhabitants were located at their site, but they may have been at
+another nearby location. The spacing of settlements may indicate a
+corporate form of social organization
+\citep[p.96]{Clark2004-uw, Rice1998-rz}. If this is accurate, then the
+communities inhabiting the sites may likely have existed in some form
+prior to the establishment of any particular site and have continued on.
+It is these communities that the network analysis is attempting to
+capture. Still, evidence indicates that most sites were contemporaneous.
+For example, Cline Terrace Mound and Schoolhouse Point were most
+probably occupied at the same time \citep{Lyons2013-ya}. It is best,
+though, to think of these interpretations as estimations based on
+incomplete data.
 
-```{r}
-# network functions 
-makeGraph = function(d, nLinks = NULL){
-  # igraph
-  g = graph_from_adjacency_matrix(d,mode = 'undirected',diag = F,weighted = T)
-  V(g)$name = nodes$id %>% as.character()
-  V(g)$ArchaMapName = nodes$ArchaMapName
-  V(g)$Area = nodes$Area
-  V(g)$Architecture = nodes$Architecture
-  V(g)$centrality = eigen_centrality(g, weights = E(g)$weight)$vector
+\hypertarget{spatial}{%
+\subsection*{Spatial}\label{spatial}}
+\addcontentsline{toc}{subsection}{Spatial}
 
-  if(!is.null(nLinks)){
-    # calculate top n (with ties) edges and convert to igraph object
-    # convert to dataframe
-    n = igraph::as_data_frame(g)
-    n = bind_rows(n,n %>% rename(from = to, to = from))
-    n %<>% # sort data frame
-      arrange(from,desc(weight)) %>%
-      # rank by site
-      group_by(from) %>%
-      mutate(rank = rank(-weight)) %>% 
-      ungroup()
-    nSub = n %>% 
-      group_by(from) %>% 
-      slice_max(n = nLinks, order_by = desc(rank)) %>% 
-      ungroup()
+The simplest network is the spatial network. This network was created by
+calculating the Euclidean distance from every site to every other site.
+Euclidean distance is a shortcut that does not represent actual travel
+routes. Least cost paths provide more accurate data \citep[see][ for an
+application in Tonto Basin]{Caseldine2022-uu}, yet an analysis in
+similar terrain indicates that for distances longer than those in this
+study the benefits of least cost path distances in place of Euclidean
+distance are negligible \citep{Bischoff2017-gx}. Least cost paths are
+also hampered by the presence of Roosevelt Lake and other modern impacts
+on the landscape. The spatial network created here is equivalent to
+proximal point analysis, which was used in one of the earliest network
+studies in archaeology \citep{Terrell1977-co} and continues to be used
+to study potential pathways of interaction
+\citep[see][]{Broodbank2000-pg, Collar2013-fb}. The inclusion of a
+spatial network serves as a null hypothesis to test the other networks
+against.
 
-    gRanks = graph_from_data_frame(nSub,directed = F, vertices = nodes)
-    gRanks %<>% simplify()
-    V(gRanks)$centrality = eigen_centrality(gRanks, weights = E(gRanks)$weight)$vector
-    return(gRanks)
-  } else {
-    return(g)
-  }
-}
+\hypertarget{architecture}{%
+\subsection*{Architecture}\label{architecture}}
+\addcontentsline{toc}{subsection}{Architecture}
 
-values2graphicsMod = function (values, output = "color", palette = "Paired") 
-{
-  if (!output %in% c("color", "shape")) {
-    stop("wrong parameter: output")
-  }
-  if (is.data.frame(values)) {
-    values <- values[[1]]
-  }
-  types <- as.factor(values)
-  num_types <- length(levels(types))
-  color_map = brewer.pal(num_types, palette)
-  color_map[types]
-  if ((output == "shape") & num_types > 5) {
-    warning("only 5 distinct shapes available: some shapes have been repeated")
-  }
-  l = vector("list")
-  l[["legend.text"]] = levels(types)
-  l[["legend.pch"]] = 21
-  if (output == "color") {
-    l[["legend.col"]] = color_map[1:num_types]
-    l[["color"]] = color_map[types]
-  }
-  if (output == "shape") {
-    l[["legend.pch"]] = ((1:num_types) - 1)%%5 + 21
-    l[["shape"]] = (as.numeric(types) - 1)%%5 + 21
-  }
-  l
-}
+The architectural components consisted of compounds, roomblocks, and
+platform mounds. Compounds were the typical residential structures in
+the area and featured multiple houses built from masonry and/or adobe
+around a courtyard. Platform mounds were formed from large masonry and
+adobe retaining walls with a cell-like interior filled with trash and
+rubble. They were common in the Hohokam region throughout the Classic
+period. Some sites had pit houses from earlier occupations, but most
+excavation was focused on the Roosevelt and Gila phases so these few pit
+houses were removed from the analysis \citep[see][]{Rice1998-ku}. Each
+site was classified as a platform mound if one was present regardless of
+other architecture or as a roomblock if a platform mound was not present
+regardless of other architecture. Thus, if a site is labeled a platform
+mound it may have other architecture. Sites were thus labeled to
+emphasize the most distinctive form of architecture, but all types of
+architecture were present in the analysis, and all architectural
+features included in the analysis can be seen in Table 1.
 
-dists = list()
-```
+\hypertarget{ceramics}{%
+\subsection*{Ceramics}\label{ceramics}}
+\addcontentsline{toc}{subsection}{Ceramics}
 
-```{r}
-# Spatial network
-d = dist(coords[,c("UTMEAST","UTMNORTH")]) %>% as.matrix
-colnames(d) = rownames(d) = coords$id
-# convert to scaled similarity (1 highest)
-d = 1 - (d / max(d))
-dists$spatial = d
-```
+The ceramic network consists of 55 types--plainware and
+decorated--ranging from 181,666 Plain Brown Ware sherds to a single
+Black Mesa Black-on-white sherd. A single sherd is not much, but it is
+still indicative of trade or movement. The median sherd count was 157
+for all types. The ceramic data was accessed from the cyberSW database
+\citep{Mills2020-tb}, which keeps the full citations for the ceramic
+data. The standard caveats for ceramic analysis apply to this analysis
+as well: problems with sherd misidentification and data errors.
 
-## Spatial {-}
+\hypertarget{projectile-points}{%
+\subsection*{Projectile Points}\label{projectile-points}}
+\addcontentsline{toc}{subsection}{Projectile Points}
 
-The simplest network is the spatial network.
-This network was created by calculating the Euclidean distance from every site to every other site.
-Euclidean distance is a shortcut that does not represent actual travel routes.
-Least cost paths provide more accurate data [see @Caseldine2022-uu for an application in Tonto Basin], yet an analysis in similar terrain indicates that for distances longer than those in this study the benefits of least cost path distances in place of Euclidean distance are negligible [@Bischoff2017-gx].
-Least cost paths are also hampered by the presence of Roosevelt Lake and other modern impacts on the landscape.
-The spatial network created here is equivalent to proximal point analysis, which was used in one of the earliest network studies in archaeology [@Terrell1977-co] and continues to be used to study potential pathways of interaction [see @Broodbank2000-pg; @Collar2013-fb].
-The inclusion of a spatial network serves as a null hypothesis to test the other networks against.
+The ceramic and architectural data are available in suitable formats for
+analysis, but the original projectile point analysis was too general for
+the purposes of this study \citep[see][ for details]{Rice1994-rk}.
+Points were divided into longer or shorter categories and subdivided by
+several attributes (e.g., blade, base, or notch shape). Geometric
+morphometrics (GM) is a set of methods designed to quantitatively
+analyze shapes and is ideal for projectile point analysis. See Bischoff
+\citeyearpar{Bischoff2022-rg} for details on GM and the methods used to
+analyze these projectile points. A GM analysis was conducted on the 2D
+outlines of each complete triangular or side-notched projectile point.
+The majority of projectile points were triangular or side-notched. The
+few other points were likely curated points from earlier periods.
+Figures 2 and 3 show the results of these analyses. These figures
+demonstrate how GM methods can calculate the similarity between each
+projectile point shape. These results were classified into closely
+related clusters and used in the same way as ceramic types.
 
-## Architecture {-}
+\begin{figure}
+\includegraphics[width=1\linewidth]{figures/TontoSideDistanceNetwork} \caption{Network graph displaying side-notched points from Tonto Basin as nodes with ties showing the morphometric distance between points.
+Darker colors represent stronger ties.
+Note that only the strongest 10\% of ties are shown.
+From Bischoff (2022: Figure 14).}\label{fig:TontoSideDistanceNetwork}
+\end{figure}
 
-```{r}
-# architecture network
-# convert to adjacency matrix
-tmp = sites %>% 
-  separate_rows(architecture, sep = ", ") %>% 
-  left_join(nodes %>% select(ArchaMapName,id))
-m = table(tmp$id,tmp$architecture) %>% as.matrix
-# calculate jaccard distance
-d = vegdist(m,method = "jaccard", na.rm = T) %>% as.matrix
-# convert to jaccard similarity
-d = 1-d
-dists$architecture = d
-```
+\begin{figure}
+\includegraphics[width=1\linewidth]{figures/TontoTriangularDistanceNetwork} \caption{Network graph displaying triangular points from Tonto Basin as nodes with ties showing the morphometric distance between points.
+Darker colors represent stronger ties.
+Note that only the strongest 10\% of ties are shown.
+From Bischoff (2022: Figure 15).}\label{fig:TontoTriangularDistanceNetwork}
+\end{figure}
 
-The architectural components consisted of compounds, roomblocks, and platform mounds.
-Compounds were the typical residential structures in the area and featured multiple houses built from masonry and/or adobe around a courtyard.
-Platform mounds were formed from large masonry and adobe retaining walls with a cell-like interior filled with trash and rubble.
-They were common in the Hohokam region throughout the Classic period.
-Some sites had pit houses from earlier occupations, but most excavation was focused on the Roosevelt and Gila phases so these few pit houses were removed from the analysis [see @Rice1998-ku].
-Each site was classified as a platform mound if one was present regardless of other architecture or as a roomblock if a platform mound was not present regardless of other architecture.
-Thus, if a site is labeled a platform mound it may have other architecture.
-Sites were thus labeled to emphasize the most distinctive form of architecture, but all types of architecture were present in the analysis, and all architectural features included in the analysis can be seen in Table 1.
+\hypertarget{results}{%
+\section*{Results}\label{results}}
+\addcontentsline{toc}{section}{Results}
 
-## Ceramics {-}
+A major benefit of network analysis is the visual exploration of the
+network. The results of the network analysis are provided partially to
+address the major questions of this paper--the correlations between
+types of material culture and their potential causes--and partially as a
+visual exploration with corresponding evidence from network metrics.
 
-```{r}
-# ceramic stats
-# cdata = ceramicDF %>% rownames_to_column("sites") %>%  janitor::adorn_totals()
-# summary(cdata[16,-1] %>% unlist %>% as.integer)
+Figure 4 shows the networks with the top five strongest ties between
+each node. This number of ties represents the fewest ties that connects
+all nodes in the spatial network. Fewer ties leaves the Cline Mesa sites
+disconnected from the rest of the network. Bass Point Mound forms a
+crucial bridge between the Cline Mesa sites and the rest of the network.
+Because the network is not complete, one cannot argue that there were
+not other sites between Cline Mesa and the Schoolhouse Mesa sites, but
+it still represents an intermediary location. Its geographic position
+overlooking the confluence of the Salt River and Tonto Creek would make
+an ideal meeting place for parties coming down from the Tonto Creek arm
+of what is now Roosevelt Lake or coming up from the Salt River arm. The
+Rock Island area consists of a single site, Bass Point Mound (a platform
+mound), although it was not heavily excavated. If spatial distance was
+an important factor in social interaction, then we would expect Bass
+Point Mound to consistently be a highly central node due to its central
+location. Table 2 shows the multilayer eigenvector centrality for points
+and ceramics only. Bass Point Mound has the highest multilayer network
+centrality with 0.95 (p = 0.87). Figure 5 provides the eigenvector data
+by type of network. Bass Point Mound has low centrality in the
+architectural network and the networks that do not connect the spatial
+components (three and four closest connections), which gives it a lower
+spatial centrality as well. The ceramic and point networks provide
+strong evidence that Bass Point Mound's spatial location was
+advantageous for forming connections between the sites in this study.
 
-# ceramic network
-# convert to adjacency matrix
-m = ceramicDF %>% as.matrix
-# calculate jaccard distance
-d = vegdist(m,method = "jaccard", na.rm = T) %>% as.matrix
-# convert to jaccard similarity
-d = 1-d
-dists$ceramics = d
-```
+\begin{figure}
+\includegraphics[width=1\linewidth]{figures/TontoNetworksMultiNet} \caption{Graphs showing the networks used in this analsyis.
+Each network is displayed with the five strongest ties between each node.
+Legend: 1: AZ U:3:128 (ASM); 2: AZ U:4:032 (ASM); 3: AZ V:5:119 (ASM); 4: Bass Point Mound; 5: Cline Terrace Mound, Monster Ruin; 6: Indian Point Complex; 7: Pinto Point Mound; 8: Saguaro Muerto; 9: Sand Dune Site; 10: Schoolhouse Point Mesa Complex; 11: Schoolhouse Point Mound}\label{fig:TontoNetworksMultiNet}
+\end{figure}
 
-The ceramic network consists of 55 types--plainware and decorated--ranging from 181,666 Plain Brown Ware sherds to a single Black Mesa Black-on-white sherd.
-A single sherd is not much, but it is still indicative of trade or movement.
-The median sherd count was 157 for all types.
-The ceramic data was accessed from the cyberSW database [@Mills2020-tb], which keeps the full citations for the ceramic data.
-The standard caveats for ceramic analysis apply to this analysis as well: problems with sherd misidentification and data errors.
+\begin{table}
 
-## Projectile Points {-}
+\caption{\label{tab:meanEigen}Multilayer Eigenvector Centrality for Ceramics and Points}
+\centering
+\begin{tabular}[t]{lr}
+\toprule
+Site & Centrality\\
+\midrule
+Bass Point Mound & 0.95\\
+Sand Dune Site & 0.84\\
+Pinto Point Mound & 0.76\\
+Indian Point Complex & 0.65\\
+AZ U:3:128 (ASM) & 0.64\\
+\addlinespace
+Schoolhouse Point Mesa Complex & 0.59\\
+Schoolhouse Point Mound & 0.55\\
+Cline Terrace Mound, Monster Ruin & 0.50\\
+Saguaro Muerto & 0.49\\
+AZ U:4:032 (ASM) & 0.46\\
+\addlinespace
+AZ V:5:119 (ASM) & 0.30\\
+\bottomrule
+\end{tabular}
+\end{table}
 
-```{r}
-# projectile point network
-# convert to adjacency matrix
-m = table(outTonto$fac$id,outTonto$fac$pointCluster) %>% as.matrix
-# calculate jaccard distance
-d = vegdist(m,method = "jaccard") %>% as.matrix
-# convert to jaccard similarity
-d = 1-d
-dists$points = d
-```
+Visual inspection of the network graphs provides several other
+observations. One is that the four geographic areas represented do not
+form strong components in the non-spatial networks like they do in the
+spatial network. The architecture is thoroughly mixed, and the ceramics
+and points form single component networks. The formation of single
+components--rather than distinct sections of the graph--is typical of
+artifact similarity networks. The Livingston sites exhibit the most
+clustering in the ceramic network, although some of the connections go
+through Cline Mesa sites, which are as geographically distant as
+possible for these sites. The Schoolhouse Mesa sites always group
+together in the ceramic and point networks, as do the Livingston
+compound sites and separately the Livingston roomblock and platform
+mound. The Cline Mesa sites exhibit more variability.
 
-The ceramic and architectural data are available in suitable formats for analysis, but the original projectile point analysis was too general for the purposes of this study [see @Rice1994-rk for details].
-Points were divided into longer or shorter categories and subdivided by several attributes (e.g., blade, base, or notch shape).
-Geometric morphometrics (GM) is a set of methods designed to quantitatively analyze shapes and is ideal for projectile point analysis.
-See Bischoff [-@Bischoff2022-rg] for details on GM and the methods used to analyze these projectile points.
-A GM analysis was conducted on the 2D outlines of each complete triangular or side-notched projectile point.
-The majority of projectile points were triangular or side-notched.
-The few other points were likely curated points from earlier periods.
-Figures `r figs` and `r figs + 1` show the results of these analyses.
-These figures demonstrate how GM methods can calculate the similarity between each projectile point shape.
-These results were classified into closely related clusters and used in the same way as ceramic types.
+\begin{figure}
+\includegraphics[width=1\linewidth]{figures/TontoNetworksEigenvector} \caption{Boxplots showing the eigenvector centrality for each network.
+Values were calculated using networks created using between 3 to 10 of the strongest ranking ties.
+Legend: 1: AZ U:3:128 (ASM); 2: AZ U:4:032 (ASM); 3: AZ V:5:119 (ASM); 4: Bass Point Mound; 5: Cline Terrace Mound, Monster Ruin; 6: Indian Point Complex; 7: Pinto Point Mound; 8: Saguaro Muerto; 9: Sand Dune Site; 10: Schoolhouse Point Mesa Complex; 11: Schoolhouse Point Mound.}\label{fig:TontoNetworksEigenvector}
+\end{figure}
 
-```{r TontoSideDistanceNetwork, fig.cap= 'Network graph displaying side-notched points from Tonto Basin as nodes with ties showing the morphometric distance between points.
-Darker colors represent stronger ties.
-Note that only the strongest 10\\% of ties are shown.
-From Bischoff (2022: Figure 14).'}
-figs = figs + 1
-knitr::include_graphics(here("figures/TontoSideDistanceNetwork.png"))
-```
+Another observation from the visual inspection of these networks is that
+there is some clustering due to architecture in the ceramic and point
+networks. Four of the five compound-only sites cluster at one end of the
+ceramic network, and a different mix of four out of five cluster
+together at one end of the projectile point network. The platform mounds
+are split in the ceramic network, but all cluster together in the
+projectile point network. Table 3 shows the mean eigenvector centrality
+vectors for the ceramic and point networks by type of architecture.
+Surprisingly the platform mounds are the least central, on average, for
+the ceramic network, but they are the most central for the projectile
+point network. The roomblocks, on the hand, are the most central for the
+ceramic network and by far the least central in the projectile point
+network. They take a larger drop in centrality in the projectile point
+network than the platform mound sites do in the ceramic network. It is
+expected that roomblocks have high centrality in the ceramic network,
+but not that platform mounds should have the lowest centrality. Clearly,
+centrality in the projectile point network does not equal centrality in
+the ceramic network.
 
-```{r TontoTriangularDistanceNetwork, fig.cap= 'Network graph displaying triangular points from Tonto Basin as nodes with ties showing the morphometric distance between points.
-Darker colors represent stronger ties.
-Note that only the strongest 10\\% of ties are shown.
-From Bischoff (2022: Figure 15).'}
-figs = figs + 1
-knitr::include_graphics(here("figures/TontoTriangularDistanceNetwork.png"))
-```
+\begin{table}
 
-# Results {-}
+\caption{\label{tab:meanEigenVals}Mean Eigenvalues for Ceramic and Point Networks by Type of Architecture}
+\centering
+\begin{tabular}[t]{lrrr}
+\toprule
+Architecture & ceramics & points & mean\\
+\midrule
+platform mound & 0.59 & 0.79 & 0.69\\
+roomblock & 0.76 & 0.38 & 0.57\\
+compound & 0.65 & 0.48 & 0.56\\
+\bottomrule
+\end{tabular}
+\end{table}
 
-```{r include=FALSE}
-corr = list()
-eigen = list()
-for (nLinks in 3:10){
-  print(nLinks)
-  net = ml_empty("Tonto")
-  g2List = list()
-  nms = names(dists)
-  for(n in nms){
-    print(n)
-    g = makeGraph(dists[[n]], nLinks = nLinks)
-    eigen[[paste0(nLinks,"-",n)]] = V(g)$centrality
-    g2List[[n]] = g
-    add_igraph_layer_ml(n = net, g = g,name = n)
-  }
+The visual analysis, combined with the eigenvector centrality, indicated
+some spatial and architectural correlation. The multilayer Pearson
+correlation provides a more direct way to compare these layers, as shown
+in figure 6. This analysis provides a clear contrast between layers.
+None of the results, on average, showed a negative correlation--meaning
+that ties existed in one network where ties did not exist in the other
+network. Only one comparison had a strong correlation--the point and
+spatial networks were strongly correlated (\textbf{x} = 0.63; p = 0.98).
+The visual inspection and centrality analysis provided some indication
+of this, but the multilayer network comparison provides strong
+verification. All other layer comparisons had an approximately
+equivalent, positive correlation, but with only a weak strength. The one
+exception was architecture and points where the correlation was
+approximately zero. The visual inspection and eigenvector analysis
+demonstrated some interesting interactions with architecture in the
+network, but the overall correlation demonstrates that architecture
+cannot be used to predict the presence or absence of network ties in the
+projectile point network.
 
-  # layer_comparison_ml(net, method = "jaccard.edges", mode = "all") %>% 
-  # print
-  corr[[nLinks]] = layer_comparison_ml(net, method = "pearson.degree", mode = "all") %>% as.data.frame() %>% 
-    mutate(nLinks = nLinks)
-}
+\begin{figure}
+\includegraphics[width=1\linewidth]{figures/TontoNetworksCorrelation} \caption{Boxplots showing the Pearson correlation between each network.
+Values were calculated using networks created from 3 to 10 of the strongest ranking ties.}\label{fig:TontoNetworksCorrelation}
+\end{figure}
 
-corrDF = bind_rows(corr) %>% 
-  rownames_to_column("from") %>% 
-  pivot_longer(-c(from,nLinks), names_to = "to") %>% 
-  mutate(from = str_remove_all(from,"[^A-z]")) %>% 
-  mutate(nLinks = factor(as.character(nLinks))) %>% 
-  rowwise()  %>% 
-  filter(from != to) %>% 
-  mutate(id = paste(sort(c(from,to)),collapse = "->")) %>% 
-  select(-from,-to) %>% 
-  distinct_all()#%>% 
-# pivot_wider(names_from = nLinks)
-corrDF %>% 
-  ggplot(aes(x = value, y = id)) + 
-  geom_boxplot() +
-  theme_gdocs() +
-  xlab("Pearson correlation") +
-  ylab("Network comparison")
-# ggsave(here("figures/TontoNetworksCorrelation.png"),dpi = 450, width = 6.5, height = 4.5)
-# browseURL(here("figures/TontoNetworksCorrelation.png"))
+\hypertarget{discusssion}{%
+\section*{Discusssion}\label{discusssion}}
+\addcontentsline{toc}{section}{Discusssion}
 
-eigenDF = bind_rows(eigen) %>% 
-  rowid_to_column("id") %>% 
-  pivot_longer(-id) %>% 
-  left_join(nodes) %>% 
-  mutate(name = str_remove_all(name,"[^A-z]"),
-         id = factor(id))
+This analysis has three main findings regarding networks in Tonto Basin:
+(1) that point networks correlate with space; (2) that roomblock sites
+are highly central in the ceramic network and have low centrality in the
+point network; and (3) that ceramic and point networks are significantly
+different from each other. In terms of the null hypotheses stated in the
+introduction, only one of the networks exhibited strong spatial
+correlation and while all but one pairing (architecture and points)
+exhibited some positive correlation, all of the correlations except the
+points and space were weak. Given these network results, how might they
+be interpreted in terms of social behavior?
 
-eigenDF %>% 
-  ggplot(aes(x = value, y = id, color = Architecture)) + 
-  geom_boxplot() +
-  theme_gdocs() + 
-  xlab("Eigenvector centrality") +
-  ylab("Site ID") +
-  guides(color = guide_legend(title = element_blank(), nrow = 1)) +
-  theme(legend.position = "bottom") +
-  facet_wrap(~name, scales = "free_x")
-# ggsave(here("figures/TontoNetworksEigenvector.png"),dpi = 450, width = 6.5, height = 8.5)
-# browseURL(here("figures/TontoNetworksEigenvector.png"))
+Exchange would be a primary driver in these network dynamics, and clear
+evidence exists for exchange of several types of material culture. Trade
+within the basin does not appear to have been dominated by any one group
+and may have been competitive in nature \citep[p.~127-128]{Rice1998-cu}.
+Ceramic exchange certainly occurred at a regional level, but even local
+ceramics were circulated within the basin
+\citep{Heidke2000-nr, Miksa1995-qw, Stark1992-vu}, possibly in return
+for food \citep[p.290-291]{Clark2004-uw}. Projectile points (either as
+part of the arrow or separately) may have also been commonly exchanged.
+There are several examples of the exchange of bows and arrows
+ethnographically throughout the world
+\citep[e.g.,][]{Mauss1966-nm, Nishiaki2013-lc, Wiessner1983-ei}, in
+North America generally \citep[e.g.,][]{Hoffman1896-xl, Radin1923-gf},
+and in the Southwest specifically
+\citep[e.g.,][]{Beaglehole1936-ul, Dittert1959-rz, Fewkes1898-oj, griffen1969a, Parsons1939-xg, Simpson1953-ob}.
+Besides Watt's study of knappers, Sliva's
+\citeyearpar[p.~539]{Sliva2002-oz} analysis of sites from the Roosevelt
+Community Development Study suggests craft specialization in projectile
+point production. Mortuary offerings were possibly the intended function
+for many points \citep[p.~543]{Sliva2002-oz}, which may have spurred the
+increased craft specialization. Shell and stone jewelry, among other
+artifacts were also produced and exchanged within the basin. A potential
+loci for exchange are the platform mound sites. The purpose of these
+sites is still debated, but many argue that they served an integrative
+function
+\citetext{\citealp[e.g.,][]{Abbott2006-fg}; \citealp{Adler1990-ws}; \citealp{Clark2004-uw}; \citealp{Craig1995-qo}; \citealp[p.~112-165]{Craig1994-bg}; \citealp[p.~439]{Doelle1995-qd}}.
+If platform mounds were places where exchange between communities
+regularly took place, then we would expect platform mounds to have
+higher centrality. Centrality varied among the platform mounds, but on
+average, platform mounds did have the highest centrality.
 
-g2List = list()
-nms = names(dists)
-for(n in nms){
-  print(n)
-  g = makeGraph(dists[[n]], nLinks = 5)
-  eigen[[paste0(nLinks,"-",n)]] = V(g)$centrality
-  g2List[[n]] = g
-}
+As discussed previously, cultural differences were expected between the
+Tonto Creek and Salt River arms of Roosevelt Lake. This was due in part
+to differences in long-distance ceramic exchange and obsidian sourcing
+\citep{Lyons2013-ya, Simon2001-am}. This analysis did not use obsidian
+or focus on non-local ceramics and therefore did not capture differences
+based on these factors. There was some expectation that more separation
+in the material networks would be apparent between the Cline Mesa sites
+in the Tonto Creek arm and the other sites (minus Bass Point Mound that
+lies at the confluence). This was not born out in the visual inspection
+of the analysis with significant mixing of the areas in the network.
+There were, however, indications that spatial distance was an important
+factor. Spatial distance would align with the proposed differences in
+the Tonto Creek and Salt River arms of the lake. The only strong
+correlation between networks was the point and spatial network, which
+indicates that being near another site is a good indication of the shape
+of a projectile point. There was however, little correlation overall
+between ceramics and spatial distance.
 
-netDF = map2_df(g2List,names(g2List),function(g,n){
-  ggnetwork(g) %>% mutate(layer = n)
-})
+There is a potential chain of association between immigrants and
+roomblocks, roomblocks or locations with roomblocks as centers of Salado
+production, and Salado pottery as a widespread phenomenon that provides
+an expectation for roomblock sites to be central to pottery networks.
+While this association is circumstantial and not expected to be uniform
+across the Hohokam region, roomblocks did have the highest centrality in
+the ceramic networks. Another reason for the high centrality of
+roomblock sites is that they lacked access to adequate farmland and had
+to trade pottery for food \citep[p.~291]{Clark2004-uw}. This second
+explanation may help explain why roomblock sites had low centrality in
+the point networks, presuming that points were not exchanged for food.
+It is beyond this simple analysis to determine the precise reasons, but
+it is clear that roomblock sites were important for ceramic circulation
+in Tonto Basin. On the other hand, sites with roomblocks had the lowest
+centrality in the point networks. This suggests that the influence the
+occupants of roomblock sites had in pottery networks may not have
+extended to other spheres of interaction.
 
-ggplot(netDF,aes(x = x,y = y, xend = xend, yend = yend, shape = Architecture, color = Area, label = name, size = centrality)) +
-  geom_edges(color = "gray", size = .75) +
-  geom_nodes() +
-  geom_nodetext_repel(color = "black", size = 3) +
-  scale_shape_manual(values = 15:18) +
-  scale_color_brewer(palette = "Set1") +
-  scale_size_continuous(range = c(3,4.5)) +
-  facet_wrap(~layer) +
-  theme_blank() +
-  theme(panel.border = element_rect(fill = NA),
-        strip.background = element_rect(fill = "white"),
-        legend.position = "bottom") +
-  guides(size = "none", shape = guide_legend(ncol = 2,title = NULL), color = guide_legend(ncol = 2, title = NULL))
-# ggsave(here("figures/TontoNetworksMultiNet.png"),dpi = 450, width = 6.5, height = 8)
-# browseURL(here("figures/TontoNetworksMultiNet.png"))
-paste0(nodes$id,": ",nodes$ArchaMapName) %>% paste(collapse = "; ")
+As discussed previously, my simplistic model of interaction in Tonto
+Basin assumes that the ceramic network represents categorical
+identification among women. Recall also that I expected architecture to
+represent categorical identity. Thus, the correlation between
+architecture and ceramics, at least as represented by roomblocks, is an
+expected find and good corroborating evidence that these types of
+material culture represent markers of identity demonstrating belonging
+to a particular social group.
 
-meanEigenVals = eigenDF %>% 
-  group_by(id,Area,Architecture,name) %>% 
-  summarize(value = round(mean(value),2)) %>% 
-  arrange(name) %>% 
-  pivot_wider(names_from = name, values_from = value) %>% 
-  mutate(id = as.integer(id)) %>% 
-  left_join(nodes %>% select(id,Site = ArchaMapName)) %>% 
-  ungroup()
+The point networks were expected to represent relational identification
+among men, at least in this case study. Because relational
+identification is related to frequent interaction, spatial distance is a
+crucial component. It is much harder to interact with someone when they
+are far away. Thus, the correlation between the point and spatial
+networks also makes theoretical sense.
 
-meanEigenValsSites = meanEigenVals %>% 
-  select(Site,Area,Architecture,ceramics,points) %>% 
-  rowwise() %>% 
-  mutate(Centrality = round(mean(c(ceramics,points)),2)) %>% 
-  select(-ceramics,-points)
+Perhaps the reason roomblocks were more central to the ceramic network
+is because immigrants to the basin had less access to farmland and had
+to make pottery to get food, as mentioned. This would explain higher
+centrality for roomblocks and lower centrality for point networks. It
+does not explain the remaining differences between point and ceramic
+networks. Projectile points at least were highly gendered, and ceramics
+probably were as well. The evidence presented here demonstrates that
+ceramic and point networks vary significantly, and I believe gender
+likely played an important role.
 
-# # resampling analysis
-# # purpose reshuffle the points and ceramics to determine if the correlations can appear by chance
-# library(furrr)
-# future::plan('multisession')
-# sensitivityResults = furrr::future_map(1:10000, function(s){
-#   corr = list()
-#   eigen = list()
-#   for (nLinks in 3:10){
-#     net = ml_empty("Tonto")
-#     g2List = list()
-#     nms = names(dists)
-#     for(n in nms){
-#       m = dists[[n]]
-#       m = matrix(sample(m,121, replace = T),ncol = 11)
-#       g = makeGraph(m, nLinks = nLinks)
-#       eigen[[paste0(nLinks,"-",n)]] = V(g)$centrality
-#       g2List[[n]] = g
-#       add_igraph_layer_ml(n = net, g = g,name = n)
-#     }
-#     corr[[nLinks]] = layer_comparison_ml(net, method = "pearson.degree", mode = "all") %>% as.data.frame() %>%
-#       mutate(nLinks = nLinks)
-#   }
-# 
-#   corrDF = bind_rows(corr) %>%
-#     rownames_to_column("from") %>%
-#     pivot_longer(-c(from,nLinks), names_to = "to") %>%
-#     mutate(from = str_remove_all(from,"[^A-z]")) %>%
-#     mutate(nLinks = factor(as.character(nLinks))) %>%
-#     rowwise()  %>%
-#     filter(from != to) %>%
-#     mutate(id = paste(sort(c(from,to)),collapse = "->")) %>%
-#     select(-from,-to) %>%
-#     distinct_all() %>%
-#     mutate(run = s)
-# 
-#   eigenDF = bind_rows(eigen) %>%
-#     rowid_to_column("id") %>%
-#     pivot_longer(-id) %>%
-#     left_join(nodes) %>%
-#     mutate(name = str_remove_all(name,"[^A-z]"),
-#            id = factor(id),
-#            run = s)
-#   return(list(corrDF = corrDF, eigenDF = eigenDF))
-# })
-# sensitivityCorr = map_df(sensitivityResults,function(i)bind_rows(i$corrDF))
-# saveRDS(sensitivityCorr,"data/SensitivityCorr.Rds")
-sensitivityCorr = readRDS("data/SensitivityCorr.Rds")
-# sensitivityEigen = map_df(sensitivityResults,function(i)bind_rows(i$eigenDF))
-# saveRDS(sensitivityEigen,"data/sensitivityEigen.Rds")
-sensitivityEigen = readRDS("data/sensitivityEigen.Rds")
-pF = function(val,var, df = "sensitivityEigen"){
-  if(df == "sensitivityEigen"){
-  df = get(df)
-  sub = df %>% filter(name %in% var)
-  ecdfF = ecdf(sub$value)
-  round(ecdfF(val),2)
-  }else{
-  df = get(df)
-  sub = df %>% filter(id %in% var)
-  ecdfF = ecdf(sub$value)
-  round(ecdfF(val),2)
-  }
-}
-# pF(.95,"ceramics",df = "sensitivityEigen")
-# pF(.35,"architecture->ceramics",df = "sensitivityCorr")
-```
+\hypertarget{conclusion}{%
+\section*{Conclusion}\label{conclusion}}
+\addcontentsline{toc}{section}{Conclusion}
 
-A major benefit of network analysis is the visual exploration of the network.
-The results of the network analysis are provided partially to address the major questions of this paper--the correlations between types of material culture and their potential causes--and partially as a visual exploration with corresponding evidence from network metrics.
+This is one of the first archaeological applications of multilayer
+network analysis to consider multiple types of material culture. This is
+advantageous for studying how types of material culture do or do not
+co-vary, which aids interpretations of the social interactions that
+created these patterns. This analysis used architectural, ceramic,
+projectile point, and spatial data from 11 sites in Tonto Basin. The
+data come primarily from occupations between AD 1275 and 1325. A network
+was created for each type of data and combined into a multilayer
+network. Visual network analysis, eigenvector centrality, and multilayer
+network Pearson correlation were used to study and compare the networks.
+The findings indicate that the projectile point network was strongly
+correlated with the spatial network--indicating that sites near each
+other were more likely to have similar projectile points. Furthermore,
+sites with roomblocks indicative of immigrants from the north and east
+of Tonto Basin were, on average, the most highly central sites in the
+ceramic network; however, they were the least central in the point
+network. Immigrants may have relied on pottery production to integrate
+with the local networks, but this relationship did not hold for
+projectile points. Finally, the results demonstrate major differences
+between the ceramic and point networks. This findings has major
+implications for network studies, because many rely on only a single
+artifact type. If different types of material culture do not co-vary
+regularly, then that indicates archaeologists must do more to include
+other types of material culture to better understand the complex social
+networks that existed in the past. Furthermore, projectile points and
+ceramics have strong associations with gender. Differences in the
+ceramic and point networks suggest differences existed between the
+social networks of each gender.
 
-Figure `r figs` shows the networks with the top five strongest ties between each node.
-This number of ties represents the fewest ties that connects all nodes in the spatial network.
-Fewer ties leaves the Cline Mesa sites disconnected from the rest of the network.
-Bass Point Mound forms a crucial bridge between the Cline Mesa sites and the rest of the network.
-Because the network is not complete, one cannot argue that there were not other sites between Cline Mesa and the Schoolhouse Mesa sites, but it still represents an intermediary location.
-Its geographic position overlooking the confluence of the Salt River and Tonto Creek would make an ideal meeting place for parties coming down from the Tonto Creek arm of what is now Roosevelt Lake or coming up from the Salt River arm.
-The Rock Island area consists of a single site, Bass Point Mound (a platform mound), although it was not heavily excavated.
-If spatial distance was an important factor in social interaction, then we would expect Bass Point Mound to consistently be a highly central node due to its central location.
-Table `r tbls` shows the multilayer eigenvector centrality for points and ceramics only.
-Bass Point Mound has the highest multilayer network centrality with 0.95 (p = `r pF(.95,c("ceramics","points"),df = "sensitivityEigen")`).
-Figure `r figs + 1` provides the eigenvector data by type of network.
-Bass Point Mound has low centrality in the architectural network and the networks that do not connect the spatial components (three and four closest connections), which gives it a lower spatial centrality as well.
-The ceramic and point networks provide strong evidence that Bass Point Mound's spatial location was advantageous for forming connections between the sites in this study.
+This analysis uses a handful of sites from Tonto Basin. Much more data
+exists and the results discussed here would be greatly strengthened by
+including more sites and data. What would be perhaps even more useful
+would be to include sourced obsidian and ceramics. Networks created from
+artifacts where the origin is known can be used to more strongly infer
+various types of social interaction. This type of multilayer analysis
+would also be useful to conduct in other regions. It would be
+particularly useful for comparative studies to know in what
+circumstances different types of material culture do or do not co-vary.
 
-```{r TontoNetworksMultiNet, fig.cap= 'Graphs showing the networks used in this analsyis.
-Each network is displayed with the five strongest ties between each node.
-Legend: 1: AZ U:3:128 (ASM); 2: AZ U:4:032 (ASM); 3: AZ V:5:119 (ASM); 4: Bass Point Mound; 5: Cline Terrace Mound, Monster Ruin; 6: Indian Point Complex; 7: Pinto Point Mound; 8: Saguaro Muerto; 9: Sand Dune Site; 10: Schoolhouse Point Mesa Complex; 11: Schoolhouse Point Mound'}
-figs = figs + 1
-knitr::include_graphics(here("figures/TontoNetworksMultiNet.png"))
-```
+\hypertarget{acknowledgements}{%
+\section*{Acknowledgements}\label{acknowledgements}}
+\addcontentsline{toc}{section}{Acknowledgements}
 
-```{r meanEigen, echo=FALSE}
-tbls = tbls + 1
-cap = "Multilayer Eigenvector Centrality for Ceramics and Points"
-meanEigenValsSites %>% 
-  select(Site,Centrality) %>% 
-  arrange(desc(Centrality)) %>% 
-  knitr::kable(caption= cap,
-               booktabs = T,
-               format.args = list(big.mark = ",",
-                                  scientific = FALSE))
-  # as_hux() %>% 
-  # theme_article() %>% 
-  # set_caption(cap)
+Josh Watts, Matt Peeples, Melissa Powell, Chris Caseldine, and several
+volunteers assisted with this research. Chris Caseldine's comments
+greatly improved this paper. All mistakes are my own. The many
+individuals who contributed to the Roosevelt Platform Mound Study also
+deserve recognition.
 
-```
-Visual inspection of the network graphs provides several other observations.
-One is that the four geographic areas represented do not form strong components in the non-spatial networks like they do in the spatial network.
-The architecture is thoroughly mixed, and the ceramics and points form single component networks.
-The formation of single components--rather than distinct sections of the graph--is typical of artifact similarity networks.
-The Livingston sites exhibit the most clustering in the ceramic network, although some of the connections go through Cline Mesa sites, which are as geographically distant as possible for these sites.
-The Schoolhouse Mesa sites always group together in the ceramic and point networks, as do the Livingston compound sites and separately the Livingston roomblock and platform mound.
-The Cline Mesa sites exhibit more variability.
-
-```{r TontoNetworksEigenvector, fig.cap= 'Boxplots showing the eigenvector centrality for each network.
-Values were calculated using networks created using between 3 to 10 of the strongest ranking ties.
-Legend: 1: AZ U:3:128 (ASM); 2: AZ U:4:032 (ASM); 3: AZ V:5:119 (ASM); 4: Bass Point Mound; 5: Cline Terrace Mound, Monster Ruin; 6: Indian Point Complex; 7: Pinto Point Mound; 8: Saguaro Muerto; 9: Sand Dune Site; 10: Schoolhouse Point Mesa Complex; 11: Schoolhouse Point Mound.'}
-figs = figs + 1
-knitr::include_graphics(here("figures/TontoNetworksEigenvector.png"))
-```
-
-Another observation from the visual inspection of these networks is that there is some clustering due to architecture in the ceramic and point networks.
-Four of the five compound-only sites cluster at one end of the ceramic network, and a different mix of four out of five cluster together at one end of the projectile point network.
-The platform mounds are split in the ceramic network, but all cluster together in the projectile point network.
-Table `r tbls` shows the mean eigenvector centrality vectors for the ceramic and point networks by type of architecture.
-Surprisingly the platform mounds are the least central, on average, for the ceramic network, but they are the most central for the projectile point network.
-The roomblocks, on the hand, are the most central for the ceramic network and by far the least central in the projectile point network.
-They take a larger drop in centrality in the projectile point network than the platform mound sites do in the ceramic network.
-It is expected that roomblocks have high centrality in the ceramic network, but not that platform mounds should have the lowest centrality.
-Clearly, centrality in the projectile point network does not equal centrality in the ceramic network.
-
-```{r meanEigenVals}
-cap = "Mean Eigenvalues for Ceramic and Point Networks by Type of Architecture"
-eigenDF %>% 
-  filter(name %in% c("ceramics","points")) %>% 
-  group_by(name,Architecture) %>% 
-  summarize(value = round(mean(value),2)) %>% 
-  pivot_wider(names_from = name, values_from = value) %>% 
-  ungroup() %>% 
-  rowwise() %>% 
-  mutate(mean = round(mean(c(ceramics,points)),2)) %>% 
-  arrange(desc(mean)) %>%
-  # as_hux() %>% 
-  # theme_article() %>% 
-  # set_caption(cap)
-  knitr::kable(caption= cap,
-               booktabs = T)
-```
-
-The visual analysis, combined with the eigenvector centrality, indicated some spatial and architectural correlation.
-The multilayer Pearson correlation provides a more direct way to compare these layers, as shown in figure `r figs`.
-This analysis provides a clear contrast between layers.
-None of the results, on average, showed a negative correlation--meaning that ties existed in one network where ties did not exist in the other network.
-Only one comparison had a strong correlation--the point and spatial networks were strongly correlated (**x** = `r psCorr = corrDF %>% filter(id == "points->spatial") %>% pull(value) %>% mean %>% round(.,2); psCorr`; p = `r pF(psCorr,"points->spatial",df = "sensitivityCorr")`).
-The visual inspection and centrality analysis provided some indication of this, but the multilayer network comparison provides strong verification.
-All other layer comparisons had an approximately equivalent, positive correlation, but with only a weak strength.
-The one exception was architecture and points where the correlation was approximately zero.
-The visual inspection and eigenvector analysis demonstrated some interesting interactions with architecture in the network, but the overall correlation demonstrates that architecture cannot be used to predict the presence or absence of network ties in the projectile point network.
-
-```{r TontoNetworksCorrelation, fig.cap= 'Boxplots showing the Pearson correlation between each network. Values were calculated using networks created from 3 to 10 of the strongest ranking ties.'}
-figs = figs + 1
-knitr::include_graphics(here("figures/TontoNetworksCorrelation.png"))
-```
-
-# Discusssion {-}
-
-This analysis has three main findings regarding networks in Tonto Basin: (1) that point networks correlate with space; (2) that roomblock sites are highly central in the ceramic network and have low centrality in the point network; and (3) that ceramic and point networks are significantly different from each other.
-In terms of the null hypotheses stated in the introduction, only one of the networks exhibited strong spatial correlation and while all but one pairing (architecture and points) exhibited some positive correlation, all of the correlations except the points and space were weak.
-Given these network results, how might they be interpreted in terms of social behavior?
-
-Exchange would be a primary driver in these network dynamics, and clear evidence exists for exchange of several types of material culture.
-Trade within the basin does not appear to have been dominated by any one group and may have been competitive in nature [@Rice1998-cu, p.
-127-128].
-Ceramic exchange certainly occurred at a regional level, but even local ceramics were circulated within the basin [@Heidke2000-nr; @Miksa1995-qw; @Stark1992-vu], possibly in return for food [@Clark2004-uw, p.290-291].
-Projectile points (either as part of the arrow or separately) may have also been commonly exchanged.
-There are several examples of the exchange of bows and arrows ethnographically throughout the world [e.g., @Mauss1966-nm; @Nishiaki2013-lc; @Wiessner1983-ei], in North America generally [e.g., @Hoffman1896-xl; @Radin1923-gf], and in the Southwest specifically [e.g., @Beaglehole1936-ul; @Dittert1959-rz; @Fewkes1898-oj; @griffen1969a; @Parsons1939-xg; @Simpson1953-ob].
-Besides Watt's study of knappers, Sliva's [-@Sliva2002-oz, p.
-539] analysis of sites from the Roosevelt Community Development Study suggests craft specialization in projectile point production.
-Mortuary offerings were possibly the intended function for many points [@Sliva2002-oz, p.
-543], which may have spurred the increased craft specialization.
-Shell and stone jewelry, among other artifacts were also produced and exchanged within the basin.
-A potential loci for exchange are the platform mound sites.
-The purpose of these sites is still debated, but many argue that they served an integrative function [e.g., @Abbott2006-fg; @Adler1990-ws; @Clark2004-uw; @Craig1995-qo; @Craig1994-bg, p.
-112-165; @Doelle1995-qd, p.
-439].
-If platform mounds were places where exchange between communities regularly took place, then we would expect platform mounds to have higher centrality.
-Centrality varied among the platform mounds, but on average, platform mounds did have the highest centrality.
-
-As discussed previously, cultural differences were expected between the Tonto Creek and Salt River arms of Roosevelt Lake.
-This was due in part to differences in long-distance ceramic exchange and obsidian sourcing [@Lyons2013-ya; @Simon2001-am].
-This analysis did not use obsidian or focus on non-local ceramics and therefore did not capture differences based on these factors.
-There was some expectation that more separation in the material networks would be apparent between the Cline Mesa sites in the Tonto Creek arm and the other sites (minus Bass Point Mound that lies at the confluence).
-This was not born out in the visual inspection of the analysis with significant mixing of the areas in the network.
-There were, however, indications that spatial distance was an important factor.
-Spatial distance would align with the proposed differences in the Tonto Creek and Salt River arms of the lake.
-The only strong correlation between networks was the point and spatial network, which indicates that being near another site is a good indication of the shape of a projectile point.
-There was however, little correlation overall between ceramics and spatial distance.
-
-There is a potential chain of association between immigrants and roomblocks, roomblocks or locations with roomblocks as centers of Salado production, and Salado pottery as a widespread phenomenon that provides an expectation for roomblock sites to be central to pottery networks.
-While this association is circumstantial and not expected to be uniform across the Hohokam region, roomblocks did have the highest centrality in the ceramic networks.
-Another reason for the high centrality of roomblock sites is that they lacked access to adequate farmland and had to trade pottery for food [@Clark2004-uw, p.
-291].
-This second explanation may help explain why roomblock sites had low centrality in the point networks, presuming that points were not exchanged for food.
-It is beyond this simple analysis to determine the precise reasons, but it is clear that roomblock sites were important for ceramic circulation in Tonto Basin.
-On the other hand, sites with roomblocks had the lowest centrality in the point networks.
-This suggests that the influence the occupants of roomblock sites had in pottery networks may not have extended to other spheres of interaction.
-
-As discussed previously, my simplistic model of interaction in Tonto Basin assumes that the ceramic network represents categorical identification among women.
-Recall also that I expected architecture to represent categorical identity.
-Thus, the correlation between architecture and ceramics, at least as represented by roomblocks, is an expected find and good corroborating evidence that these types of material culture represent markers of identity demonstrating belonging to a particular social group.
-
-The point networks were expected to represent relational identification among men, at least in this case study.
-Because relational identification is related to frequent interaction, spatial distance is a crucial component.
-It is much harder to interact with someone when they are far away.
-Thus, the correlation between the point and spatial networks also makes theoretical sense.
-
-Perhaps the reason roomblocks were more central to the ceramic network is because immigrants to the basin had less access to farmland and had to make pottery to get food, as mentioned.
-This would explain higher centrality for roomblocks and lower centrality for point networks.
-It does not explain the remaining differences between point and ceramic networks.
-Projectile points at least were highly gendered, and ceramics probably were as well.
-The evidence presented here demonstrates that ceramic and point networks vary significantly, and I believe gender likely played an important role.
-
-# Conclusion {-}
-
-This is one of the first archaeological applications of multilayer network analysis to consider multiple types of material culture.
-This is advantageous for studying how types of material culture do or do not co-vary, which aids interpretations of the social interactions that created these patterns.
-This analysis used architectural, ceramic, projectile point, and spatial data from `r nrow(sites)` sites in Tonto Basin.
-The data come primarily from occupations between AD 1275 and 1325.
-A network was created for each type of data and combined into a multilayer network.
-Visual network analysis, eigenvector centrality, and multilayer network Pearson correlation were used to study and compare the networks.
-The findings indicate that the projectile point network was strongly correlated with the spatial network--indicating that sites near each other were more likely to have similar projectile points.
-Furthermore, sites with roomblocks indicative of immigrants from the north and east of Tonto Basin were, on average, the most highly central sites in the ceramic network; however, they were the least central in the point network.
-Immigrants may have relied on pottery production to integrate with the local networks, but this relationship did not hold for projectile points.
-Finally, the results demonstrate major differences between the ceramic and point networks.
-This findings has major implications for network studies, because many rely on only a single artifact type.
-If different types of material culture do not co-vary regularly, then that indicates archaeologists must do more to include other types of material culture to better understand the complex social networks that existed in the past.
-Furthermore, projectile points and ceramics have strong associations with gender.
-Differences in the ceramic and point networks suggest differences existed between the social networks of each gender.
-
-This analysis uses a handful of sites from Tonto Basin.
-Much more data exists and the results discussed here would be greatly strengthened by including more sites and data.
-What would be perhaps even more useful would be to include sourced obsidian and ceramics.
-Networks created from artifacts where the origin is known can be used to more strongly infer various types of social interaction.
-This type of multilayer analysis would also be useful to conduct in other regions.
-It would be particularly useful for comparative studies to know in what circumstances different types of material culture do or do not co-vary.
-
-# Acknowledgements {-}
-
-Josh Watts, Matt Peeples, Melissa Powell, Chris Caseldine, and several volunteers assisted with this research.
-Chris Caseldine's comments greatly improved this paper.
-All mistakes are my own.
-The many individuals who contributed to the Roosevelt Platform Mound Study also deserve recognition.
-
-# Disclosure Statement {-}
+\hypertarget{disclosure-statement}{%
+\section*{Disclosure Statement}\label{disclosure-statement}}
+\addcontentsline{toc}{section}{Disclosure Statement}
 
@@
 
-# References {-}
+\bibliographystyle{tfcad}
+\bibliography{references2.bib}
+
+\end{document}

### 2022-12-28 manuscript.Rmd
Added additional references for women making pottery.

@@
 Women made pottery and men made projectile points.
-This was not always true of course, but this fits the available data for the Hohokam [@Crown1996-xb; @VanPool2010-im, p.
-253].
+This was not always true of course, but this fits the available data and expectations for the Hohokam [@Crown1996-xb; @Harry2010-hj; @VanPool2010-im, p.
+253; @Whittlesey2010-ox].
+An examination of the ethnographic record for the O'odham people, recognized descendants of the Hohokam, provides additional references that women made pottery [@Bahr2011-or, p.4; @Castetter1935-pq, p.
+5-6; @Chona1936-mg, p.
+44; @Joseph1949-ga, p.
+57].
 An examination of the burial record for the RPMS study shows no obvious indication that women made pottery, only that both men and women were buried with pottery [@Loendorf1998-ln, Table 10.7].

### 2023-01-11 manuscript.Rmd
updated with new references and discussion

@@
 output:
-  rticles::tf_article: null
   bookdown::word_document2:
@@
     reference_docx: TF.docx
+  rticles::tf_article: null
 author:
@@
 Many studies consider material culture holistically, yet network analysis has the advantage of focusing specifically on the relationships between entities (often communities) and how the structure of a network can benefit particular entities.
-This study takes architecture, ceramic, projectile point, and site location data from the Roosevelt Platform Mound Study and combines these data into a multilayer network analysis.
+This study uses architecture, ceramic, projectile point, and site location data from the Roosevelt Platform Mound Study and combines these data in a multilayer network analysis.
 This analysis provides a way to test the co-variance of these types of material culture with each other and with spatial variation.
 The results demonstrate a strong association between projectile points and spatial distance.
-Sites containing roomblock architecture are highly central in the ceramic network but have low centrality in the projectile point network.
 Overall, the ceramic and projectile point networks exhibit significant differences.
 This indicates that the social networks that created these patterns had different social mechanisms.
-One potential cause of these differences is gendered spheres of interaction with men exchanging projectile points and women exchanging ceramics.
+One potential cause of these differences is gendered spheres of interaction with men producing and exchanging projectile points and women producing and exchanging ceramics.
 bibliography: references2.bib
@@
 
-Network science has many applications for archaeologists.
-It can be particularly useful for breaking out of traditional spatial categories and examining data through new lenses [@Feinman2020-nf; @Holland-Lulewicz2021-eb].
+Network science has many applications for archaeologists, and it can be particularly useful for breaking out of traditional spatial categories and examining data through new lenses [@Feinman2020-nf; @Holland-Lulewicz2021-eb].
 Multilayer network analysis combines multiple networks into one analytical framework [@Kivela2014-mh].
 Social systems are complex webs of interrelations.
-There is too much involved to model the material culture of even a technologically simple society.
-But certainly analyzing more than one type of material culture provides a more complete understanding of the social dynamics involved in the networks of interaction we wish to understand.
+There is too much involved to model the material culture of even a technologically simple society, but certainly analyzing more than one type of material culture provides a more complete understanding of the social dynamics involved in the networks of interaction we wish to understand.
 Yet, there are few examples of multilayer networks in archaeology [@Giomi2022-dn; @Upton2019-yg] and none I am aware of that extend beyond ceramics.
@@
 Network studies are particularly prominent in the Southwest and there are examples of network studies that consider multiple types of material culture.
-For example, Mills and colleagues [-@Mills2013-gc] used ceramics and obsidian to understand the transformation of large scale social networks.
+For example, Mills and colleagues [-@Mills2013-gc] used ceramics and obsidian to understand the transformation of large-scale social networks.
 Peeples [-@Peeples2018-ib] used ceramics and architecture to understand identity and social change in the Cibola region.
@@
 Given the frequent focus on single types of artifacts in network research, a primary question I seek to address is whether networks based on different types of material culture are positively correlated.
-This is an important question because archaeologists often make inferences based solely on ceramics or other types of material culture.
-Evidence regarding the ways material culture co-vary will help archaeologists direct future research.
+This is important because archaeologists often make inferences based solely on ceramics or other types of material culture.
+Evidence regarding the ways material culture co-vary will help give better context to these studies.
 Furthermore, I also seek causal explanations for the co-variance, or lack thereof, for material culture in the case study I present here.
@@
 Tonto Basin holds great potential for archaeological research due to the large cultural resource management projects conducted in the region primarily in the 1980s-1990s [@Ahlstrom1991-eb; @Ciolek-Torrello1994-ds; @Doelle1992-lu; @Rice1998-ku], as well as the large number of syntheses and other studies focusing on the area [e.g., @Clark2004-uw; @Dean2000-uc; @Elson1995-tb; @Elson2000-zq;  @Hill2015-at; @Huntley2016-em; @Lange1992-gp; @Lyons2003-yy; @Lyons2006-to; @Lyons2012-pr;  @Neuzil2008-zd; @Oliver2001-tu; @Stark1998-mu; @Watts2013-ub].
-This analysis uses data from the Roosevelt Platform Mounds Study (RPMS), the largest of these projects [@Rice1998-ku], to examine basic architectural data, typed ceramics, and a recent projectile point analysis [@Bischoff2022-rg].
+This analysis uses data from the Roosevelt Platform Mounds Study (RPMS), the largest of these projects [@Rice1998-ku], to examine basic architectural data, typed ceramics, and data from a recent projectile point analysis [@Bischoff2022-rg].
 There are two null hypotheses tested in this analysis: (1) that the data is spatially correlated--meaning that sites nearest to each other will be most alike--and (2) that the architecture, ceramics, and projectile points will all be positively correlated--meaning that similar types of architecture, ceramics, and projectile points will be found at the same sites.
@@
 As will be demonstrated, the results indicate a positive correlation between projectile point similarity and spatial distance.
-Furthermore, ceramic and projectile point networks (hereafter point networks) exhibit significant differences.
+Additionally, ceramic and projectile point networks (hereafter point networks) exhibit significant differences.
 I posit that the differences in these networks are connected to the identities of the individuals creating the original social networks: particularly that of gender.
@@
 One aspect of Tonto Basin that has caught the attention of researchers is its location in a border zone between the Hohokam, Mogollon, and Ancestral Pueblo regions [e.g., @Caseldine2022-uu; @Clark2001-ck; @Elson1994-hk; @Hill2015-at; @Huntley2016-em; @Lyons2012-pr; @Lyons2013-ya; @Neuzil2008-zd; @Wood2000-ze].
-Of particular note is the presence of masonry roomblock architecture and pottery uncharacteristic of the local Hohokam traditions, which has been interpreted as Kayenta immigration into Tonto Basin [@Clark2001-ck; @Lyons2003-yy; @Lyons2006-to; @Stark1998-mu].
-This migration into the Tonto Basinwas part of a larger migration from north to south and is associated with the origin of the Salado phenomenon.
+Of particular note is the presence of masonry roomblock architecture and pottery uncharacteristic of the local Hohokam traditions, which has been interpreted as Kayenta immigration into Tonto Basin [@Clark2001-ck; @Lyons2003-yy; @Lyons2006-to; @Stark1998-mu], although some have attributed the architectural changes to warfare or population aggregation [c.f.
+@Oliver2001-tu; @Wood2000-ze].
+This migration into the Tonto Basin was part of a larger migration from north to south and is associated with the origin of the Salado phenomenon.
 Salado pottery production was widespread across southern Arizona, but often the largest sources of production was centered at the location of a former Kayenta enclave [@Hill2015-at; @Huntley2016-em; @Lyons2012-pr; @Neuzil2008-zd].
@@
 A network approach is an ideal way to examine the relationships between sites.
-Evidence that projectile points move between sites [e.g., @Watts2013-ub] indicates that this is also a productive line of analysis, and one that is less-commonly pursued compared to architecture and ceramics.
+Evidence that projectile points move between communities in the Tonto Basin [e.g., @Sliva2002-oz; @Watts2013-ub] indicates that this is also a productive line of analysis, and one that is less-commonly pursued compared to architecture and ceramics.
 
@@
 Watts [-@Watts2013-ub] used projectile points from Tonto Basin in just such a study.
-His analysis of flake scar patterning identified individual knappers, or at least clusters of knappers with similar knapping styles, and the distribution of their projectile points around Tonto Basin.
+His analysis of flake scar patterning identified individual knappers, or at least clusters of people with similar knapping styles, and the distribution of their projectile points around Tonto Basin.
 He found strong connections between many parts of the eastern Tonto Basin.
@@
 The ceramics are grouped into types based primarily on decoration, which is strongly indicative of group membership.
-What this means is that links between nodes in the cer The details of architecture and ceramics used in this study are discussed in a future section, but it is important to note that these designations as categorical or relational are contextually dependent on this study.
+What this means is that links between nodes in the ceramic network are more likely to indicate belonging to a similar social group.
+The details of architecture and ceramics used in this study are discussed in a future section, but it is important to note that these designations as categorical or relational are contextually dependent on this study.
 Clark [-@Clark2001-ck] has an excellent discussion on material culture and its relationship to identity.
@@
 Women made pottery and men made projectile points.
-This was not always true of course, but this fits the available data and expectations for the Hohokam [@Crown1996-xb; @Harry2010-hj; @VanPool2010-im, p.
+This was not always true of course, but this fits the available data and expectations for the Hohokam [@Crown1996-xb; @Harry2010-hj; @Shackley2005-hn; @VanPool2010-im, p.
 253; @Whittlesey2010-ox].
@@
 This does not mean that a man could not move pots from one site to another or that a woman could not do the same with a projectile point, but, in general, differences in point and ceramic networks are most likely to indicate differences in gender networks.
+
+Shackley [-@Shackley2005-hn], using projectile point typology data from Hoffman [-@Hoffman1997-hb] as well as obsidian provenience and ceramic data for preclassic Hohokam sites in the Phoenix Basin, found evidence for three distinct projectile point traditions.
+Shackley and Hoffman believe these may correlate with warrior sodalities, and Shackley argues that the male projectile point exchange systems functioned in different ways than the female ceramic exchange systems.
+In this case, the projectile points likely correlate with a categorical identity, but this provides an example of how men and women's networks have been argued to vary in Hohokam archaeology.
 
@@
   select(Site = ArchaMapName,Area,Architecture = architecture,Phase,Ceramics,Points) %>%
-  # as_hux() %>% set_width(1) %>% set_number_format(everywhere,c(5,6),fmt_pretty()) %>% 
-  # theme_article() %>% 
-  # set_caption(cap)
-  knitr::kable(caption= cap,
-               booktabs = T,
-               format.args = list(big.mark = ",",
-                                  scientific = FALSE)) %>%
-  kable_styling(latex_options = "scale_down")
+  as_hux() %>% set_width(1) %>% set_number_format(everywhere,c(5,6),fmt_pretty()) %>%
+  theme_article() %>%
+  set_caption(cap)
+  # knitr::kable(caption= cap,
+  #              booktabs = T,
+  #              format.args = list(big.mark = ",",
+  #                                 scientific = FALSE)) %>%
+  # kable_styling(latex_options = "scale_down")
 ```
@@
 This can be a challenge for network analysis, because we can end up with networks where every node is connected to every other node.
-The decision to binarize a network--remove the weakest links and then consider each link of equal value--has its drawbacks [@Peeples2013-tj], but is necessary in this case.
+The decision to binarize a network--remove the weakest links and then consider each link of equal value--has its drawbacks [@Peeples2013-tj] but is necessary in this case.
 A solution is to assign weights to each link that defines the strength of the tie.
@@
 Eigenvector centrality was mentioned in the previous section, but it is also calculated as a multilayer eigenvector centrality.
-In this case the centrality measure is simply the mean of the eigenvector centraliy for each separate layer, as the layers are not interdependent [@Frost2022-lr].
+In this case the centrality measure is simply the mean of the eigenvector centrality for each separate layer, as the layers are not interdependent [@Frost2022-lr].
 Multilayer eigenvector centrality results are only presented for the combination of ceramic and point layers.
@@
 
-The robustness of results were tested by randomly sampling the underlying network matrices 10,000 times and comparing the results to the random samples to obtain a p-value.
+The robustness of results was tested by randomly sampling the underlying network matrices 10,000 times and comparing the results to the random samples to obtain a p-value.
 This provides a baseline that determines how likely it is to obtain a given result by chance.
@@
 Network analysts have grappled with this question and often use resampling or similar methods to deal with this problem [@Bischoff2021-sf; @Bolland1988-zm; @Borgatti2006-jl; @Brughmans2023-uj; @Costenbader2003-st; @Galaskiewicz1991-qb; @Gjesfjeld2015-hw; @Lee2009-wr; @Rivera-Hutinel2012-ik].
-Regardless of statistical methods, unaccounted for biases will still produce invalid results [@Bischoff2021-sf].
+Regardless of statistical methods, unaccounted for biases will still produce invalid network results [@Bischoff2021-sf].
 The most important bias that may affect these findings is chronology.
@@
 Bass Point Mound forms a crucial bridge between the Cline Mesa sites and the rest of the network.
-Because the network is not complete, one cannot argue that there were not other sites between Cline Mesa and the Schoolhouse Mesa sites, but it still represents an intermediary location.
+Because the network is not complete, one cannot argue that there were no other sites between Cline Mesa and the Schoolhouse Mesa sites, but it still represents an intermediary location.
 Its geographic position overlooking the confluence of the Salt River and Tonto Creek would make an ideal meeting place for parties coming down from the Tonto Creek arm of what is now Roosevelt Lake or coming up from the Salt River arm.
@@
 
-```{r TontoNetworksMultiNet, fig.cap= 'Graphs showing the networks used in this analsyis.
+```{r TontoNetworksMultiNet, fig.cap= 'Graphs showing the networks used in this analysis.
 Each network is displayed with the five strongest ties between each node.
@@
   arrange(desc(Centrality)) %>% 
-  knitr::kable(caption= cap,
-               booktabs = T,
-               format.args = list(big.mark = ",",
-                                  scientific = FALSE))
-  # as_hux() %>% 
-  # theme_article() %>% 
-  # set_caption(cap)
+  # knitr::kable(caption= cap,
+  #              booktabs = T,
+  #              format.args = list(big.mark = ",",
+  #                                 scientific = FALSE))
+  as_hux() %>%
+  theme_article() %>%
+  set_caption(cap)
 
@@
 Shell and stone jewelry, among other artifacts were also produced and exchanged within the basin.
-A potential loci for exchange are the platform mound sites.
+Potential loci for exchange are the platform mound sites.
 The purpose of these sites is still debated, but many argue that they served an integrative function [e.g., @Abbott2006-fg; @Adler1990-ws; @Clark2004-uw; @Craig1995-qo; @Craig1994-bg, p.
@@
 This is advantageous for studying how types of material culture do or do not co-vary, which aids interpretations of the social interactions that created these patterns.
-This analysis used architectural, ceramic, projectile point, and spatial data from `r nrow(sites)` sites in Tonto Basin.
-The data come primarily from occupations between AD 1275 and 1325.
+This analysis used architectural, ceramic, projectile point, and spatial data from `r nrow(sites)` sites in Tonto Basin dating primarily from occupations between AD 1275 and 1325.
 A network was created for each type of data and combined into a multilayer network.
@@
 Finally, the results demonstrate major differences between the ceramic and point networks.
-This findings has major implications for network studies, because many rely on only a single artifact type.
+This finding has major implications for network studies, because many rely on only a single artifact type.
 If different types of material culture do not co-vary regularly, then that indicates archaeologists must do more to include other types of material culture to better understand the complex social networks that existed in the past.
 Furthermore, projectile points and ceramics have strong associations with gender.
-Differences in the ceramic and point networks suggest differences existed between the social networks of each gender.
+Differences in the ceramic and point networks suggest differences existed between the men's and women's social networks.
 
@@
 All mistakes are my own.
-The many individuals who contributed to the Roosevelt Platform Mound Study also deserve recognition.
+The many individuals who conducted the Roosevelt Platform Mound Study deserve recognition for contributing the data.
 

### 2023-01-12 manuscript.Rmd
minor revisions

@@
 output:
+  rticles::tf_article: NULL
   bookdown::word_document2:
@@
     reference_docx: TF.docx
-  rticles::tf_article: null
 author:
@@
   Network analysis has a strong foundation in Southwest archaeology, yet the combination of multiple types of artifacts in one analysis--multilayer network analysis--has not been formally applied except within a single artifact type.
-Many studies consider material culture holistically, yet network analysis has the advantage of focusing specifically on the relationships between entities (often communities) and how the structure of a network can benefit particular entities.
+Many studies consider material culture holistically, yet network analysis has the advantage of focusing specifically on the relationships between entities (often communities) and how the structure of a network can benefit entities.
 This study uses architecture, ceramic, projectile point, and site location data from the Roosevelt Platform Mound Study and combines these data in a multilayer network analysis.
@@
   \def\UrlBreaks{\do\/\do-\do?}
+  \usepackage{lineno}
+  \linenumbers
 csl: american-antiquity.csl
@@
 There is too much involved to model the material culture of even a technologically simple society, but certainly analyzing more than one type of material culture provides a more complete understanding of the social dynamics involved in the networks of interaction we wish to understand.
-Yet, there are few examples of multilayer networks in archaeology [@Giomi2022-dn; @Upton2019-yg] and none I am aware of that extend beyond ceramics.
+Yet, there are few examples of multilayer networks in archaeology [e.g., @Giomi2022-dn; @Upton2019-yg] and none I am aware of that extend beyond ceramics.
 
 Network studies are particularly prominent in the Southwest and there are examples of network studies that consider multiple types of material culture.
-For example, Mills and colleagues [-@Mills2013-gc] used ceramics and obsidian to understand the transformation of large-scale social networks.
-Peeples [-@Peeples2018-ib] used ceramics and architecture to understand identity and social change in the Cibola region.
+To name just two: Mills and colleagues [-@Mills2013-gc] used ceramics and obsidian to understand the transformation of large-scale social networks; Peeples [-@Peeples2018-ib] used ceramics and architecture to understand identity and social change in the Cibola region.
 This paper is an application of multilayer network analysis that seeks to further network research by examining multiple types of material culture in one framework.
 Given the frequent focus on single types of artifacts in network research, a primary question I seek to address is whether networks based on different types of material culture are positively correlated.
-This is important because archaeologists often make inferences based solely on ceramics or other types of material culture.
+This matters because archaeologists often make inferences based solely on ceramics or other types of material culture.
 Evidence regarding the ways material culture co-vary will help give better context to these studies.
@@
 
-## Tonto Basin
+## Tonto Basin {-}
 
-Tonto Basin is located in east-central Arizona and features Tonto Creek flowing from the Mogollon Rim into the Salt River (figure `r figs`).
+Tonto Basin is in east-central Arizona and features Tonto Creek flowing from the Mogollon Rim into the Salt River (figure `r figs`).
 Parts of the region have been intensively studied, while certain time periods and much of the uplands have been less intensively sampled [see @Clark2021-lu for a recent overview].
-The basin was also conveniently located along regional travel routes, allowing major settlements--particularly those with platform mounds--to participate and benefit from this exchange [@Caseldine2022-uu; @Wood2000-ze, p.
+The basin was conveniently located along regional travel routes, allowing major settlements--particularly those with platform mounds--to participate and benefit from this exchange [@Caseldine2022-uu; @Wood2000-ze, p.
 129-133].
@@
 
-One aspect of Tonto Basin that has caught the attention of researchers is its location in a border zone between the Hohokam, Mogollon, and Ancestral Pueblo regions [e.g., @Caseldine2022-uu; @Clark2001-ck; @Elson1994-hk; @Hill2015-at; @Huntley2016-em; @Lyons2012-pr; @Lyons2013-ya; @Neuzil2008-zd; @Wood2000-ze].
-Of particular note is the presence of masonry roomblock architecture and pottery uncharacteristic of the local Hohokam traditions, which has been interpreted as Kayenta immigration into Tonto Basin [@Clark2001-ck; @Lyons2003-yy; @Lyons2006-to; @Stark1998-mu], although some have attributed the architectural changes to warfare or population aggregation [c.f.
-@Oliver2001-tu; @Wood2000-ze].
+Tonto Basin's position in a border zone between the Hohokam, Mogollon, and Ancestral Pueblo regions has received much attention [e.g., @Caseldine2022-uu; @Clark2001-ck; @Elson1994-hk; @Hill2015-at; @Huntley2016-em; @Lyons2012-pr; @Lyons2013-ya; @Neuzil2008-zd; @Wood2000-ze].
+Of particular note is the presence of masonry roomblock architecture and pottery uncharacteristic of the local Hohokam traditions, which has been interpreted as Kayenta immigration into Tonto Basin [@Clark2001-ck; @Lyons2003-yy; @Lyons2006-to; @Stark1998-mu], although some have attributed the architectural changes to warfare or population aggregation [@Oliver2001-tu; @Wood2000-ze].
 This migration into the Tonto Basin was part of a larger migration from north to south and is associated with the origin of the Salado phenomenon.
 Salado pottery production was widespread across southern Arizona, but often the largest sources of production was centered at the location of a former Kayenta enclave [@Hill2015-at; @Huntley2016-em; @Lyons2012-pr; @Neuzil2008-zd].
-The connection between immigration and an important type of decorated pottery in the region suggests that sites with roomblocks may have higher centrality in ceramic networks.
+The connection between immigration and Salado pottery suggests that sites with roomblocks may have higher centrality in ceramic networks.
 Having high centrality means they have more connections to other nodes (i.e., sites).
@@
 
-## Defining Networks and Relations
+## Defining Networks and Relations {-}
 
@@
 The types of similarity used in this study are varied depending on the type of network.
-Part of this analysis is a visual approach and some of the methods require non-weighted links, thus I will be only using the strongest links.
+Part of this analysis is a visual approach and some of the methods require non-weighted links, thus I use only the strongest links.
 There is rarely a clear dividing line between similar and not similar.
@@
 This involved calculating the strength of similarity between each site in the network and then ranking the strength of similarity.
-The top **n** connections between each site were kept and the rest were discarded.
+The top **n** connections (**n** varies between 3 and 10) between each site were kept and the rest were discarded.
 In practice, not every node will have the same number of connections.
@@
 Either Pearson or Spearman rank correlation can be computed to determine the strength and direction of correlation between each layer.
-Bródka and colleagues [-@Brodka2018-vz] have provided an R package to compute these statistics, which I have used in this study.
+Bródka and colleagues [-@Brodka2018-vz] have provided an R [@RCore2022-a] package to compute these statistics, which I have used in this study.
 They recommend the Pearson correlation in most circumstances.
@@
 Multilayer eigenvector centrality results are only presented for the combination of ceramic and point layers.
-These networks can be thought of as dependent variables for the spatial and architectural networks--meaning spatial distance and architecture are considered to be important variables determining the structure of the ceramic and point networks.
+These networks can be thought of as dependent variables for the spatial and architectural networks--meaning spatial distance and architecture are considered important variables determining the structure of the ceramic and point networks.
 
@@
 For example, Cline Terrace Mound and Schoolhouse Point were most probably occupied at the same time [@Lyons2013-ya].
-It is best, though, to think of these interpretations as estimations based on incomplete data.
+It is best, though, to think of these interpretations as estimations based on incomplete data and to recognize that these networks are an analytical tool and not a representation of an ancient face-to-face interaction network.
 
@@
 They were common in the Hohokam region throughout the Classic period.
-Some sites had pit houses from earlier occupations, but most excavation was focused on the Roosevelt and Gila phases so these few pit houses were removed from the analysis [see @Rice1998-ku].
+Some sites had pit houses dating to earlier occupations, so these few pit houses were removed from the analysis [see @Rice1998-ku].
 Each site was classified as a platform mound if one was present regardless of other architecture or as a roomblock if a platform mound was not present regardless of other architecture.
 Thus, if a site is labeled a platform mound it may have other architecture.
-Sites were thus labeled to emphasize the most distinctive form of architecture, but all types of architecture were present in the analysis, and all architectural features included in the analysis can be seen in Table 1.
+Sites were labeled in this manner to emphasize the most distinctive form of architecture, but all types of architecture were present in the analysis, and all architectural features included in the analysis can be seen in Table 1.
 
@@
 Figure `r figs` shows the networks with the top five strongest ties between each node.
-This number of ties represents the fewest ties that connects all nodes in the spatial network.
-Fewer ties leaves the Cline Mesa sites disconnected from the rest of the network.
+This number of ties represents the fewest ties that connect all nodes in the spatial network.
+Fewer ties leave the Cline Mesa sites disconnected from the rest of the network.
 Bass Point Mound forms a crucial bridge between the Cline Mesa sites and the rest of the network.
@@
 cap = "Mean Eigenvalues for Ceramic and Point Networks by Type of Architecture"
+tbls = tbls + 1
 eigenDF %>% 
@@
   arrange(desc(mean)) %>%
-  # as_hux() %>% 
-  # theme_article() %>% 
-  # set_caption(cap)
-  knitr::kable(caption= cap,
-               booktabs = T)
+  as_hux() %>%
+  theme_article() %>%
+  set_caption(cap)
+  # knitr::kable(caption= cap,
+               # booktabs = T)
 ```
@@
 
-# Discusssion {-}
+# Discussion {-}
 
@@
 Potential loci for exchange are the platform mound sites.
-The purpose of these sites is still debated, but many argue that they served an integrative function [e.g., @Abbott2006-fg; @Adler1990-ws; @Clark2004-uw; @Craig1995-qo; @Craig1994-bg, p.
+The purpose of these sites is debated, but many argue that they served an integrative function [e.g., @Abbott2006-fg; @Adler1990-ws; @Clark2004-uw; @Craig1995-qo; @Craig1994-bg, p.
 112-165; @Doelle1995-qd, p.
@@
 While this association is circumstantial and not expected to be uniform across the Hohokam region, roomblocks did have the highest centrality in the ceramic networks.
-Another reason for the high centrality of roomblock sites is that they lacked access to adequate farmland and had to trade pottery for food [@Clark2004-uw, p.
+Another proposed reason for the high centrality of roomblock sites is that they lacked access to adequate farmland and had to trade pottery for food [@Clark2004-uw, p.
 291].
 This second explanation may help explain why roomblock sites had low centrality in the point networks, presuming that points were not exchanged for food.
-It is beyond this simple analysis to determine the precise reasons, but it is clear that roomblock sites were important for ceramic circulation in Tonto Basin.
+It is beyond this simple analysis to determine the precise reasons, but clearly roomblock sites were important for ceramic circulation.
 On the other hand, sites with roomblocks had the lowest centrality in the point networks.
@@
 Projectile points at least were highly gendered, and ceramics probably were as well.
-The evidence presented here demonstrates that ceramic and point networks vary significantly, and I believe gender likely played an important role.
+The evidence presented here demonstrates that ceramic and point networks vary significantly, and gender likely played an important role.
+More research will be needed to determine how gender roles affected the networks.
 
@@
 Finally, the results demonstrate major differences between the ceramic and point networks.
-This finding has major implications for network studies, because many rely on only a single artifact type.
+This finding has major implications for network studies because many rely on only a single artifact type.
 If different types of material culture do not co-vary regularly, then that indicates archaeologists must do more to include other types of material culture to better understand the complex social networks that existed in the past.
 Furthermore, projectile points and ceramics have strong associations with gender.
-Differences in the ceramic and point networks suggest differences existed between the men's and women's social networks.
+Differences in the ceramic and point networks suggest differences existed between the social networks of men and women.
 
-This analysis uses a handful of sites from Tonto Basin.
+This analysis used a handful of sites from Tonto Basin.
 Much more data exists and the results discussed here would be greatly strengthened by including more sites and data.
-What would be perhaps even more useful would be to include sourced obsidian and ceramics.
+What would be perhaps more useful would be to include sourced obsidian and ceramics.
 Networks created from artifacts where the origin is known can be used to more strongly infer various types of social interaction.
@@
 It would be particularly useful for comparative studies to know in what circumstances different types of material culture do or do not co-vary.
+It is my hope that this study highlights the usefulness of a multilayer network approach utilizing multiple types of material culture.
 
@@
 
-Josh Watts, Matt Peeples, Melissa Powell, Chris Caseldine, and several volunteers assisted with this research.
+Josh Watts, Matt Peeples, Melissa Powell, Chris Caseldine, Michael Brooks, and Ronald Vasquez assisted with this research.
 Chris Caseldine's comments greatly improved this paper.
 All mistakes are my own.
-The many individuals who conducted the Roosevelt Platform Mound Study deserve recognition for contributing the data.
+The many individuals who conducted the Roosevelt Platform Mound Study deserve recognition for their hard work.
 
